---
title: "Extending the API"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Extending the API}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

In this vignette we will cover some general guidelines that ensure your anvl functions come without surprises.
This is primarily intended for extending the API -- either in your own package or contributing to {anvl} itself -- but is also helpful when writing your own scripts.

The general guidelines are:

1. The function must be pure.
1. Consistent input and output types:
   1. The dynamic (arrayish) inputs should accept `AnvlArray`s as well as R vectors of length 1 and `array`s.
   1. The function should only output `AnvlArray`s.
1. The function should work with arbitrary devices.
1. The function should (unless there are specific reasons) work in eager and jit mode.
1. Use static arguments when you require data-dependent input checks.
1. Tag the function with `#' @jit` so it is jit-wrapped at package build time
   (see *Jit-wrapping API Functions* below).

## Pure Functions

This is extensively covered in the *JIT Deep Dive*, so we won't repeat it here.
While the subsequent sections mostly address issues that are relevant in eager mode, purity is the primary requirement to enable usage of `jit()` with your function.

## Consistent Input and Output Types

Functions in anvl have dynamic (arrayish) and static (standard R values) inputs.
However, it can also be convenient to pass R objects as dynamic inputs and let anvl convert them.
To enable this, there are the `as_anvl_array()` and `as_anvl_arrays()` converters.
You should call them at the top of your function.
Not only will these functions convert the inputs, they will also check them for compatibility, specifically w.r.t. their device and backend.
If they don't live on the same device, an error will be thrown.

This matters most in eager mode: under `jit()`, {anvl} sees the whole call and
places the inputs itself. But the converters mean the same thing in both modes,
so writing your function this way costs nothing when it *is* jitted -- and see
point 4 below for what it buys.

The advantage of this input standardization is best illustrated with an example.

Consider the naive implementation of reshaping, which will fail when called on an R vector:

```{r, error = TRUE}
library(anvl)
# x: dynamic, shape: static
nv_reshape_naive <- function(x, shape) {
  if (!identical(shape(x), shape)) {
    prim_reshape(x, shape)
  } else {
    x
  }
}
nv_reshape_naive(1L, c(2, 2))
```

This is because `shape()` reports the shape of the *value*, and an R vector that is neither length-1 nor an `array()` is not an anvl value at all -- canonicalizing at the top is what turns a valid input into something the rest of the function can work with uniformly.

Also, consider this function that converts an input to a specific dtype (or keeps it as-is if `dtype` is `NULL`).
The problem is that in the no-op case, we return a static R object instead of (as intended) an `AnvlArray`.

```{r}
# x: dynamic, dtype: static
nv_convert_naive <- function(x, dtype) {
  if (is.null(dtype)) {
    return(x)
  }
  prim_convert(x, dtype)
}
nv_convert_naive(1L, "i16")
nv_convert_naive(1L, NULL)
```

By canonicalizing inputs, such pitfalls can be avoided.

Finally, note that primitives such as `prim_convert()` already canonicalize their inputs, so if you are only wrapping primitives (or other `nv_<op>` functions that already canonicalize), you might not have to do this yourself.

### Start an eager function with a single `as_anvl_arrays()` call

When a function takes more than one arrayish input, normalize **all of them in
one `as_anvl_arrays(...)` call** at the top. Not one `as_anvl_array()` per
argument, and not part-way down the body -- the call has to see the whole set,
because three of the four things it does are decisions *across* the arguments.

```{r}
nv_scaled_sum <- function(x, y, z) {
  args <- as_anvl_arrays(x, y, z)
  args[[1L]] * nv_add(args[[2L]], args[[3L]])
}
```

**1. It puts every input on one device.** An R value has no device, so it is
placed on one when it is converted -- and on its own it would be placed on the
*default* device, whatever the other arguments are on. Converting the set
together places it where its `AnvlArray` siblings live instead:

```r
naive <- function(x, y, z) x * nv_add(y, z)
naive(nv_scalar(1, device = "cuda"), 2, 3)
#> Error: invalid input `rhs`: it lives on a different device than an earlier
#> input; all inputs must share a device unless a target device is fixed

nv_scaled_sum(nv_scalar(1, device = "cuda"), 2, 3)   # fine
```

It also rejects a call that mixes devices or backends, which is the check you
want at the boundary rather than somewhere in the middle of the body.

**2. It does the type promotion.** With a `.promote` rule the whole set comes
back at one data type. Which one is the rule's job to say, and there are three:

```{r}
# the common data type of the inputs -- what a binary operator does
as_anvl_arrays(nv_array(1L), 1.5, .promote = promote_common())
# the data type one particular input already has
as_anvl_arrays(x = nv_array(1L), y = 1.5, .promote = promote_like("x"))
# a data type the function names itself
as_anvl_arrays(nv_array(1L), 1.5, .promote = promote_dtype("f64"))
```

Promoting *realizes* an R value at the target rather than converting it there:
it is built at that data type from the R data, so it arrives with every digit it
had. Converting afterwards instead -- canonicalizing first and then calling
`nv_convert()` -- rounds through the default `f32` on the way, so
`x_f64 * sqrt(2)` comes out right only in the first spelling. **If your
function's result data type depends on its arguments, say so with a rule.**

Which rule you want follows from what the function is. `nv_add()` is
`promote_common()`: neither operand is privileged. `nv_clamp(min_val, x,
max_val)` is `promote_like("x")`: the result is `x`'s type and the bounds come
along, even where that narrows them. `nv_rnorm(shape, state, dtype = )` is
`promote_dtype(dtype)`: the type is the function's own argument, not something
read off an input.

An argument that has no business taking part is named out with `only`. Every
input is still aligned onto one device and converted -- it just does not take
the target's data type. `nv_ifelse()` is the case: its predicate has to stay a
`bool`, or the primitive underneath would be handed a float predicate.

```{r}
as_anvl_arrays(
  pred = nv_array(c(TRUE, FALSE)),
  true_value = nv_array(c(1L, 2L), dtype = "i8"),
  false_value = 3L,
  .promote = promote_common(only = c("true_value", "false_value"))
)
```

`.promote` also takes a *list* of rules, for a function whose arguments fall into
several groups that are promoted independently:

```{r}
as_anvl_arrays(
  x = nv_array(1L), y = 1.5,
  a = nv_array(1L, dtype = "i8"), b = 2L,
  .promote = list(
    promote_common(only = c("x", "y")),
    promote_common(only = c("a", "b"))
  )
)
```

The rules must name disjoint sets -- a value cannot be brought to two data types
-- and all of them are resolved before any is applied, so one group's target
never depends on another group having been converted first.

`promote_yield()` is the fourth rule, and the odd one out: it reads its target
off the inputs that already have a data type and *never converts them*, so only
the R values move. It is what the primitives use, and what you want for a
function whose arguments must agree but which should not silently widen the
array it was handed.

```{r, error = TRUE}
# the R value takes the array's data type
as_anvl_arrays(nv_array(1, dtype = "f64"), 1.5, .promote = promote_yield())
# ... but only within its own category: an integer literal stays an integer
as_anvl_arrays(nv_array(1, dtype = "f64"), 1L, .promote = promote_yield())
```

### Arguments may be trees

An argument can be a nested list of arrayish values rather than a single one.
The device and the `.promote` target are decided over every *leaf* of every
argument, and each argument comes back with the structure it had:

```{r}
as_anvl_arrays(
  nv_array(1L, dtype = "i8"),
  list(a = 1.5, b = list(2L)),
  .promote = promote_common()
)
```

A rule names whole arguments, never individual values inside one, so a tree
takes part as a whole: all of its leaves contribute to the target and all of
them are brought to it.

**3. Everything downstream is then well-defined.** After the call each argument
is an `AnvlArray` (or, under `jit()`, a traced box), so `dtype()`, `device()`
and `backend()` answer. On a bare R value they do not -- an R value has no data
type of its own until it is used (see `?RDataArray`):

```{r, error = TRUE}
(function(x) dtype(x))(2)
```

```{r}
(function(x) dtype(as_anvl_array(x)))(2)
```

If you only need to *look* at the data type -- an "is this a float?" branch --
`peek_dtype()` answers for a bare R value too, with the one it would take.

**4. It keeps eager mode and `jit()` in step.** Without it, a function whose
arguments happen to all be R values never reaches anvl at all: base R evaluates
it, and you get an R vector rather than an array.

```{r}
g <- function(x, y) x + y
g(1, 2)         # base R's `+`: an R double
jit(g)(1, 2)    # anvl: an f32 array
```

```{r}
h <- function(x, y) {
  args <- as_anvl_arrays(x, y)
  args[[1L]] + args[[2L]]
}
h(1, 2)
jit(h)(1, 2)
```

`as_anvl_array()` and `as_anvl_arrays()` convert -- always, and the same way in
both modes. That is what makes the two columns above agree, and it is worth
insisting on: a function that meant two different things depending on whether it
was called under `jit()` would silently give two different answers for one call.

## Arbitrary Devices

In order to ensure that your function works with inputs from arbitrary devices, you need to be careful when creating new constants within your function.
Let's say you are creating your function and working on GPU:

```{r}
nv_add_one_naive <- function(x) {
  x <- as_anvl_array(x)
  x + nv_fill(1L, shape(x), device = "cuda")
}
```

As long as you are adding ones on a CUDA GPU, this function will work fine!
However, if you suddenly use it on the CPU, it will fail, because we can't add a CPU array to a CUDA array.
Constants should always be initialized on the same device as the inputs.
If there are multiple inputs and you called `as_anvl_arrays()` on them at the top, you know that there is only a single device.

One way to achieve this is to simply pass the input's device to `nv_fill()`:

```{r}
nv_add_one1 <- function(x) {
  # Converting is what gives the value a device to read back here.
  x <- as_anvl_array(x)
  x + nv_fill(1L, shape(x), device = device(x))
}
```

Another option is to rely on `nv_<op>_like` functions.
These take in another `AnvlArray` as their first input and use its properties as the defaults for their arguments.
In this case, the created array will assume the data type, shape and device from the input array.

```{r}
nv_add_one2 <- function(x) {
  x + nv_fill_like(x, 1L)
}
```

Note that when you only want to use a function with `jit()`, you can just omit specifying the device at all, as `jit()` is smart enough to place it on the correct device.

## Static Arguments to Enable Input Checks

One restriction of the XLA compiler is that it does not really allow for runtime checks.
Let's say you want to sample from a Bernoulli distribution with probability `p`.
If you make `p` a dynamic input, you can't check that it is within `[0, 1]`, so you need to make it a static input.
Don't convert it to an `AnvlArray` before checking its value.
Later in the function, it will actually be converted, but from XLA's point of view, it will just be a constant within the compiled program and not a dynamic input.

```{r}
nv_rbernoulli <- function(initial_state, p) {
  initial_state <- as_anvl_array(initial_state)
  stopifnot((p >= 0) && (p <= 1))

  # returns: (state, sample)
  out <- nv_runif(1L, initial_state)
  out_state <- out[[1L]]
  x <- nv_convert(out[[2L]] <= p, "i32")
  list(out_state, x)
}
nv_rbernoulli(nv_rng_state(1), 0.2)[[2L]]
```

## Jit-wrapping API Functions

Most user-facing API functions in anvl are wrapped in `jit(f, ...)` so that calling them traces and compiles a single program instead of executing each operation eagerly.
The wrapping is driven by the `@jit` roclet (see `?jit_roclet`).

In `R/api*.R`, tag any function that performs more than one primitive operation with `#' @jit`:

```r
#' @export
#' @jit
nv_log2 <- function(x) {
  x <- as_anvl_array(x)
  nv_log(x) / log(2)
}
```

If the function has static arguments (anything that is not an arrayish input -- axes, shape, dtype, control flags, functions used as templates, ...), list them with `static = c(...)` using either positional indices or argument names:

```r
#' @export
#' @jit static = c(2L, 3L)
nv_mean <- function(x, axes = NULL, drop = TRUE) {
  ...
}

#' @export
#' @jit static = c("axis")
nv_concatenate <- function(..., axis = NULL) {
  ...
}
```

Use names rather than positions whenever an argument lives after `...`, since `...` has no fixed position.

**When to skip `#' @jit`.** Don't tag a function whose body is essentially a single primitive call -- direct aliases (`nv_log <- prim_log`), `make_do_binary(prim_X)` factories, or thin wrappers that just validate and forward to one primitive. The underlying primitive is already jit-wrapped, so adding another `jit()` layer adds tracing overhead without fusing anything new.
Also skip pure I/O (`nv_save`, `nv_serialize`), backend constructors (`nv_array`, `nv_scalar`, `nv_matrix`), and device/state objects (`nv_device`, `nv_rng_state`).

The roclet writes the list of tagged functions to `R/jit-registry.R` on every `devtools::document()` run, and `R/zzz.R` applies that registry at package source time so the wrapped functions are byte-compiled with the rest of the package.
