---
title: "Extending the API"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Extending the API}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
library(anvl)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

In this vignette we will cover some general guidelines that ensure your anvl functions come without surprises.
This is primarily intended for extending the API -- either in your own package or contributing to {anvl} itself -- but is also helpful when writing your own scripts.

The general guidelines are:

1. The function must be pure.
1. Consistent input and output types:
   1. The dynamic (arrayish) inputs should accept `AnvlArray`s as well as R vectors of length 1 and `array`s.
   1. The function should only output `AnvlArray`s.
1. The function should work with arbitrary devices.
1. The function should (unless there are specific reasons) work in eager and jit mode.
1. Use static arguments when you require data-dependent input checks.
1. Tag the function with `#' @jit` so it is jit-wrapped at package build time
   (see *Jit-wrapping API Functions* below).

## Pure Functions

This is extensively covered in the *JIT Deep Dive*, so we won't repeat it here.
While the subsequent sections mostly address issues that are relevant in eager mode, purity is the primary requirement to enable usage of `jit()` with your function.

## Consistent Input and Output Types

Functions in anvl have `AnvlArray` inputs and R data inputs.
R data can be used dynamically, when it is passed to an argument that expects an `AnvlArray`, or statically, when it is passed to an argument that `jit()` marked as static.
To make this work, you should call `as_anvl_arrays()` on all dynamic input values.
If there is only a single one, you can also use `as_anvl_array()`.

They do four things, and three of them are decisions *across* the whole argument set, which is why a function with several arrayish arguments should normalize them in **one** `as_anvl_arrays()` call rather than one `as_anvl_array()` per argument:

1. every input is put on one device, and a call mixing devices or backends is rejected,
2. the type promotion rules of the `.promote` argument are applied, which handle both placement of R values, as well as conversions between
   `AnvlArray`s, e.g., to make `nv_add(nv_scalar(1, "f32"), nv_scalar(1, "f64"))` work.
3. afterwards, `dtype()` is well defined and the remaining body does not have to distinguish between R value inputs and `AnvlArray` inputs.
4. it reduces differences between eager and jit mode, see the `vignette("gotchas")` for a discussion.

Below, we add a function that performs the addition after coercing all inputs to `f32`;

```{r}
nv_add_f32 <- function(x, y) {
  args <- as_anvl_arrays(x, y, .promote = promote_dtype("f32", coerce = TRUE))
  do.call(prim_add, args)
}
nv_add_f32(1, nv_scalar(1, "f64"))
```

## Arbitrary Devices

In order to ensure that your function works with inputs from arbitrary devices, you need to be careful when creating new constants within your function.
Let's say you are creating your function and working on GPU:

```{r}
nv_add_one_naive <- function(x) {
  x <- as_anvl_array(x)
  x + nv_fill(1L, shape(x), device = "cuda")
}
```

As long as you are adding ones on a CUDA GPU, this function will work fine!
However, if you suddenly use it on the CPU, it will fail, because we can't add a CPU array to a CUDA array.
Constants should always be initialized on the same device as the inputs.
If there are multiple inputs and you called `as_anvl_arrays()` on them at the top, you know that there is only a single device.

One way to achieve this is to simply pass the input's device to `nv_fill()`:

```{r}
nv_add_one1 <- function(x) {
  # Converting is what gives the value a device to read back here.
  x <- as_anvl_array(x)
  x + nv_fill(1L, shape(x), device = device(x))
}
```

Another option is to rely on `nv_<op>_like` functions.
These take in another `AnvlArray` as their first input and use its properties as the defaults for their arguments.
In this case, the created array will assume the data type, shape and device from the input array.

```{r}
nv_add_one2 <- function(x) {
  x + nv_fill_like(x, 1L)
}
```

Note that when you only want to use a function with `jit()`, you can just omit specifying the device at all, as `jit()` is smart enough to place it on the correct device.

## Static Arguments to Enable Input Checks

One restriction of the XLA compiler is that it does not really allow for runtime checks.
Let's say you want to sample from a Bernoulli distribution with probability `p`.
If you make `p` a dynamic input, you can't check that it is within `[0, 1]`, so you need to make it a static input.
Don't convert it to an `AnvlArray` before checking its value.
Later in the function, it will actually be converted, but from XLA's point of view, it will just be a constant within the compiled program and not a dynamic input.

```{r}
nv_rbernoulli <- function(initial_state, p) {
  initial_state <- as_anvl_array(initial_state)
  stopifnot((p >= 0) && (p <= 1))

  # returns: (state, sample)
  out <- nv_runif(1L, initial_state)
  out_state <- out[[1L]]
  x <- nv_convert(out[[2L]] <= p, "i32")
  list(out_state, x)
}
nv_rbernoulli(nv_rng_state(1), 0.2)[[2L]]
```

## Jit-wrapping API Functions

Most user-facing API functions in anvl are wrapped in `jit(f, ...)` so that calling them traces and compiles a single program instead of executing each operation eagerly.
The wrapping is driven by the `@jit` roclet (see `?jit_roclet`).

In `R/api*.R`, tag any function that performs more than one primitive operation with `#' @jit`:

```r
#' @export
#' @jit
nv_log2 <- function(x) {
  x <- as_anvl_array(x)
  nv_log(x) / log(2)
}
```

If the function has static arguments (anything that is not an arrayish input -- axes, shape, dtype, control flags, functions used as templates, ...), list them with `static = c(...)` using either positional indices or argument names:

```r
#' @export
#' @jit static = c(2L, 3L)
nv_mean <- function(x, axes = NULL, drop = TRUE) {
  ...
}

#' @export
#' @jit static = c("axis")
nv_concatenate <- function(..., axis = NULL) {
  ...
}
```

Use names rather than positions whenever an argument lives after `...`, since `...` has no fixed position.

**When to skip `#' @jit`.** Don't tag a function whose body is essentially a single primitive call -- direct aliases (`nv_log <- prim_log`), `make_do_binary(prim_X)` factories, or thin wrappers that just validate and forward to one primitive. The underlying primitive is already jit-wrapped, so adding another `jit()` layer adds tracing overhead without fusing anything new.
Also skip pure I/O (`nv_save`, `nv_serialize`), backend constructors (`nv_array`, `nv_scalar`, `nv_matrix`), and device/state objects (`nv_device`, `nv_rng_state`).

The roclet writes the list of tagged functions to `R/jit-registry.R` on every `devtools::document()` run, and `R/zzz.R` applies that registry at package source time so the wrapped functions are byte-compiled with the rest of the package.
