---
title: "Gotchas"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Gotchas}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette lists various things to be aware of, specifically in relation to base R.

```{r, include = FALSE}
library(anvl)
```

## Row-major vs column-major ordering

R stores matrices and arrays in *column-major* order, while {anvl} (following XLA) uses *row-major* order.
For most operations, this is an internal implementation detail that does not change the semantics.
However, for reshaping operations such as `nv_flatten()` there is a difference.

Consider the 2x2 matrix below:

```{r}
m <- matrix(1:4, nrow = 2)
m
```

In base R, `as.vector()` flattens it column-by-column, so we get `1, 2, 3, 4`:

```{r}
as.vector(m)
```

In {anvl}, reshaping to a length-4 vector traverses the data row-by-row, so we get `1, 3, 2, 4`:

```{r}
nv_flatten(m)
```

If you need column-major flattening in {anvl}, transpose first:

```{r}
nv_flatten(t(m))
```

## No recycling

Base R *recycles* the shorter operand when two vectors of different lengths are combined elementwise:

```{r}
c(1, 2, 3, 4) + c(1, 2)
```

{anvl} only auto-broadcasts *scalars* (operands with shape `integer()`).
Adding a scalar to an array works as you would expect:

```{r}
nv_array(1:4) + 10L
```

But combining two non-scalar arrays of different shapes errors, even when one shape is a "tile" of the other:

```{r, error = TRUE}
nv_array(1:4) + nv_array(1:2)
```

When two non-scalar arrays differ only by size-1 axes (numpy-style broadcasting, e.g. shape `(2, 3)` and `(1, 3)`), use `nv_broadcast_arrays()` to align them explicitly first:

```{r}
a <- nv_matrix(1:6, nrow = 2)
shape(a)
b <- nv_matrix(c(10, 20, 30), nrow = 1)
shape(b)
xs <- nv_broadcast_arrays(a, b)
lapply(xs, shape)
xs[[1]] + xs[[2]]
```

Note that even `nv_broadcast_arrays()` cannot replicate R's recycling for shapes like `(4)` and `(2)` -- the shapes must be broadcast-compatible in the numpy sense.

## No `NA`s

R has a dedicated missing-value marker (`NA`) for every atomic type.
{anvl} arrays do not -- there is no representation of "missing" at the XLA level, only `NaN` for floating point numbers.
At a floating-point dtype, an `NA` silently turns into a `NaN`:

```{r}
nv_array(NA_real_)
```

```{r}
nv_array(c(1, NA, 3))
```

Round-tripping back to R is not guaranteed to produce `NA`, but can also yield `NaN`.
What comes back depends on the data type the value was built at, so it is pinned here:

```{r}
as_array(nv_array(c(1, NA, 3), dtype = "f32"))
```

The signed integer dtypes are a special case.
R spells `NA_integer_` as the bit pattern `-2147483648`, which is also what an `i32` array stores, so the value survives the round trip -- but on the device it is an ordinary, very negative integer that nothing distinguishes from a missing value.
Both directions warn about it:

```{r}
nv_scalar(NA_integer_, dtype = "i32")
```

```{r}
as.integer(nv_scalar(NA_integer_, dtype = "i32"))
```

Nothing on the device reserves that bit pattern, so it also arises from ordinary
arithmetic -- and such a value comes back to R as an `NA` too:

```{r}
as_array(nv_scalar(-2147483647L) - nv_scalar(1L))
```

At every other dtype there is no bit pattern for an `NA` to land on, so a missing value is an error, including for `bool`:

```{r, error = TRUE}
nv_scalar(NA)
```

```{r, error = TRUE}
nv_array(c(1L, NA, 3L), dtype = "i16")
```

Array creators take no argument to change any of this: what happens to an `NA` is fixed by the dtype you build at.
Scan the data yourself with `anyNA()` if you want to hear about a missing value the dtype accepts.

Converters like `as_array()` do take a `check` argument, with three levels: `"warn"` (the default) warns about a value R's type cannot hold, `"err"` makes it an error, and `FALSE` skips the scan.

```{r, error = TRUE}
as_array(nv_scalar(NA_integer_, dtype = "i32"), check = "err")
```

## Subnormal floating-point values

Subnormal values are extremely small nonzero numbers, very close to zero.
They allow floating-point numbers to extend below their normal range, but with fewer significant digits of precision.
For R's usual double-precision numbers (IEEE 754 binary64), subnormal magnitudes range from approximately $4.94 \times 10^{-324}$ to just below $2.23 \times 10^{-308}$.
More precisely, their magnitudes are at least $2^{-1074}$ but strictly less than $2^{-1022}$, which is the smallest positive normal value.

With {anvl}'s default PJRT backend, these tiny values can be stored in an array and read back into R unchanged, but may become zero when used in calculations.
On CPUs, XLA enables a mode that treats subnormal inputs and results as zero, often called "flushing to zero".
The exact behavior depends on the platform, backend, and operation.
Single-precision (`f32`) values are affected in the same way, below approximately $1.18 \times 10^{-38}$.
Comparisons flush their inputs too, so for example a negative subnormal can test as greater than or equal to zero.

Running the following example demonstrates this effect, with "Observed" comments corresponding to results from the PJRT CPU backend.

```{r}
library(anvl)

with_backend("pjrt", {
  x <- nv_scalar(1e-310, dtype = "f64")

  as.vector(x)  # Preserved: approximately 1e-310

  # Subnormal input; mathematically normal result (should give ~1e-210).
  as.vector(x * nv_scalar(1e100, dtype = "f64"))  # Observed: 0

  # Normal inputs; mathematically subnormal result (should give ~1e-310).
  as.vector(
    nv_scalar(1e-200, dtype = "f64") *
      nv_scalar(1e-110, dtype = "f64")
  )  # Observed: 0

  # Comparisons flush their inputs as well.
  as.vector(nv_scalar(-1e-310, dtype = "f64") >= 0)  # Observed: TRUE

  # The same happens in single precision.
  as.vector(nv_scalar(1e-40, dtype = "f32"))  # Preserved: approximately 1e-40
  as.vector(nv_scalar(1e-40, dtype = "f32") * 1e30)  # Observed: 0
})
```

If the first argument to `with_backend()` is changed to `"quickr"` -- the experimental, optional backend described in the [internals vignette](internals.html) -- then you should observe the results of subnormal computations are preserved correctly.

This is also present, for example, in the [JAX gotchas guide](https://docs.jax.dev/en/latest/notebooks/Common_Gotchas_in_JAX.html#miscellaneous-divergences-from-numpy) which also illustrates values preserved in storage but flushed during operations on some backends.

## No unsigned integers

R's `integer` type is signed 32-bit (range `-2147483648` to `2147483647`).
{anvl} also exposes unsigned integer dtypes (`ui8`, `ui16`, `ui32`, `ui64`) backed by XLA, but R has no native counterpart.
For values that fit into R's signed integer range, the round-trip works as expected:

```{r}
as_array(nv_array(c(0L, 200L, 255L), dtype = "ui8"))
```

Because `ui32` does not fit into R's native integer type, it will be converted to `bit64::integer64` data type:


```{r}
big <- nv_array(2147483647L, dtype = "ui32") + 1L
as_array(big)
```

However, for `ui64`, we also convert to `integer64`, which does not cover the whole range, so wrapping is possible.
`as_array()` warns about it, and `check = "err"` makes it an error:

```{r, error = TRUE}
big <- nv_array(0L, dtype = "ui64") - 1L
big
as_array(big)
as_array(big, check = "err")
```

## Differences between eager and jit-mode

We try to keep the semantics of eager and jit-mode as close as possible.
There are a few reasons why this might not be the case, which are listed here:

## The function does not canonicalize its inputs.

One difference arises when eager functions do not canonicalize their inputs.
Here, canoicalizing refers to the conversion or dynamic R inputs to `AnvlArray`s.
This can be done via `as_anvl_array` for single arguments and `as_anvl_arrays` for multiple arguments.

E.g., the following function does not behave the same in jit-mode and in eager mode.
The untyped `1` materializes at the default float, so under the default `f32` the
jitted result is rounded where the eager one is not (at an `f64` default the two
would agree, which is exactly the point: the answer depends on a default rather
than on the code):

```{r}
add_pi <- function(x) {
  x + pi
}
as.numeric(jit(add_pi)(1)) == add_pi(1)
```

We can fix this, by canonicalizing the inputs:

```{r}
add_pi2 <- function(x) {
  x <- as_anvl_array(x)
  x + pi
}
jit(add_pi2)(1) == add_pi2(1)
```

Not canonicalizing inputs can also be a problem for device placement.
If we were to call the function below as `threeway_add(1, 2, nv_scalar(3, "cuda"))`, then the
`1 + 2` would first move the `1` and `2` to the default device (which is CPU by default) and the second addition would then fail because it would attempt to call `nv_add` with mixed-device inputs.

```{r}
threeway_add <- function(x, y, z) {
  nv_add(nv_add(x, y), z)
}
```

Therefore, you should always canonicalize your inputs in eager functions.
