---
title: "Installation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Installation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Currently, {anvl} is not available on CRAN, so you either have to install it via [r-universe](https://r-xla.r-universe.dev/) or from [GitHub](https://github.com/r-xla/anvl).

The system library required during runtime is `libprotobuf`.
Source installation requires a `C++20` compiler and `protoc` (protobuf compiler).

You can install the latest release from r-universe (prebuilt binary).

```{r, eval = FALSE}
install.packages("anvl", repos = c("https://cloud.r-project.org", "https://r-xla.r-universe.dev"))
```

The development version can be installed via:

```{r, eval = FALSE}
pak::pak("r-xla/anvl")
```

Afterwards, you need to install additional dependencies via:

```{r, eval = FALSE}
anvl::install_anvl()
```

If you do not run this, interactive use of {anvl} will ask you for confirmation to download the additional dependencies.
You can opt into always downloading the additional required dependencies by configuring the `PJRT_INSTALL` variable:

| Value | Effect |
|-------|--------|
| `1` | Always download, without asking. Use this in CI, scripts and Docker builds. |
| `0` | Never download; abort with instructions instead. |
| unset | Ask in an interactive session, abort in a non-interactive one. |

## CUDA Setup

```{r, include = FALSE}
cuda_pkg <- getFromNamespace("cuda_r_package", "pjrt")()
cuda_version <- getFromNamespace("cuda_version", "pjrt")()
```

The additional dependencies that {anvl} installs includes the {`r cuda_pkg`} R package, which only requires a CUDA `r cuda_version`-compatible driver to be installed.

When the {`r cuda_pkg`} package is not installed, the correct runtime libraries need to be installed on the system and discoverable via `LD_LIBRARY_PATH`.
The specific versions of the CUDA runtime libraries provided with {`r cuda_pkg`} are listed [here](`r sprintf("https://github.com/r-xla/%s/blob/main/inst/components.tsv", cuda_pkg)`).

To trouble-shoot the CUDA installation, run the following in a new R session for maximum debug output.

```{r, eval = FALSE}
Sys.setenv(PJRT_DEBUG = "1", TF_CPP_MIN_LOG_LEVEL = "0")
anvl::nv_scalar(1, device = "cuda")
```

Note that if another package (such as {torch}) is using different CUDA versions, there might be some issues.
In this case, use separate R processes, e.g. via {mirai}.

## Docker

Prebuilt Docker images are available in [r-xla/docker](https://github.com/r-xla/docker).
This includes a CUDA and CPU build for amd64/x86-64 architecture:

### Available Images

| Image | Description |
|-------|-------------|
| `anvl-cpu` | CPU support, based on `rocker/r-ver` |
| `anvl-cuda` | GPU support with CUDA 12.8 |

Note that running the GPU container requires the [NVIDIA Container Toolkit](https://docs.nvidia.com/datacenter/cloud-native/container-toolkit/latest/install-guide.html) to be installed on the host.
Once installed (and the Docker daemon restarted), pass `--gpus all` to `docker run` to expose the host GPUs to the container:

```bash
docker run --rm -it --gpus all ghcr.io/r-xla/anvl-cuda:latest R
```

You can verify that the GPU is visible inside the container by running `nvidia-smi`, or from R:

```r
anvl::nv_scalar(1, device = "cuda")
```

### Tags

Each image is available with two tags:

| Tag | Description |
|-----|-------------|
| `:latest` | Built from the `main` branch (rebuilt daily) |
| `:release` | Built from the latest release |
