---
title: "Internals"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Internals}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


## Transforming Code

While a real anvil is made for reshaping metal, this package is a tool for reshaping code.
We refer to such a rewriting of code as a **transformation**, of which there are three types:

1. `R` $\rightarrow$ `AnvlGraph`:
   Generic `R` functions are too complicated to handle, so the first step in {anvl} is always to convert them into a computational `AnvlGraph` object via **tracing**.
   Such an `AnvlGraph` is similar to `Jaxpr` objects in JAX.
   It operates only on `GraphNode`s -- the graph's stand-ins for arrays -- and applies `AnvlPrimitive` operations to them.
1. `AnvlGraph` $\rightarrow$ `AnvlGraph`:
   It is possible to transform `AnvlGraph`s into other `AnvlGraph`s.
   Their purpose is to change the functionality of the code.
   At the time of writing, there is essentially only one such transformation, namely reverse-mode automatic differentiation via `gradient()`.
1. `AnvlGraph` $\rightarrow$ executable:
   In order to perform the actual computation, the `AnvlGraph` needs to be converted into an executable.
   The main backend is `"pjrt"` (via {stablehlo} and {pjrt}, compiling with XLA). There is also an experimental [quickr](https://github.com/t-kalinowski/quickr) backend.

### Tracing R Functions into Graphs

All functionality in the {anvl} package is centered around the `AnvlGraph` class.
While it is in principle possible to create `AnvlGraph`s by hand, these are usually created by tracing R functions.
In general, when we want to convert some code into another form (in our case, R code into an `AnvlGraph`), there are two approaches:

1. Static analysis, which would require operating on the abstract syntax tree (AST) of the code.
2. Dynamic analysis (aka "tracing"), which executes the code and records selected operations.

The former approach is followed by the {quickr} package, while we go with tracing.
We start with a simple yet illustrative example that either adds or multiplies two inputs `x` and `y` depending on the value of `op`.

```{r}
library(anvl)
f <- function(x, y, op) {
  if (op == "add") {
    nv_add(x, y)
  } else if (op == "mul") {
    nv_mul(x, y)
  } else {
    stop("Unsupported operation")
  }
}
```

To do this, we use `anvl::trace_fn()`, which takes in an `R` function and a list of `AbstractArray` inputs that specify the input types.

```{r}
aten <- nv_aval("f32", c())
aten
graph <- trace_fn(f, list(x = aten, y = aten, op = "mul"))
graph
```

The output of `trace_fn()` is now an `AnvlGraph` object that represents the computation.
The fields of the `AnvlGraph` are:

* `inputs`, which are `GraphValue`s that represent the inputs to the function.
* `outputs`, which are `GraphValue`s that represent the outputs of the function.
* `calls`, which are `PrimitiveCall`s that take in `GraphNode`s (and parameters) and produce output `GraphValue`s.
* `constants`, which are the `GraphValue`s for values closed over by the traced function (see *Constant Handling*).
* `in_tree`, `out_tree`, which record the nesting structure of the function's inputs and outputs.

What happens during `trace_fn()` is that a new `GraphDescriptor` is created and the inputs `x` and `y` are converted into `GraphBox` objects.
Then, the function `f` is simply evaluated with the `GraphBox` objects as inputs.
During this evaluation, we need to distinguish between two cases:

1. A "standard" `R` function is called:
   Here, nothing special happens and the function is simply evaluated.
2. An `anvl` function is called:
   Here, the operation that underlies the function is recorded in the `GraphDescriptor`.

The evaluation of the `if` statement is an example for the first category.
Because we set `op = "mul"`, only the second branch is executed.
Then, we are calling `nv_mul`, which attaches a `PrimitiveCall` that represents the multiplication of the two arrays to the `$calls` of the `GraphDescriptor`.
Note that `nv_mul` is itself not a primitive: it performs some type promotion and broadcasting if needed before calling into the primitive `prim_mul`.

A `PrimitiveCall` object consists of the following fields:

* `primitive`: The primitive function that was called.
* `inputs`: The inputs to the primitive function.
* `params`: The parameters (non-arrays) to the primitive function.
* `outputs`: The outputs of the primitive function.

When the evaluation of `f` is complete, the `$outputs` field of the `GraphDescriptor` is set and the `AnvlGraph` is subsequently created from the `GraphDescriptor`.
The only difference between the `AnvlGraph` and the `GraphDescriptor` is that the latter has some utility fields that are useful during graph creation, but for the purposes of this tutorial, you can think of them as being the same.

### Transforming Graphs into other Graphs

Once the `R` function is staged out into a simpler format, it is ready to be transformed.
The {anvl} package does not in any way dictate how such an `AnvlGraph` to `AnvlGraph` transformation can be implemented.
For most interesting transformations, however, we need to store some information for each {anvl} primitive function.
In the case of the gradient, we need to store the derivative rules.
For this, the `AnvlPrimitive` metadata object attached to each primitive has a `rules` field that can be populated.
The derivative rules are stored as functions under the `"reverse"` name.
Each primitive is an exported `prim_*` function; `[[` on it reads a rule:

```{r}
prim_mul[["reverse"]]
```

The `transform_gradient()` function uses these rules to compute the gradient of a function.
For this specific transformation, we walk the graph backwards and apply the derivative rules, which appends the "reverse pass" to the graph.
Besides the forward graph, the transformation takes in the `wrt` argument, which specifies with respect to which arguments to compute the gradient.

```{r}
bwd_graph <- transform_gradient(graph, wrt = c("x", "y"))
bwd_graph
```

### Lowering a Graph

In order to execute an `AnvlGraph`, we need to convert it into a -- wait for it -- executable.
Here, we show how to compile using the PJRT backend.
First, we will translate the `AnvlGraph` into the StableHLO representation via the {stablehlo} package.
Then, we will compile this program using the XLA compiler that is accessible via the {pjrt} package.

Like for the gradient transformation, the rules of how to do this transformation are attached to each primitive.

```{r}
prim_mul[["stablehlo"]]
```

The `stablehlo()` function creates a `stablehlo::Func` object and sequentially translates the `PrimitiveCall`s into StableHLO operations.

```{r}
func <- stablehlo(graph)[[1L]]
func
```

Now, we can compile the function via `pjrt_compile()`.

```{r}
hlo_str <- stablehlo::repr(func)
program <- pjrt::pjrt_program(src = hlo_str, format = "mlir")
exec <- pjrt::pjrt_compile(program)
```

To run the function, we need to extract the underlying buffers from the arrays before passing them to the executable, which will output a `PJRTBuffer` that we can easily convert to an `AnvlArray`.

```{r}
x <- nv_scalar(3, "f32")
y <- nv_scalar(4, "f32")
out <- pjrt::pjrt_execute(exec, x$data, y$data)
out
nv_array(out)
```

## The User Interface

In the previous section, we have shown how the transformations are implemented under the hood.
The actual user interface is a little more convenient and follows JAX's interface.

### `jit()`

The `jit()` function allows converting a regular `R` function into a just-in-time compiled function that can be executed on `AnvlArray`s.
We apply it to our simple example function, where we mark the non-array parameter `op` as "static".
This means that the value of this parameter needs to be known at compile time.

```{r}
f_jit <- jit(f, static = "op")
f_jit(x, y, "add")
```

One might think that `jit()` first calls `trace_fn()`, then runs `stablehlo()`, followed by `pjrt_compile()`.
This is, however, not what is happening, as this requires the input types to be known.
Instead, `f_jit` is a "lazy" function that will only perform these steps once the inputs are provided.
However, if those steps were applied every time the `f_jit` function is called, this would be very inefficient, because tracing and compiling take some time.
Therefore, the function `f_jit` also contains a cache (implemented as an `xlamisc::LRUCache`), which will check whether there is already a compiled executable for the given inputs.
For this, the types of all `AnvlArray`s need to match exactly (data type and shape) and all static arguments need to be identical.
For example, if we run the function with `AnvlArray`s of the same type, but different values, the function won't be recompiled, which we can see by checking the size of the cache, which is already 1, because we have called it on `x` and `y` above.

```{r}
cache_size <- function(f) environment(f)$cache$size
cache_size(f_jit)
```

After calling it with arrays of the same types and identical static argument values, the size of the cache remains 1:

```{r}
f_jit(nv_scalar(-99, "f32"), nv_scalar(2, "f32"), "add")
cache_size(f_jit)
```

When we execute the function with arrays of different `dtype` or `shape`, the function will be recompiled:

```{r}
f_jit(nv_scalar(1, "i32"), nv_scalar(2, "i32"), "add")
cache_size(f_jit)
```

Also, if we provide different values for static arguments, the function will be recompiled:

```{r}
f_jit(nv_scalar(1, "f32"), nv_scalar(2, "f32"), "mul")
cache_size(f_jit)
```


### `gradient()`

Just like `jit()`, `gradient()` also returns a function that will lazily create the graph and transform it, once the inputs are provided.

```{r}
g <- gradient(f, wrt = c("x", "y"))
```

To actually compute the gradient, we wrap it in `jit()`:

```{r}
g_jit <- jit(g, static = "op")
g_jit(x, y, "add")
```

We can also use `g` inside another function:

```{r}
h <- function(x, y) {
  z <- nv_add(x, y)
  g(z, x, "mul")
}
h_jit <- jit(h)
h_jit(x, y)
```

So, what is happening here?
Once the inputs `x` and `y` are provided to `h_jit`, a new `GraphDescriptor` is created and the inputs `x` and `y` are converted into `GraphBox` objects.
Then, the addition of `x` and `y` is recorded in the `GraphDescriptor`.
The call into `g()` is a bit more involved.
First, a new `GraphDescriptor` is created and the forward computation of `g` is recorded.
Subsequently, the reverse pass will be added to the descriptor, after which it will be converted into an `AnvlGraph`.
This `AnvlGraph` will then be inlined into the parent `GraphDescriptor` (representing the whole function `h`), which is then converted into the main `AnvlGraph`.
We can look at this graph below, where `trace_fn` internally converts the `AnvlArray`s `x` and `y` into their abstract representation.

```{r}
h_graph <- trace_fn(h, list(x = x, y = y))
h_graph
```

Afterwards, this graph is lowered to StableHLO and subsequently compiled.

## More Internals

### Constant Handling

Constants are handled specially in {anvl}.
Consider the program below:


```{r}
y <- nv_array(rnorm(1000000L))
graph <- trace_fn(function(x) {
  x + y + 1
}, list(x = nv_scalar(1L)))
graph
```

Here, `y` is a closed-over constant and it is included in the `$constants` field of the graph, just like the literal `1`.

```{r}
graph$constants
```

When compiling such a program to StableHLO, constants are treated differently depending on their shape (we follow JAX's approach here).
That is, constants with 1 element are **inlined** into the program, whereas other constants are added as inputs to the StableHLO program.
This is because inlining large constants into the executable is inefficient.
However, if we didn't inline small scalars, the compiler would be unable to do constant folding.

```{r}
out <- stablehlo(graph)
out[[1L]]
out[[2L]]
```

Also, before compiling, we remove unused constants.
Captured constants can become unused when we apply code transformations like below, where the gradient of the function w.r.t. `x` does not depend on the captured `y`:

```r
f <- function(x) {
x + y
}
transform_gradient(trace_fn(f, list(x = nv_scalar(1))))
```

In principle, the compiler is able to do this itself, but because we pass constants as inputs to the program, we need to handle it ourselves.

Further note that:

1. R literals are embedded directly into the program.
2. Currently, constants with the same value (that refer to different `AnvlArray`s) are not deduplicated, which we might change in the future.

### R Values Without a Data Type

An R value entering a program -- a length-1 vector or an `array()` -- has no data
type of its own. It is built into the program at the data type its use site
needs, from the R data itself, which is what keeps `x / sqrt(2)` exact for an
`f64` `x`. The machinery lives in `R/graph.R`.

A value written in the body of a traced function is built where it is used. An
*argument* of a jitted function cannot be: its value is unknown while tracing
(the cache keys on R storage type and shape only), so it becomes an input whose
aval is an `RData` -- an input like any other, except that it has no data type
yet -- and which data type it is supplied at is settled only once the body has
run. `finalize_rdata_inputs()` picks the narrowest data type that holds every
use site, stamps the input's aval with it and records the R storage type in
the graph's `rdata_types` beside it, and adds a
convert for each further data type the body asked for.

```{r}
graph <- trace_fn(
  function(x, v) list(x * v, nv_convert(v, "f32")),
  list(x = nv_aval("f64", integer()), v = nv_aval("double", integer()))
)
graph
```

The `f64[] <- double` marks an input the caller supplies as an R double and the
runtime uploads as `f64` -- rather than at the default for its R storage type.
`graph_input_dtypes()` reads those off for the backends.

A value is only ever built directly at a data type of its own R category: a
double at a float, an integer at an integer, a logical at `bool`. Any other
target is built at the category's natural type (`f64` / `i32` / `bool`) and
converted by the program, so narrowing follows XLA's semantics rather than R's.

## Device Inference in `jit()`

Device handling in `jit()` is quite complicated.
Some things that are important to be aware of:

1. We don't know the inferred device just from looking at the input, as we might have something like:
  `jit(\(x) x + nv_scalar(1, device = "cuda"))` where we might only learn about the device during tracing.
  This means the data is only converted at the end.
2. There are different backends.
   There might be a function like `jit(\(dev) nv_scalar(1, device = dev), backend = "auto")`.
   But with the current implementation of `jit()`, the tracing is handled by the backend's `jit` method,
   so we need to determine the backend from the input arguments. Therefore, the `device = device_arg("dev")`
   needs to be specified:

   ```r
   f <- jit(\(dev) nv_scalar(1, device = dev), backend = "auto", device = device_arg("dev"))
   f(nv_device("cpu", "pjrt"))
   ```

   If we made the main `jit()` function trace eagerly, we could determine the backend during
   tracing, but this is not really needed yet.

3. `device = device_arg()` is only accepted together with `backend = NULL` or `backend = "auto"`.
   A concrete backend combined with `device_arg()` is rejected, because with a concrete backend
   the device can simply be passed via a static argument.

### Nested Inputs and Outputs

TODO

## Dichotomy of anvl functions

Here, we will dig deeper into the dichotomy of {anvl} functions such as `prim_add`.
In the *Get Started* vignette, we have learned that these functions can either be called directly on `AnvlArray`s to transform data, or used within `jit()` blocks to build up programs.
Here, we will explain what this actually does and why this is possible.

The core problem this dichotomy solves is that it is a mental burden to always keep two versions of an {anvl} function:

1. The `jit()`ted version that can be used to transform arrays.
2. The non-`jit()`ted one that can be used to build up programs.

With our implementation, the following is possible:

```{r}
library(anvl)
prim_add(nv_scalar(1), nv_scalar(2))

times_2 <- jit(function(x) {
  nv_mul(x, 2)
})

times_4 <- jit(function(x) {
  times_2(times_2(x))
})

times_2(nv_scalar(2))
times_4(nv_scalar(2))
```

Otherwise, we would need the following:

```{r}
times_2_r <- function(x) {
  nv_mul(x, 2)
}
times_2_jit <- jit(times_2_r)
times_4 <- jit(function(x) {
  times_2_r(times_2_r(x))
})
```

This is rather cumbersome, as there are always two versions of a function and the first solution is preferable.
Internally, we have implemented this by wrapping every primitive function in `jit()` and making a `jit()`ted function behave differently depending on whether we are in another `jit()` call or not.

If we are in a `jit()` call and call into a `jit(f)`, then `f` is evaluated inline and re-traced.
Otherwise, the standard jit path is followed.

However, for the {anvl} API this now means that special care needs to be taken that everything
works in jit-mode and in eager-mode.
The most important points are:

1. Canonicalize inputs at the start using `as_anvl_array()` / `as_anvl_arrays()`.
2. Propagate the device from the inputs:
   1. For functions with dynamic inputs: use `nv_*_like` for constant creation and pass input operands
   2. For functions without dynamic inputs, add `device` arg and pass it to constant creators.
