---
title: "Logistic Regression"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Logistic Regression}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

In this vignette, we will implement logistic regression from scratch.

## Problem

Logistic regression models the probability that an observation belongs to a particular class.
Given a feature matrix $X$ and a weight vector $\beta$, the predicted probability is:

$$P(y = 1 | X) = \sigma(X \beta + \alpha)$$

  where $\sigma$ is the logistic function $\sigma(z) = \frac{1}{1 + e^{-z}}$ and $\alpha$ is the bias term.

In practice, logistic regression is typically fitted using Iteratively Reweighted Least Squares (IRLS), which is a form of Newton's method.
Here, we will use a simple gradient descept algorithm for didactic purposes.

## Data

We'll use the classic Titanic dataset from base R, which contains survival data for the most famous shipwreck in history.
The `Titanic` object is a contingency table, so we first expand it into individual observations.

```{r}
library(anvl)
set.seed(42)

# Expand contingency table to individual observations
titanic_df <- as.data.frame(Titanic)
titanic <- titanic_df[rep(seq_len(nrow(titanic_df)), titanic_df$Freq), 1:4]
rownames(titanic) <- NULL

# Create design matrix with dummy variables
X <- model.matrix(~ Class + Sex + Age, data = titanic)[, -1]
y <- as.integer(titanic$Survived == "Yes")

# Standardize features for better gradient descent convergence
X <- scale(X)

n <- nrow(X)
p <- ncol(X)
```

The dataset contains `r n` observations and `r p` predictor variables.
Our goal is to predict survival (`Survived`) based on three categorical features:

- **Class**: Passenger class (1st, 2nd, 3rd, or Crew) - encoded as dummy variables with 1st class as the reference category
- **Sex**: Male or Female - encoded with Male as the reference category
- **Age**: Child or Adult - encoded with Child as the reference category

```{r}
summary(titanic)
```

The general survival rate was `r sprintf("%.2f", mean(y) * 100)`%.

Now we convert the data to `AnvlArray`s.

```{r}
# `X` is numeric, so it takes the default float on its own; `y` is an R
# integer, which would take the default *integer* instead, so we name the float
# we want. Everything downstream follows, whatever the defaults are set to.
X_tensor <- nv_array(X)
y_tensor <- nv_array(y, dtype = default_float(), shape = c(n, 1L))
```

## Model

The logistic regression model consists of computing the linear combination of features and then applying the logistic function.

```{r}
predict_proba <- function(X, beta, alpha) {
  logits <- X %*% beta + alpha
  nv_logistic(logits)
}
```

For binary classification, we use the binary cross-entropy loss:

$$\mathcal{L} = -\frac{1}{n} \sum_{i=1}^{n} \left[ y_i \log(p_i) + (1 - y_i) \log(1 - p_i) \right]$$

We need to be careful with numerical stability when computing the logarithm of probabilities close to 0 or 1.
We'll add a small epsilon to avoid taking the log of exactly 0.

```{r}
binary_cross_entropy <- function(y_true, y_pred) {
  eps <- 1e-7
  y_pred_clipped <- nv_clamp(eps, y_pred, 1 - eps)
  loss <- -(y_true * log(y_pred_clipped) + (1 - y_true) * log(1 - y_pred_clipped))
  mean(loss)
}
```

We combine the prediction and loss computation into a single function.

```{r}
model_loss <- function(X, y, beta, alpha) {
  y_pred <- predict_proba(X, beta, alpha)
  binary_cross_entropy(y, y_pred)
}
```

Using {anvl}'s automatic differentiation, we can obtain the gradients of the loss with respect to the model parameters.

```{r}
model_loss_grad <- gradient(model_loss, wrt = c("beta", "alpha"))
```

## Training

We will implement the training loop using `nv_while()`.
This keeps the entire training loop within a single compiled function, which is more efficient than repeatedly calling a JIT-compiled function from R, especially for small models.

```{r}
fit_logreg <- jit(function(X, y, beta, alpha, n_epochs, lr) {
  output <- nv_while(
    list(beta = beta, alpha = alpha, epoch = nv_scalar(0L)),
    \(beta, alpha, epoch) epoch < n_epochs,
    \(beta, alpha, epoch) {
      grads <- model_loss_grad(X, y, beta, alpha)
      list(
        beta = beta - lr * grads$beta,
        alpha = alpha - lr * grads$alpha,
        epoch = epoch + 1L
      )
    }
  )
  list(beta = output$beta, alpha = output$alpha)
})
```

We initialize the parameters and train the model with a single function call.

```{r}
beta_init <- nv_array(rnorm(p), shape = c(p, 1L))
alpha_init <- nv_scalar(0)

result <- fit_logreg(
  X_tensor, y_tensor,
  beta_init, alpha_init,
  nv_scalar(50000L),
  nv_scalar(0.1)
)

result
```

Let's verify our implementation by comparing with R's built-in `glm()`.

```{r}
glm_fit <- glm(y ~ X, family = binomial)
```

Now let's compare the coefficients:

```{r, echo = FALSE}
anvl_coef <- c(as_array(result$alpha), as.vector(result$beta))
glm_coef <- coef(glm_fit)

# Clean up parameter names (remove "X" prefix from glm output)
param_names <- gsub("^X", "", names(glm_coef))

comparison <- data.frame(
  Parameter = param_names,
  anvl = anvl_coef,
  glm = glm_coef,
  row.names = NULL
)
comparison
```

The estimates are very close, confirming that our gradient descent implementation converges to the same solution as `glm()`.
