---
title: "Subsetting"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Subsetting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

In this vignette, you will learn how to subset arrays in {anvl} and how to update subsets.
Because array shapes in {anvl} programs are static (see the [Static Shape Restriction](static_shapes.html) vignette), only certain subsetting operations are supported and they come with some surprises.

We start by listing possible subsets and whether they support dynamic values (arrays that are specified during runtime) or only static values (e.g., R literals).

| Subset | Dynamic | Static |
|--------|---------|--------|
| Single Index | Yes | Yes |
| Multiple Indices | Yes | Yes |
| Range | No | Yes |
| Mask | No | No |

Ranges cannot have dynamic values, because then the size of the subset would be unknown (what's the size of `a:b` where `a` and `b` are unknown?).
Boolean masks are not supported, because the output shape depends on the data, which is not known at compile time.
For workarounds (e.g. `sum(x[x > 0])` or `x[mask] <- update`), see the [masking pattern](static_shapes.html#the-masking-pattern) section of the Static Shape Restriction vignette.
Negative indexing (e.g., `x[-1]` to exclude elements) is currently also not supported.
For static values, this will throw an error.
For dynamic values, negative indices are treated as out-of-bounds and clamped to the valid range (see [Out-of-bounds Handling](#out-of-bounds-handling) below).
If you are missing a feature, please open an issue on GitHub.

We will start with subsetting and then move on to subset-assignment.

## Subsetting

### Subsetting 1D arrays

Let's start with some simple examples of selecting individual elements from a 1-dimensional array.
The index can be either static or dynamic and we can drop or keep the axis:

```{r}
library(anvl)
x <- nv_array(1:10)
x
```

* Static & Drop:

  ```{r}
  x[2]
  ```

* Static & Keep:

  Here we use `arr()`, a convenience helper that builds an R array without having to wrap the values in `c()` (so `arr(2L)` is equivalent to `array(2L)`, and `arr(1, 2, 3)` to `array(c(1, 2, 3))`).

  ```{r}
  x[arr(2L)]
  ```

* Dynamic & Drop:

  ```{r}
  x[nv_scalar(2L)]
  ```

* Dynamic & Keep:

  Below, we perform almost the same operation as above, except that we use an array of shape `(1)` instead of a scalar with shape `()`.
  The difference is that subsetting with the former will preserve the axis, while the latter will drop it, as we have seen above.
  This ensures that the dimensionality of the result is the same for any 1D subset specification, and does not suddenly "simplify" the result to 0D.

  ```{r}
  x[nv_array(2L)]
  ```

Next, we subset multiple elements, where we only have to distinguish between static and dynamic indices.

* Static

  ```{r}
  x[arr(2, 4, 6)]
  ```

* Dynamic

  ```{r}
  x[nv_array(c(2L, 4L, 6L))]
  ```

We use `arr()` (or equivalently `array()`) instead of a bare R vector, because otherwise the case where we use a length-1 vector would be ambiguous (do we drop or keep the axis?).
This allows us to do without a `drop` parameter.

We can also use a range that can be specified either canonically via `a:b` or using `nv_seq()`.

```{r}
x[2:5]
x[nv_seq(2, 5)]
```

Note that the `a:b` syntax works via Non-Standard Evaluation (NSE), so we can distinguish it from the actual vector `2:5`.
Internally, it is translated to `nv_seq(a, b)`.

A range that counts down selects in reverse, as it does in base R.

```{r}
x[5:2]
```

It is also possible to select the whole range by omitting the specification altogether.

```{r}
x[]
```

### Subsetting higher-dimensional arrays

We start by creating a 2-dimensional array.

```{r}
x <- nv_matrix(1:12, nrow = 3, byrow = TRUE)
x
```

Combining subsets just works like one would expect.

```{r}
x[1, ]
x[1, 2]
x[arr(1), 2:3]
x[arr(1, 3), 2:3]
x[1:2, 2:3]
x[1, 2:3]
x[arr(2, 2), ]
x[arr(2, 2)]
```

### Out-of-bounds Handling

If one specifies out-of-bounds indices, we can only throw an error if the indices are static (and therefore known at compile time).
The PJRT backend that {anvl} compiles to does not throw errors for out-of-bounds dynamic indices, but instead clamps them to the valid range:

```{r}
x[nv_array(-1L), nv_array(100L)]
x[nv_array(1L), nv_array(4L)]
```

Therefore, you need to be careful when using dynamic indexing in order to avoid bugs.

## Updating Subsets

Updating subsets supports the same syntax as subsetting.
The value to write must either have the shape of the subset, or be a scalar.

```{r}
x
x[, 3] <- nv_array(-(1:3))
x
x <- nv_matrix(1:12, nrow = 3, byrow = TRUE)
x[, 3] <- -99L
x
```

Also, it must have a data type that is convertible to the data type of the array.

```{r, error = TRUE}
x <- nv_matrix(1:12, nrow = 3, byrow = TRUE)
x[, 3] <- nv_array(c(1.5, 2.5, 3.5))
x
```

### Out-of-bounds Handling

Similar to subsetting, out-of-bounds indices can only be checked for static values.
For dynamic indices, out-of-bounds writes are simply ignored:

```{r}
x <- nv_array(1:5)
x[nv_array(c(1L, 100L, 3L))] <- nv_array(c(-1L, -2L, -3L))
x
```

Here, the write to index 100 is silently ignored, while indices 1 and 3 are updated.

### Duplicate Indices

When writing to the same element multiple times, there is no guarantee which value will be written.
Specifically, this might differ between backends (CPU vs. GPU).

```{r}
x <- nv_array(1:5)
x[arr(1L, 1L, 1L)] <- nv_array(c(10L, 20L, 30L))
x
```

### In-place Updates

In eager mode, `x[i] <- val` creates a new array by default, even when R could update `x` in place.
This is because {anvl} cannot know whether another R variable still refers to the same array, which would otherwise be modified as well.
To avoid the copy, pass `inplace = TRUE` among the subscripts, which writes the update into the memory of `x`:

```{r}
x <- nv_array(1:5)
x[2:3, inplace = TRUE] <- 0L
x
```

`nv_subset_assign()` takes the same argument: `nv_subset_assign(x, 2:3, value = 0L, inplace = TRUE)`.

This consumes the original array: its memory is [donated](efficiency.html#donation) to the result, so every other R variable that referred to it can no longer be used afterwards.

```{r, error = TRUE}
x <- nv_array(1:5)
y <- x
x[1, inplace = TRUE] <- -1L
y
```

Inside a jitted function, `inplace = TRUE` is an error.
There, `y` would stay valid, so the same code would behave differently in eager and in jit mode.
The compiler avoids unnecessary copies inside `jit()` anyway, so `inplace` is not needed there.
