---
title: "Type Promotion"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Type Promotion}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  eval = Sys.getenv("PJRT_PLATFORM", "cpu") != "metal"
)
```

## Type Promotion Rules

When combining arrays of different types (e.g., adding an `f32` to an `i32`), {anvl} needs to determine a common type.
For example, below we are adding an `f32` to an `f64`, where the former is promoted to the latter's type, because it's more expressive.

```{r}
library(anvl)
jit(nv_add)(
  nv_scalar(1.0, dtype = "f32"),
  nv_scalar(1.0, dtype = "f64")
)
```

The type-promotion rules are inspired by JAX, and they are designed for execution on accelerators like GPUs, where one often wants speed instead of precision.

The rules are defined by the `common_dtype()` function.

```{r}
common_dtype("f64", "f32")
common_dtype("i64", "f32")
```

A table with the promotion rules is below.

```{r, echo = FALSE}
library(stablehlo)
dtypes <- c("bool", "i8", "i16", "i32", "i64", "ui8", "ui16", "ui32", "ui64", "f32", "f64")
tbl <- matrix(NA_character_, length(dtypes), length(dtypes))
for (i in seq_along(dtypes)) {
  for (j in seq_along(dtypes)) {
    tbl[i, j] <- repr(common_dtype(dtypes[i], dtypes[j]))
  }
}
rownames(tbl) <- dtypes
colnames(tbl) <- dtypes
knitr::kable(tbl, format = "html", caption = "Type promotion rules (row × column)")
```

## R Values Have No Data Type

Every array in an {anvl} program has a data type -- except the R values you write into it.
`1.5` is neither an `f32` nor an `f64`; `sqrt(2)` is a number, not a number *at some precision*.
So {anvl} does not give an R value a data type when it enters a program.
It keeps the R data, and builds the value into the program at the type its use site turns out to need.

```{r}
x <- nv_scalar(1, dtype = "f64")
as_array(x / sqrt(2)) == 1 / sqrt(2)
```

The result is exact to the last bit, because the `f64` division was handed `sqrt(2)` itself.
Had the literal been stamped `f32` on the way in, that `f64` would have been built from a number that had already lost half its digits.
The same holds for a value passed as an argument:

```{r}
f <- jit(function(a, b) a / b)
as_array(f(x, sqrt(2))) == 1 / sqrt(2)
```

This works for length-1 vectors and for `array()`s, in the body of a jitted function and as its arguments alike.

## What an R Value Becomes

An R value takes the data type of the array it meets, whatever that type is:

```{r}
dtype(nv_array(1L, dtype = "i8") + 1L)
dtype(nv_array(1, dtype = "f64") + 1L)
```

It does not drag a program to a wider type just because R stores its numbers as doubles -- which is what makes it safe to write `x + 1` without thinking about it.
Two rules limit the yielding, both because the type simply cannot hold the value:

1. A value from an R `double` meets an integer type: it stays a float (the default one).
2. Any value meets `bool`: it stays what it is.

```{r}
dtype(nv_array(1L, dtype = "i32") + 1.5)
dtype(nv_array(TRUE) + 1L)
```

When an R value meets nothing but other R values, there is nothing to take a type from, and it **commits to the default** for its R type:

* `double()` -> `f32`
* `integer()` -> `i32`
* `logical()` -> `bool`

```{r}
jit(\() list(1L, 1.0, TRUE))()
```

## Committing Happens Per Operation

A value commits the first time it takes part in an operation, and from then on it is an ordinary array with an ordinary type.
Nothing is remembered about where it came from.
This is the same model PyTorch uses for Python scalars, and it is worth knowing where its edge is:

```{r}
f <- jit(function(x) {
  y <- x * 2
  y + nv_scalar(1, dtype = "f64")
})
f(sqrt(2))
```

Here `x` is only ever combined with the literal `2`, so at that point the best {anvl} knows is "a float, no width demanded" and both commit to `f32`.
The `f64` arrives one step later, on `y`, when `x` is already spent.
If you need the wider type, say so where the value enters:

```{r}
f(nv_scalar(sqrt(2), dtype = "f64"))
```

## Asking for the Type of an R Value

Because an R value has no type until it is used, asking for one is an error -- just as `dtype(1.5)` has no answer outside a program either:

```{r, error = TRUE}
jit(function(x) dtype(x))(1)
```

Its *shape* is a different matter, and is answered as usual: `()` for a length-1 vector, `dim()` for an `array()`.
This also means the `nv_*_like()` family, which takes its result's type from its argument, needs an argument that has one:

```{r, error = TRUE}
jit(function(x) nv_fill_like(x, 3))(1)
```

```{r}
jit(function(x) nv_fill_like(x, 3))(nv_scalar(1))
```
