---
title: "Type Promotion"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Type Promotion}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  eval = Sys.getenv("PJRT_PLATFORM", "cpu") != "metal"
)
```

## Type Promotion Rules

When combining arrays of different types (e.g., adding an `f32` to an `i32`), {anvl} needs to determine a common type.
For example, below we are adding an `f32` to an `f64`, where the former is promoted to the latter's type, because it's more expressive.

```{r}
library(anvl)
jit(nv_add)(
  nv_scalar(1.0, dtype = "f32"),
  nv_scalar(1.0, dtype = "f64")
)
```

The type-promotion rules are inspired by JAX, and they are designed for execution on accelerators like GPUs, where one often wants speed instead of precision.

The rules are defined by the `common_dtype()` function.

```{r}
common_dtype("f64", "f32")
common_dtype("i64", "f32")
```

A table with the promotion rules is below.

```{r, echo = FALSE}
library(stablehlo)
dtypes <- c("bool", "i8", "i16", "i32", "i64", "ui8", "ui16", "ui32", "ui64", "f32", "f64")
tbl <- matrix(NA_character_, length(dtypes), length(dtypes))
for (i in seq_along(dtypes)) {
  for (j in seq_along(dtypes)) {
    tbl[i, j] <- as.character(common_dtype(dtypes[i], dtypes[j]))
  }
}
rownames(tbl) <- dtypes
colnames(tbl) <- dtypes
knitr::kable(tbl, format = "html", caption = "Type promotion rules (row × column)")
```

The biggest differentiator between our type system and the one from JAX is the handling of array objects from the host language, which is R in our case.
While introducing more complexity, these rules prevent the loss of precision present in JAX's type system.
We describe it below.

## R Values Have No Data Type

In {anvl}'s type system, R objects do not have a concrete data type.

```{r, error = TRUE}
dtype(1)
```

The type promotion table from above therefore does not apply to them.
However, it is possible to apply an {anvl} function to R values (length-1 vectors and arrays):

```{r}
nv_exp(1)
nv_add(1, nv_scalar(1, "f64"))
nv_add(1, 2)
```

It is therefore important to understand to understand the rules that govern the materialization of R objects as `AnvlArray`s.
Generally, there are two routes:

1. An R values it commited at its default data type (`double -> f32`, `integer -> i32`, `logical -> bool`).
   This is e.g. the case in unary functions such as `nv_exp`.
2. The R values data type is inferred from other arguments, as is the case of the `nv_add` call above.
   When no concrete data type is present, `nv_add(1, 2)` falls back to the default, which is `f32` for `double`s.

Note that these rules are not universal and exceptions exist.
Some functions, such as `nv_clamp`, prioritize the data type of a specific argument, in this case `x`, the value that is being clamped.
It fails if the the boundary values cannot be promoted to `x` without loss of precision:

```{r, error = TRUE}
nv_clamp(
  min_val = nv_scalar(0, "f64"),
  x = nv_scalar(0, "f32"),
  max_val = nv_scalar(0, "f64")
)
```

Functions document their behavior, so consult their respective help page for more information.

You can canonicalize and promote inputs to a function via the `.promote` field of `as_anvl_arrays()`.
It takes a `PromotionRule`, which is a special function that takes in the arguments and outputs a data type for each one.

```{r}
as_anvl_arrays(1, 2, .promote = promote_like("f64"))
```

One common rule is `promote_common()`, which is used by functions such as `nv_add` above.
It computes the common data type of the inputs.
In this case, it returns `f32`, which is the default data type of R `double`s.


```{r}
promote_fn <- promote_common()
args <- list(1, 2, nv_scalar(1L, "i8"))
promote_fn(args)
```

And when used in `as_anvl_arrys()` it moves the inputs there:

```{r}
as_anvl_arrays(args, promote_fn)
```

For more information about the available rules, see `?promotion_rule`.
