---
title: "Type Promotion"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Type Promotion}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  eval = !startsWith(Sys.getenv("ANVL_DEFAULT_DEVICE", "cpu"), "metal")
)
```

## Type Promotion Rules

When combining arrays of different types (e.g., adding an `f32` to an `i32`), {anvl} needs to determine a common type.
For example, below we are adding an `f32` to an `f64`, where the former is promoted to the latter's type, because it's more expressive.

```{r}
library(anvl)
nv_add(
  nv_scalar(1.0, dtype = "f32"),
  nv_scalar(1.0, dtype = "f64")
)
```

The rules are defined by the `common_dtype()` function.

```{r}
common_dtype("f64", "f32")
common_dtype("i64", "f32")
```

A table with the promotion rules is below.

```{r, echo = FALSE}
library(stablehlo)
dtypes <- c("bool", "i8", "i16", "i32", "i64", "ui8", "ui16", "ui32", "ui64", "f32", "f64")
tbl <- matrix(NA_character_, length(dtypes), length(dtypes))
for (i in seq_along(dtypes)) {
  for (j in seq_along(dtypes)) {
    # `ui64` and a signed integer have no common data type.
    tbl[i, j] <- tryCatch(as.character(common_dtype(dtypes[i], dtypes[j])), error = function(e) "--")
  }
}
# The column headers are bold as table headers; the row labels are cells, so
# they are emphasized here to match.
rownames(tbl) <- sprintf("<strong>%s</strong>", dtypes)
colnames(tbl) <- dtypes
knitr::kable(tbl, format = "html", escape = FALSE, caption = "Type promotion rules (row × column)")
```

Two integer types meet at a type that holds every value of both, so `i8` and `ui8` meet at `i16`.
The exception is `ui64`: no signed integer type holds its values, and an integer never becomes a float on its own, so a `ui64` and a signed integer have no common type at all (the `--` cells above).
This is where JAX and NumPy fall back to a float -- {anvl} instead asks you to say what you want, by converting one of the operands with `nv_convert()`.

The biggest differentiator between our type system and the one from JAX is the handling of array objects from the host language, which is R in our case.
While introducing more complexity, these rules prevent the loss of precision present in JAX's type system.
We describe it below.

## R Values Have No Data Type

In {anvl}'s type system, R objects do not have a concrete data type.

```{r, error = TRUE}
dtype(1)
```

The type promotion table from above therefore does not apply to them.
However, it is possible to apply an {anvl} function to R values (length-1 vectors and arrays):

```{r}
nv_exp(1)
nv_add(1, nv_scalar(1, "f64"))
nv_add(1, 2)
```

It is therefore important to understand the rules that govern the materialization of R objects as `AnvlArray`s.
Generally, there are two routes:

1. The R value is materialized at its default data type, which `default_dtypes()` reports: a `double` and an `integer` take the defaults of the active backend, configurable through the `anvl.default_dtypes` option, and a `logical` takes `bool`.
   This is e.g. the case in unary functions such as `nv_exp`.
2. The R value's data type is inferred from the other arguments, as in the `nv_add` call above.
   When no concrete data type is present, `nv_add(1, 2)` falls back to that default again.

Either way the value is *built at* that data type rather than converted into it, so one the data type cannot hold is an error:

```{r, error = TRUE}
jit(function(x) x + (-2L))(nv_scalar(1L, "ui8"))
```

Converting an array is a different matter and keeps XLA's wraparound semantics:

```{r}
nv_convert(nv_scalar(-2L, "i32"), "ui8")
```

Note that these rules are not universal and exceptions exist.
Some functions, such as `nv_clamp`, prioritize the data type of a specific argument, in this case `x`, the value that is being clamped.
It fails if the the boundary values cannot be promoted to `x` without loss of precision:

```{r, error = TRUE}
nv_clamp(
  x = nv_scalar(0, "f32"),
  min = nv_scalar(0, "f64"),
  max = nv_scalar(0, "f64")
)
```

Functions document their behavior, so consult their respective help page for more information.

You can canonicalize and promote inputs to a function via the `.promote` field of `as_anvl_arrays()`.
It takes a `PromotionRule`, which is a special function that takes in the arguments and outputs a data type for each one.

```{r}
as_anvl_arrays(1, 2, .promote = promotion_dtype("f64"))
```

One common rule is `promotion_common()`, which is used by functions such as `nv_add` above.
It computes the common data type of the inputs.
In this case, it returns the default data type of an R `double`, as neither input has one of its own.
We can override that default via `with_default_dtypes()`, which we use below to ask for double precision.
The two elements contrast what the override does and does not touch: `materialized` is an R literal with
no data type of its own, so it follows the new default, while `yielded` meets a typed `f32` array and
takes *its* data type -- a default only decides what happens when nothing else does.

```{r}
with_default_dtypes(c(float = "f64"), {
  list(
    materialized = jit(\() 1.0)(),
    yielded = nv_array(1, dtype = "f32") + 1.5
  )
})
```

Below, we compute the common data type of the arguments.


```{r}
promotion_fn <- promotion_common()
args <- list(1, 2, nv_scalar(1L, "i8"))
promotion_fn(args)
```

And when used in `as_anvl_arrays()` it moves the inputs there:

```{r}
do.call(as_anvl_arrays, c(args, list(.promote = promotion_fn)))
```

For more information about the available rules, see `?promotion_rule`.
