---
title: "Type Promotion"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Type Promotion}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  eval = Sys.getenv("PJRT_PLATFORM", "cpu") != "metal"
)
```

## Type Promotion Rules

When combining arrays of different types (e.g., adding an `f32` to an `i32`), {anvl} needs to determine a common type.
For example, below we are adding an `f32` to an `f64`, where the former is promoted to the latter's type, because it's more expressive.

```{r}
library(anvl)
jit(nv_add)(
  nv_scalar(1.0, dtype = "f32"),
  nv_scalar(1.0, dtype = "f64")
)
```

The type-promotion rules are inspired by JAX, and they are designed for execution on accelerators like GPUs, where one often wants speed instead of precision.

The rules are defined by the `common_dtype()` function.

```{r}
common_dtype("f64", "f32")
common_dtype("i64", "f32")
```

A table with the promotion rules is below.

```{r, echo = FALSE}
library(stablehlo)
dtypes <- c("bool", "i8", "i16", "i32", "i64", "ui8", "ui16", "ui32", "ui64", "f32", "f64")
tbl <- matrix(NA_character_, length(dtypes), length(dtypes))
for (i in seq_along(dtypes)) {
  for (j in seq_along(dtypes)) {
    tbl[i, j] <- as.character(common_dtype(dtypes[i], dtypes[j]))
  }
}
rownames(tbl) <- dtypes
colnames(tbl) <- dtypes
knitr::kable(tbl, format = "html", caption = "Type promotion rules (row × column)")
```

## R Values Have No Data Type

Every array in an {anvl} program has a data type -- except the R values you write into it.
`1.5` is neither an `f32` nor an `f64`; `sqrt(2)` is a number, not a number *at some precision*.
So {anvl} does not give an R value a data type when it enters a program.
It keeps the R data, and builds the value into the program at the type its use site turns out to need.

```{r}
x <- nv_scalar(1, dtype = "f64")
as_array(x / sqrt(2)) == 1 / sqrt(2)
```

The result is exact to the last bit, because the `f64` division was handed `sqrt(2)` itself.
The same holds for a value passed as an argument:

```{r}
f <- jit(function(a, b) a / b)
as_array(f(x, sqrt(2))) == 1 / sqrt(2)
```

This works for length-1 vectors and for `array()`s, in the body of a jitted function and as its arguments alike.

## What an R Value Becomes

An R value takes the data type of the array it meets, whatever that type is:

```{r}
dtype(nv_array(1L, dtype = "i8") + 1L)
dtype(nv_array(1, dtype = "f64") + 1L)
```

It does not drag a program to a wider type just because R stores its numbers as doubles -- which is what makes it safe to write `x + 1` without thinking about it.
That matters on accelerators: `f64` arithmetic is far slower than `f32` on most GPUs, and the Metal backend does not support it at all, so an `f32` program that silently became an `f64` one because of a literal would be a real cost.

Two rules limit the yielding, both because the type simply cannot hold the value:

1. A value from an R `double` meets an integer type: it stays a float (the default one).
2. Any value meets `bool`: it stays what it is.

```{r}
dtype(nv_array(1L, dtype = "i32") + 1.5)
dtype(nv_array(TRUE) + 1L)
```

The full rule, for each of the three R types against every data type it can
meet:

```{r, echo = FALSE}
r_types <- c(logical = "bool", integer = "i32", double = "f32")
tbl <- matrix(NA_character_, length(r_types), length(dtypes))
for (i in seq_along(r_types)) {
  for (j in seq_along(dtypes)) {
    tbl[i, j] <- as.character(promote_dt_rdata(as_dtype(r_types[[i]]), as_dtype(dtypes[[j]])))
  }
}
rownames(tbl) <- sprintf("%s()", names(r_types))
colnames(tbl) <- dtypes
knitr::kable(tbl, format = "html", caption = "What an R value becomes: R type (row) meeting a data type (column)")
```

When an R value meets nothing but other R values, there is nothing to take a type from, and it **commits to the default** for its R type:

* `double()` -> `f32`
* `integer()` -> `i32`
* `logical()` -> `bool`

```{r}
jit(\() list(1L, 1.0, TRUE))()
```

## Committing Happens Per Operation

A value commits the first time it takes part in an operation, and from then on it is an ordinary array with an ordinary type, which will not yield again.
It is worth knowing where that puts the edge:

```{r}
f <- jit(function(x) {
  y <- x * 2
  y + nv_scalar(1, dtype = "f64")
})
f(sqrt(2))
```

Here `x` is only ever combined with the literal `2`, so at that point the best {anvl} knows is "a float, no width demanded" and both commit to `f32`.
The `f64` arrives one step later, on `y`, when `x` is already spent.
If you need the wider type, say so where the value enters:

```{r}
f(nv_scalar(sqrt(2), dtype = "f64"))
```

## Which Type a Function Promotes To

The rules above say what *one* type two values meet at. Which of them an `nv_*` function applies is the function's own decision, stated once at the top of it: the arguments are canonicalized together with `as_anvl_arrays()`, and its `.promote` argument names the rule.

```{r}
# the common type of the inputs -- what a binary operator does
as_anvl_arrays(nv_array(1L), 1.5, .promote = promote_common())
# the type one particular argument already has
as_anvl_arrays(x = nv_array(1, dtype = "f64"), y = 1.5, .promote = promote_like("x"))
# a type the function names itself
as_anvl_arrays(nv_array(1L), 1.5, .promote = promote_dtype("f64"))
```

`nv_add()` is `promote_common()`: neither operand is privileged, so the table above decides.
`nv_clamp(min_val, x, max_val)` is `promote_like("x")`: the result is `x`'s type and the bounds come along.
`nv_rnorm(shape, initial_state, dtype = )` is `promote_dtype(dtype)`: the type is the function's own argument rather than one read off an input.

A function that would rather read the type off its arguments but has one in mind for the case where they bring none gives `promote_common()` a `fallback`.
It applies only where *every* argument is an R value, so an argument that has a type still wins:

```{r}
as_anvl_arrays(1, 2L, .promote = promote_common(fallback = "f64"))
as_anvl_arrays(nv_array(1L), 2L, .promote = promote_common(fallback = "f64"))
```

This is how `nv_rnorm(mean = , sd = )` settles on a type when called with neither: the sample is whatever `mean` and `sd` are, and `f32` where they are bare R values with nothing to say.

The two rules that *name* a target refuse an argument that type cannot hold, rather than narrowing it silently:

```{r, error = TRUE}
nv_clamp(0, nv_array(1L), 1.5)
```

Write the literal in the array's own category instead -- an R integer reaches any data type an integer can, a double only the floats:

```{r}
nv_clamp(0L, nv_array(1L), 2L)
```


`promote_yield()` is the fourth rule and the odd one out: it reads the target off the arguments that already have a type and *never converts them*, so only the R values move -- and only within their own category.
It is what the primitives use, because a primitive should not silently widen the array it was handed.

```{r, error = TRUE}
as_anvl_arrays(nv_array(1, dtype = "f64"), 1.5, .promote = promote_yield())
# an integer literal cannot become a float this way -- that is promotion, and
# belongs to the `nv_*` layer
as_anvl_arrays(nv_array(1, dtype = "f64"), 1L, .promote = promote_yield())
```

An argument that should stay out of the promotion is named out with `only`; it is still converted, just not to the target.
`nv_ifelse()` needs this, since its predicate has to stay a `bool`.
Groups of arguments that promote independently are combined with `promote_grouped()`.
See `?promote_rule` for the full set, and `vignette("extending_api")` for writing a function around them.

These four are not a closed set.
A rule is a *function* of the call's arguments that answers with the data type each one is brought to, so a promotion {anvl} has never heard of is written the same way its own are:

```{r}
# every argument at the widest float in the call, and never below f32
promote_widest_float <- function(args) {
  widths <- vapply(args, function(a) {
    dt <- peek_dtype(to_abstract(a))
    if (is_dtype_float(dt)) dtype_width(dt) else 0L
  }, integer(1))
  rep(list(as_dtype(paste0("f", max(c(32L, widths))))), length(args))
}
as_anvl_arrays(nv_array(1L), 2.5, nv_array(1, dtype = "f64"), .promote = promote_widest_float)
```

The arguments reach the rule as the caller passed them, R values still uncommitted -- which is what lets the rule decide what they become, and why it asks `peek_dtype()` rather than `dtype()`.
`promote_grouped()` mixes rules of either origin.

Whichever rule applies, an R value among the arguments is *built* at the target rather than converted to it -- which is what makes the exactness of the previous sections hold for `nv_*` functions too, and not just for a literal that happens to meet an array.

## Asking for the Type of an R Value

Because an R value has no type until it is used, asking for one is an error. This is
the same answer inside a program and outside one -- `dtype(1.5)` has none to give
either:

```{r, error = TRUE}
jit(function(x) dtype(x))(1)
```

```{r, error = TRUE}
dtype(1.5)
```

What you can ask is which type it *would* take if nothing claimed it, which
`peek_dtype()` answers without settling the value:

```{r}
peek_dtype(1.5)
peek_dtype(1L)
```

Its *shape* is a different matter, and is answered as usual: `()` for a length-1 vector, `dim()` for an `array()`.
This also means the `nv_*_like()` family, which takes its result's type from its argument, needs an argument that has one:

```{r, error = TRUE}
jit(function(x) nv_fill_like(x, 3))(1)
```

```{r}
jit(function(x) nv_fill_like(x, 3))(nv_scalar(1))
```
