---
title: "`nv_qnorm` vs JAX vs R — performance & accuracy"
subtitle: "Standard normal quantile: value & gradient, `log_p` on/off, 32- & 64-bit"
date: today
author: Opus 5 supervised by Louis Aslett
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-tools: true
    code-summary: "Show code"
    embed-resources: true
    theme: cosmo
    fig-width: 9
    fig-height: 7
    tbl-cap-location: top
execute:
  warning: false
  message: false
  cache: false
---

```{r}
#| label: setup
#| include: false

## Everything regenerates on `quarto render`; no result is hard-coded, and the
## document is self-contained -- it sources no local file and depends on nothing
## but installed packages.
##
## Portable configuration -- no machine-specific path is baked in. Override on
## the render line, e.g.
##   QNORM_PY=/path/to/venv/bin/python quarto render qnorm-jax-comparison.qmd
##
##   QNORM_PY     Python interpreter providing jax, numpy and google-benchmark.
##                Default: this repo's uv venv at
##                ../../../py-benchmarks/.venv/bin/python, else python3 on PATH.
##   QNORM_FULL   "0" for a fast reduced-grid preview; default runs the full grid.
##   QNORM_PERF   "0" skips the performance report entirely (it renders blank);
##                accuracy still runs. Independent of QNORM_FULL -- use it when
##                iterating on the accuracy panels without waiting on timings.
##   QNORM_JAX    "0" drops the JAX series (no Python needed at all).
##   QNORM_DENSE  random targets per variant in the dense sweep (default 2000).
##   DEVICE       Compute device for BOTH anvl and JAX: "cpu" (default) or "cuda"
##                for GPU. GPU needs the matching PJRT plugin (anvl) and a
##                CUDA-enabled jaxlib (JAX) installed on the target machine.
env_or <- function(var, default = "") {
  v <- Sys.getenv(var, unset = ""); if (nzchar(v)) v else default
}
FULL    <- env_or("QNORM_FULL", "1") != "0"
PERF    <- env_or("QNORM_PERF", "1") != "0"
USEJAX  <- env_or("QNORM_JAX",  "1") != "0"
DENSE_N <- as.integer(env_or("QNORM_DENSE", if (FULL) "2000" else "60"))
DEVICE  <- env_or("DEVICE", "cpu")
## anvl reads PJRT_PLATFORM in default_device(); nv_array()/jit() then place and
## compile on it. JAX is steered per-subprocess via JAX_PLATFORMS (run_py_chunk).
Sys.setenv(PJRT_PLATFORM = DEVICE)

suppressPackageStartupMessages({
  library(anvl)
  library(Rmpfr)
  library(jsonlite)
  library(bench)
  library(ggplot2)
  library(plotly)
  library(dplyr)
  library(tidyr)
  library(DT)
  library(htmltools)
})

WORK <- tempfile("qnorm_jax_"); dir.create(WORK)

## Python interpreter (needs jax, numpy, google-benchmark); env-overridable.
PY_BIN <- ""
if (USEJAX) {
  PY_BIN <- env_or("QNORM_PY")
  if (!nzchar(PY_BIN)) {
    repo_py <- "../../../py-benchmarks/.venv/bin/python"
    PY_BIN <- if (file.exists(repo_py)) repo_py else unname(Sys.which("python3"))
  }
  if (!nzchar(PY_BIN) || !file.exists(PY_BIN))
    stop("No Python interpreter found. Set QNORM_PY=/path/to/venv/bin/python ",
         "(needs jax, numpy, google-benchmark), or set QNORM_JAX=0.")
  if (!identical(0L, suppressWarnings(system2(PY_BIN,
        c("-c", shQuote("import jax, numpy, google_benchmark")),
        stdout = FALSE, stderr = FALSE))))
    stop("Python at '", PY_BIN, "' lacks jax / numpy / google-benchmark; ",
         "point QNORM_PY at an environment that has all three.")
}

## --- run an embedded (eval:false) python chunk in the project venv ----------
## system2() pastes `env` into the shell command unquoted, so any value with a
## space in it is parsed as a separate word and the call silently runs something
## else entirely. Quote the value half of each NAME=value pair.
quote_env <- function(kv) {
  i <- regexpr("=", kv, fixed = TRUE)
  paste0(substr(kv, 1L, i), shQuote(substring(kv, i + 1L)))
}
run_py_chunk <- function(label, envvars = character()) {
  src <- knitr::knit_code$get(label)
  py  <- file.path(WORK, paste0(label, ".py"))
  writeLines(src, py)
  status <- system2(PY_BIN, shQuote(py),
                    env = quote_env(c(envvars, paste0("JAX_PLATFORMS=", DEVICE))),
                    stdout = TRUE, stderr = TRUE)
  st <- attr(status, "status")
  if (!is.null(st) && st != 0)
    stop("python chunk '", label, "' failed:\n", paste(status, collapse = "\n"))
  invisible(status)
}

## --- shared configuration --------------------------------------------------
DTYPES <- c("f64", "f32")
SIZES  <- if (FULL) c(1, 10, 100, 1e3, 1e4, 1e5, 1e6, 1e7) else c(1, 100, 1e4)
SEED   <- 1L
PREC   <- 512L                 # bits of mpfr ground-truth precision
PERF_PASSES <- 3L              # interleaved randomised timing passes
REGIME_N <- if (FULL) 1e6 else 1e4        # size used by the regime panel
COMPILE_N <- if (FULL) 30011L else 3011L  # base length for fresh-shape compiles

## The two static-flag settings under test. `log_p` describes the *input*: a bare
## probability, or its logarithm. They are separate compiled programs (`log_p` is
## a static jit argument), and they reach very different parts of the domain, so
## every panel is faceted by them rather than pooled.
VARIANTS <- list(
  list(id = "value", log_p = FALSE, label = "log_p = FALSE"),
  list(id = "log",   log_p = TRUE,  label = "log_p = TRUE")
)
VAR_LABEL <- vapply(VARIANTS, `[[`, "", "label")
names(VAR_LABEL) <- vapply(VARIANTS, `[[`, "", "id")

M_ANVL <- "anvl nv_qnorm"
M_JAX  <- "JAX ndtri"
M_R    <- "R qnorm"
PAL <- c("#2c7fb8", "#d95f0e", "#444444")
names(PAL) <- c(M_ANVL, M_JAX, M_R)
METHODS <- if (USEJAX) c(M_ANVL, M_JAX, M_R) else c(M_ANVL, M_R)
```

```{python}
#| label: perf-jax-src
#| eval: false

# JAX performance for one (dtype, size, input regime): times the quantile value
# and value+grad, for both `log_p` settings, with google-benchmark. Parameters
# arrive via environment variables; a google-benchmark JSON is written for R.
import os
import jax
jax.config.update("jax_enable_x64", True)
import jax.numpy as jnp
import numpy as np
import google_benchmark as gb
from jax.scipy.special import ndtri

DT = {"f64": jnp.float64, "f32": jnp.float32}


# What JAX ships: jax.scipy.special.ndtri, a transcription of Cephes ndtri.
# There is no log-probability entry point anywhere in JAX (no `logppf`, no
# `ndtri_exp`), so the log variant can only be served by composing ndtri(exp(lp)).
def qfun(a, log_p):
    return ndtri(jnp.exp(a)) if log_p else ndtri(a)


# Input recipe, mirrored exactly by gen_inputs() on the R side. For timing only
# the *distribution* has to match, not the draws, so each language samples
# locally from the same band with a fixed seed.
def gen(regime, log_p, n, rng):
    if regime == "uniform":
        p = rng.uniform(0.0, 1.0, n)
    elif regime == "central":
        p = rng.uniform(0.2, 0.8, n)
    elif regime == "tail":
        p = 10.0 ** rng.uniform(-30.0, -3.0, n)
    elif regime == "deep tail":
        p = 10.0 ** rng.uniform(-300.0, -100.0, n)
    elif regime == "mixed":
        p = 10.0 ** rng.uniform(-300.0, 0.0, n)
    else:
        raise ValueError(regime)
    return np.log(p) if log_p else p


dt = DT[os.environ["PJ_DTYPE"]]
size = int(os.environ["PJ_SIZE"])
regime = os.environ["PJ_REGIME"]
rng = np.random.default_rng(int(os.environ["PJ_SEED"]))


def make_bench(fn, arr, pair):
    def b(state):
        while state:
            out = fn(arr)
            if pair:                        # (value, grad): materialise both
                np.asarray(out[0]); np.asarray(out[1])
            else:
                np.asarray(out)             # force host materialisation
    return b


for vid, log_p in (("value", False), ("log", True)):
    arr = jnp.asarray(gen(regime, log_p, size, rng), dtype=dt)
    # value+grad, not grad alone: reverse mode yields both from one forward pass,
    # so that is what a sampler step or an optimiser iteration actually costs.
    fns = (
        ("value", jax.jit(lambda z, lp=log_p: qfun(z, lp)), False),
        ("vgrad", jax.jit(jax.value_and_grad(
            lambda z, lp=log_p: jnp.sum(qfun(z, lp)))), True),
    )
    for op, fn, pair in fns:
        out = fn(arr)                       # warm up / compile
        np.asarray(out[0] if pair else out)
        gb.register(make_bench(fn, arr, pair), name=f"{vid}|{op}")

# `<n>x` fixes the iterations per repetition rather than letting a time budget
# choose it, so each repetition's real_time is a mean over a known, small batch
# and the minimum over repetitions is the same statistic the R side reports.
gb.main([
    "perf_jax", "--benchmark_format=json",
    f"--benchmark_out={os.environ['PJ_OUT']}", "--benchmark_out_format=json",
    f"--benchmark_min_time={os.environ['PJ_ITERS']}x",
    f"--benchmark_repetitions={os.environ['PJ_REPS']}",
])
```

```{python}
#| label: compile-jax-src
#| eval: false

# Wall-clock of the very first call on a shape never traced before: trace, lower,
# XLA compile, plus one (negligible) execution. Minimum over several distinct
# fresh shapes -- a single measurement varies by tens of percent and the first is
# systematically the highest.
import os, json, time
import jax
jax.config.update("jax_enable_x64", True)
import jax.numpy as jnp
import numpy as np
from jax.scipy.special import ndtri

DT = {"f64": jnp.float64, "f32": jnp.float32}


def qfun(a, log_p):
    return ndtri(jnp.exp(a)) if log_p else ndtri(a)


rng = np.random.default_rng(int(os.environ["PJ_SEED"]))
sizes = [int(s) for s in os.environ["PJ_SIZES"].split(",")]
rows = []
for dname, dt in DT.items():
    for vid, log_p in (("value", False), ("log", True)):
        for op in ("value", "vgrad"):
            best = None
            for n in sizes:                 # each a shape never compiled before
                u = rng.uniform(0.0, 1.0, n)
                arr = jnp.asarray(np.log(u) if log_p else u, dtype=dt)
                if op == "value":
                    fn = jax.jit(lambda z, lp=log_p: qfun(z, lp))
                else:
                    fn = jax.jit(jax.value_and_grad(
                        lambda z, lp=log_p: jnp.sum(qfun(z, lp))))
                t0 = time.perf_counter()
                out = fn(arr)
                jax.block_until_ready(out)
                t = (time.perf_counter() - t0) * 1000.0
                best = t if best is None else min(best, t)
            rows.append({"dtype": dname, "variant": vid, "op": op, "ms": best})
json.dump(rows, open(os.environ["PJ_OUT"], "w"))
```

```{python}
#| label: opcount-jax-src
#| eval: false

# StableHLO for the JAX side. `.lower(x).as_text()` is the same dialect anvl
# emits and at the same stage -- what the stack hands to XLA, before XLA's own
# inlining, CSE and fusion. The *text* is handed back to R, which counts both
# stacks with one parser, so neither can be counted by a rule the other isn't.
import os, json
import jax
jax.config.update("jax_enable_x64", True)
import jax.numpy as jnp
import numpy as np
from jax.scipy.special import ndtri

DT = {"f64": jnp.float64, "f32": jnp.float32}


def qfun(a, log_p):
    return ndtri(jnp.exp(a)) if log_p else ndtri(a)


n = int(os.environ["PJ_N"])
rows = []
for dname, dt in DT.items():
    for vid, log_p in (("value", False), ("log", True)):
        p = np.full(n, 0.3)
        arr = jnp.asarray(np.log(p) if log_p else p, dtype=dt)
        fns = (
            ("value", jax.jit(lambda z, l=log_p: qfun(z, l))),
            ("value+grad", jax.jit(jax.value_and_grad(
                lambda z, l=log_p: jnp.sum(qfun(z, l))))),
        )
        for op, fn in fns:
            lowered = fn.lower(arr)
            rows.append({"dtype": dname, "variant": vid, "op": op,
                         # what the stack hands to XLA ...
                         "mlir": lowered.as_text(),
                         # ... and what XLA made of it: the optimised, scheduled
                         # HLO of the executable that actually runs.
                         "hlo": lowered.compile().as_text()})
json.dump(rows, open(os.environ["PJ_OUT"], "w"))
```

```{python}
#| label: acc-jax-src
#| eval: false

# JAX on inputs supplied by R: the quantile and its elementwise derivative, for
# every requested (variant, dtype). Reads a spec JSON from PJ_SPEC, writes
# results to PJ_OUT. Everything is cast to float64 for lossless transport (f32 is
# a subset of f64); non-finite values become JSON null, which R reads as NA.
import os, json
import jax
jax.config.update("jax_enable_x64", True)
import jax.numpy as jnp
import numpy as np
from jax.scipy.special import ndtri

DT = {"f64": jnp.float64, "f32": jnp.float32}


def qfun(a, log_p):
    return ndtri(jnp.exp(a)) if log_p else ndtri(a)


def clean(a):
    """JSON has no Inf/NaN. Emit them as tagged strings rather than null: which
    one came back matters at the boundaries (silent +-Inf where NaN is correct is
    the worst outcome there is, because nothing downstream notices)."""
    a = np.asarray(a, np.float64)
    return [float(v) if np.isfinite(v)
            else ("nan" if np.isnan(v) else ("inf" if v > 0 else "-inf"))
            for v in a]


spec = json.load(open(os.environ["PJ_SPEC"]))
out = {"meta": {"jax": jax.__version__}, "data": {}}
for var in spec["variants"]:
    vid, log_p = var["id"], bool(var["log_p"])
    inputs = np.atleast_1d(np.asarray(var["inputs"], np.float64))
    out["data"][vid] = {}
    for dname in spec["dtypes"]:
        # R has already rounded the inputs to this dtype, so both stacks receive
        # bit-identical values; asarray here is exact.
        z = jnp.asarray(inputs, dtype=DT[dname])
        val = jax.jit(lambda a, lp=log_p: qfun(a, lp))(z)
        # grad of the summed output is the per-element derivative, since the
        # function is elementwise -- the same construction anvl differentiates.
        gr = jax.jit(jax.grad(lambda a, lp=log_p: jnp.sum(qfun(a, lp))))(z)
        out["data"][vid][dname] = {"value": clean(val), "grad": clean(gr)}
json.dump(out, open(os.environ["PJ_OUT"], "w"))
```

This document validates the new anvl `nv_qnorm()` against JAX's
`jax.scipy.special.ndtri` and base R's `stats::qnorm`, on
`r toupper(DEVICE)`. Every combination of **dtype** (32-/64-bit), **`log_p`**
(the input is a probability $p$, or its logarithm $\log p$) and **quantity** (the
value, or its derivative with respect to the input) is measured, for the standard
normal (`mean = 0`, `sd = 1`, `lower_tail = TRUE`).

Performance uses the R [`bench`](https://bench.r-lib.org) package and Python
[`google-benchmark`](https://github.com/google/benchmark); accuracy uses `Rmpfr`
at `r PREC`-bit precision as ground truth, obtained by a Newton solve in log
space (below). Three points of asymmetry between the implementations are worth
knowing before reading any panel:

* **JAX has no log-probability quantile at all** — no `logppf`, no `ndtri_exp` —
  so its `log_p = TRUE` series can only be `ndtri(exp(lp))`, which stops
  wherever `exp` does (around $\log p = -745$, i.e. $x \approx -38.5$).
* **base R has no automatic differentiation**, so its gradient series is the
  analytic derivative composed from native R: $\mathrm{d}x/\mathrm{d}p = 1/\phi(x)$,
  written `1 / dnorm(qnorm(p))`, and its log-input counterpart
  $\mathrm{d}x/\mathrm{d}\log p = p/\phi(x)$, written
  `exp(lp - dnorm(qnorm(lp, log.p = TRUE), log = TRUE))`.
* **base R is double precision only**, so it appears in the 64-bit panels and not
  the 32-bit ones.

```{r}
#| label: helpers

## --- input recipes, mirrored by gen() in the python chunks -----------------
gen_inputs <- function(regime, log_p, n) {
  p <- switch(regime,
    uniform     = runif(n),
    central     = runif(n, 0.2, 0.8),
    tail        = 10^runif(n, -30, -3),
    `deep tail` = 10^runif(n, -300, -100),
    mixed       = 10^runif(n, -300, 0),
    stop("unknown regime ", regime))
  if (log_p) base::log(p) else p
}
REGIMES <- c("central", "tail", "deep tail", "mixed")

## --- the three implementations, as timed and as evaluated -----------------
## nv_qnorm is tagged `@jit static c("lower_tail", "log_p")`, so it is already
## jit-compiled: calling it directly exercises the compiled path, and wrapping it
## in jit() again would measure something the user never runs.
anvl_value <- function(arr, log_p) as.vector(nv_qnorm(arr, log_p = log_p))
anvl_grad_fn <- function(log_p) {
  jit(gradient(function(p) nv_reduce_sum(nv_qnorm(p, log_p = log_p)), wrt = "p"))
}
anvl_value_grad_fn <- function(log_p) {
  jit(value_and_gradient(function(p) nv_reduce_sum(nv_qnorm(p, log_p = log_p)),
                         wrt = "p"))
}

r_value <- function(p, log_p) {
  if (log_p) stats::qnorm(p, log.p = TRUE) else stats::qnorm(p)
}
## No AD in base R: the derivative is the analytic one, composed from qnorm and
## dnorm, which is what an R user actually writes.
r_grad <- function(p, log_p, q = r_value(p, log_p)) {
  if (log_p) exp(p - stats::dnorm(q, log = TRUE)) else 1 / stats::dnorm(q)
}

## --- dtype arithmetic ------------------------------------------------------
DTYPE_EPS    <- c(f64 = 2^-52,  f32 = 2^-23)
DTYPE_BYTES  <- c(f64 = 8L,     f32 = 4L)
DTYPE_MINEXP <- c(f64 = -1022L, f32 = -126L)   # exponent of the smallest normal

## Round a double vector to `dtype` and back, i.e. exactly the value the kernel
## will receive. The reference must be evaluated *here* rather than at the f64
## input: the double->float rounding is the caller's, not the algorithm's, and
## charging it to the algorithm makes every implementation report the same error.
round_to_dtype <- function(x, dtype) {
  sz <- DTYPE_BYTES[[dtype]]
  if (sz == 8L) return(x)
  readBin(writeBin(x, raw(), size = sz), "double", size = sz, n = length(x))
}

## Relative error at full mpfr precision: promote the computed double, difference
## in mpfr, and only then collapse to a double.
rel_err <- function(computed, gold_m) {
  re <- rep(NA_real_, length(computed))
  ok <- is.finite(computed)
  if (any(ok))
    re[ok] <- as.numeric(abs(mpfr(computed[ok], PREC) - gold_m[ok]) / abs(gold_m[ok]))
  re
}

## Error in ulps of `dtype`. Relative error cannot distinguish "correctly
## rounded" from "sloppy"; <= 0.5 ulp is correctly rounded, <= 1 ulp faithful,
## and that is the currency a special function's accuracy is normally claimed in.
## Both the difference and the ulp size are formed in mpfr, so a reference too
## large for a double (dx/dp overflows f64 past |x| ~ 38) still gives a count.
ulp_vs_gold <- function(computed, gold_m, dtype) {
  out <- rep(NA_real_, length(computed))
  ok <- is.finite(computed)
  if (any(ok)) {
    e <- pmax(floor(as.numeric(log2(abs(gold_m[ok])))), DTYPE_MINEXP[[dtype]])
    ulp <- mpfr(2, PREC)^e * DTYPE_EPS[[dtype]]
    out[ok] <- as.numeric(abs(mpfr(computed[ok], PREC) - gold_m[ok]) / ulp)
  }
  out
}

slog <- function(z) sign(z) * log10(1 + abs(z))
size_label <- function(n) {
  ifelse(n >= 1e6, paste0(n / 1e6, "M"),
         ifelse(n >= 1e3, paste0(n / 1e3, "k"),
                format(n, scientific = FALSE, trim = TRUE)))
}
legend_below <- function(p) {
  plotly::layout(p, legend = list(orientation = "h", x = 0, xanchor = "left",
                                  y = -0.18, yanchor = "top"),
                 margin = list(b = 130))
}
```

# Performance

```{r}
#| label: perf-fns

## Robust minimum wall-clock per call, in ms. The minimum, not the mean or
## median: it is the statistic least contaminated by GC and scheduler noise.
## Host materialisation is forced by the caller.
bench_ms <- function(call_fn, min_time = 0.02, min_iterations = 2L,
                     max_iterations = 200L) {
  b <- bench::mark(call_fn(), min_iterations = min_iterations,
                   min_time = min_time, max_iterations = max_iterations,
                   check = FALSE, memory = FALSE, filter_gc = FALSE)
  as.numeric(b$min) * 1000
}

## Time a group of cells in several interleaved passes, in a fresh random order
## each pass, and keep each cell's minimum. The point is not more samples but
## *unconfounded* ones: measuring every cell once in a fixed order spreads a long
## render across the cells systematically, so the method that sorts first is
## always measured first and the largest size always last and hottest. `spread`
## (max/min across passes) comes back with the estimate, so a reader can tell a
## 10% difference between two methods from a 10% difference between two
## measurements of the same one.
bench_cells <- function(cells, passes = PERF_PASSES) {
  n <- length(cells)
  if (!n) return(list(min = numeric(), spread = numeric()))
  for (cl in cells) invisible(cl())          # compile + warm every cell first
  t <- matrix(NA_real_, nrow = n, ncol = passes)
  for (p in seq_len(passes)) for (i in sample.int(n)) t[i, p] <- bench_ms(cells[[i]])
  list(min = apply(t, 1L, min), spread = apply(t, 1L, max) / apply(t, 1L, min))
}

## Every R-side cell for one (dtype, size, regime): anvl value and value+grad for
## both variants, plus base R's (f64 only, since R has no single precision).
r_side_cells <- function(dt, n, regime) {
  meta <- list(); cells <- list()
  add <- function(md, fn) {
    meta[[length(meta) + 1L]] <<- md
    cells[[length(cells) + 1L]] <<- fn
  }
  for (v in VARIANTS) {
    set.seed(SEED)
    vec <- gen_inputs(regime, v$log_p, n)
    arr <- nv_array(vec, dtype = dt)
    local({
      a <- arr; lp <- v$log_p; vid <- v$id
      add(data.frame(dtype = dt, size = n, regime = regime, variant = vid,
                     method = M_ANVL, op = "value"),
          function() anvl_value(a, lp))
      vg <- anvl_value_grad_fn(lp)
      add(data.frame(dtype = dt, size = n, regime = regime, variant = vid,
                     method = M_ANVL, op = "value+grad"),
          function() { r <- vg(a); as.vector(r$value) + as.vector(r$grad[[1L]]) })
    })
    if (identical(dt, "f64")) local({
      p <- vec; lp <- v$log_p; vid <- v$id
      add(data.frame(dtype = dt, size = n, regime = regime, variant = vid,
                     method = M_R, op = "value"),
          function() r_value(p, lp))
      add(data.frame(dtype = dt, size = n, regime = regime, variant = vid,
                     method = M_R, op = "value+grad"),
          function() { q <- r_value(p, lp); q + r_grad(p, lp, q) })
    })
  }
  list(meta = dplyr::bind_rows(meta), cells = cells)
}

## One JAX subprocess per (dtype, size, regime); all four cells inside it.
jax_perf_rows <- function(dt, n, regime) {
  if (!USEJAX) return(NULL)
  out_json <- file.path(WORK, sprintf("perf_%s_%s_%d.json", dt,
                                      gsub(" ", "_", regime), n))
  ## Repetitions scale down with size, so the small-n cells -- where one call is
  ## microseconds -- still get a decent sample without the 1e7 cells taking all day.
  reps <- max(2L * PERF_PASSES, min(60L, as.integer(2e6 / max(n, 1L))))
  run_py_chunk("perf-jax-src", c(
    sprintf("PJ_DTYPE=%s", dt), sprintf("PJ_SIZE=%d", n),
    sprintf("PJ_REGIME=%s", regime), sprintf("PJ_SEED=%d", SEED),
    sprintf("PJ_OUT=%s", out_json), "PJ_ITERS=1", sprintf("PJ_REPS=%d", reps)))
  pj <- fromJSON(out_json, simplifyVector = FALSE)$benchmarks
  ## google-benchmark reports its own time_unit; do not assume ns.
  to_ms <- c(ns = 1e-6, us = 1e-3, ms = 1)
  it <- Filter(function(b) b$run_type == "iteration", pj)
  tms <- vapply(it, function(b) b$real_time * to_ms[[b$time_unit]], numeric(1))
  key <- vapply(it, function(b) b$run_name, character(1))
  tmin <- tapply(tms, key, min)
  ## Spread must describe the *estimator*, not the raw samples, or it is not
  ## comparable across stacks: the R-side figure is already a minimum over many
  ## iterations, so chunk JAX's repetitions into the same number of groups and
  ## take max/min of those minima.
  tspr <- tapply(tms, key, function(z) {
    g <- vapply(split(z, cut(seq_along(z), PERF_PASSES, labels = FALSE)), min, 0)
    max(g) / min(g)
  })
  do.call(rbind, lapply(names(tmin), function(k) {
    parts <- strsplit(k, "|", fixed = TRUE)[[1L]]
    data.frame(dtype = dt, size = n, regime = regime, variant = parts[1],
               method = M_JAX,
               op = if (parts[2] == "value") "value" else "value+grad",
               time_ms = tmin[[k]], spread = tspr[[k]])
  }))
}

perf_block <- function(dt, n, regime) {
  rs <- r_side_cells(dt, n, regime)
  timed <- bench_cells(rs$cells)
  md <- rs$meta
  md$time_ms <- timed$min
  md$spread <- timed$spread
  out <- dplyr::bind_rows(md, jax_perf_rows(dt, as.integer(n), regime))
  gc(FALSE)
  out
}
```

```{r}
#| label: perf-run

perf_df <- NULL
if (PERF) {
  rows <- list()
  for (dt in DTYPES) for (size in SIZES) {
    rows[[length(rows) + 1L]] <- perf_block(dt, as.integer(size), "uniform")
  }
  perf_df <- dplyr::bind_rows(rows)
  perf_df$variant_label <- VAR_LABEL[perf_df$variant]
}
```

```{r}
#| label: perf-intro
#| echo: false
#| results: asis
if (PERF) cat(sprintf(
  "Minimum wall-clock time **per call** over %d interleaved passes, as the input grows from a single element to %s. Within each (dtype, size) every R-side cell is timed once per pass in a fresh random order, so no implementation is systematically measured before another or later in the render; the JAX side reports the same statistic, a minimum over individual timings. Host materialisation is forced on all three (`as.vector()` in anvl, `np.asarray()` in JAX; base R is already on the host).\n\nInputs here are `runif(n)` — passed through `log()` for the `log_p = TRUE` variant — which is the call pattern of inverse-transform sampling. That never reaches past $x \\approx -5.4$ at any practical $n$, so it exercises only the central branch; the [regime panel](#sec-regimes) below feeds each branch directly instead of assuming the cost is the same.\n\nThe derivative row is **`value+grad`**, not the gradient alone: reverse mode produces both from one forward pass, so that is what a sampler step or an optimiser iteration actually costs, and the marginal cost of the derivative is that row minus the value row. Hover any point for exact values.\n",
  PERF_PASSES, format(max(SIZES), big.mark = ",", scientific = FALSE)))
if (!PERF) cat("*Performance report disabled (`QNORM_PERF=0`). Set `QNORM_PERF=1` to include it.*\n")
```

```{r}
#| label: perf-plot-fn

perf_plot <- function(dt) {
  d <- dplyr::filter(perf_df, dtype == dt)
  d$tt <- sprintf("%s | %s | %s\nsize %s: %.4g ms  (spread %.2fx)",
                  d$method, d$variant_label, d$op,
                  format(d$size, big.mark = ",", scientific = FALSE),
                  d$time_ms, d$spread)
  g <- ggplot(d, aes(size, time_ms, colour = method, group = method, text = tt)) +
    geom_line(linewidth = 0.5) + geom_point(size = 1.3) +
    facet_grid(variant_label ~ op, scales = "free_y") +
    scale_x_log10(breaks = SIZES, labels = size_label(SIZES)) +
    scale_y_log10() +
    scale_colour_manual(values = PAL) +
    labs(x = "input length", y = "time per call (ms)", colour = NULL,
         title = sprintf("%s-bit — time per call", sub("f", "", dt))) +
    theme_minimal(base_size = 12)
  ggplotly(g, tooltip = "text") |> legend_below()
}
```

## 64-bit

```{r}
#| label: perf-f64
if (PERF) perf_plot("f64")
```

## 32-bit

```{r}
#| label: perf-f32
if (PERF) perf_plot("f32")
```

## Input regime {#sec-regimes}

```{r}
#| label: perf-regimes-run

regime_df <- NULL
if (PERF) {
  rows <- list()
  for (dt in DTYPES) for (rg in REGIMES) {
    rows[[length(rows) + 1L]] <- perf_block(dt, as.integer(REGIME_N), rg)
  }
  regime_df <- dplyr::bind_rows(rows)
  regime_df$variant_label <- VAR_LABEL[regime_df$variant]
  regime_df$regime <- factor(regime_df$regime, levels = REGIMES)
}
```

```{r}
#| label: perf-regimes-intro
#| echo: false
#| results: asis
if (PERF) cat(sprintf(
  "All three implementations claim that their cost does not depend on where in the distribution the input lands — anvl and JAX because the kernel is branchless (every regime is evaluated and selected between), base R because its branches are cheap. Each panel holds the size fixed at %s elements and puts every element squarely in one regime: `central` ($p \\in [0.2, 0.8]$), `tail` ($p \\in [10^{-30}, 10^{-3}]$), `deep tail` ($p \\in [10^{-300}, 10^{-100}]$, reachable only with `log_p = TRUE` in 32-bit) and `mixed` ($p$ log-uniform over the whole range).\n",
  format(REGIME_N, big.mark = ",", scientific = FALSE)))
```

```{r}
#| label: perf-regimes-plot

regime_plot <- function(dt) {
  d <- dplyr::filter(regime_df, dtype == dt)
  d$tt <- sprintf("%s | %s | %s\n%s: %.4g ms  (spread %.2fx)",
                  d$method, d$variant_label, d$op, d$regime, d$time_ms, d$spread)
  g <- ggplot(d, aes(regime, time_ms, colour = method, group = method, text = tt)) +
    geom_line(linewidth = 0.5) + geom_point(size = 1.6) +
    facet_grid(variant_label ~ op, scales = "free_y") +
    scale_y_log10() +
    scale_colour_manual(values = PAL) +
    labs(x = NULL, y = "time per call (ms)", colour = NULL,
         title = sprintf("%s-bit — cost by input regime, n = %s",
                         sub("f", "", dt),
                         format(REGIME_N, big.mark = ",", scientific = FALSE))) +
    theme_minimal(base_size = 12) +
    theme(axis.text.x = element_text(angle = 20, hjust = 1))
  ggplotly(g, tooltip = "text") |> legend_below()
}
if (PERF) regime_plot("f64")
```

```{r}
#| label: perf-regimes-f32
if (PERF) regime_plot("f32")
```

## Compile time {#sec-compile}

```{r}
#| label: compile-run

compile_df <- NULL
if (PERF) {
  ## Fresh, never-before-compiled shapes -- distinct per repetition, or the
  ## executable cache serves the second one for free.
  csizes <- COMPILE_N + 7L * (0:2)
  rows <- list()
  for (dt in DTYPES) for (v in VARIANTS) for (op in c("value", "value+grad")) {
    ts <- vapply(csizes, function(n) {
      set.seed(SEED)
      arr <- nv_array(gen_inputs("uniform", v$log_p, n), dtype = dt)
      fn <- if (op == "value") {
        function(a) as.vector(nv_qnorm(a, log_p = v$log_p))
      } else {
        g <- anvl_value_grad_fn(v$log_p)
        function(a) { r <- g(a); as.vector(r$value); as.vector(r$grad[[1L]]) }
      }
      t0 <- Sys.time()
      invisible(fn(arr))
      as.numeric(difftime(Sys.time(), t0, units = "secs")) * 1000
    }, 0)
    rows[[length(rows) + 1L]] <- data.frame(
      dtype = dt, variant = v$id, op = op, method = M_ANVL, ms = min(ts))
  }
  if (USEJAX) {
    out_json <- file.path(WORK, "compile_jax.json")
    run_py_chunk("compile-jax-src", c(
      sprintf("PJ_SIZES=%s", paste(csizes, collapse = ",")),
      sprintf("PJ_SEED=%d", SEED), sprintf("PJ_OUT=%s", out_json)))
    cj <- fromJSON(out_json, simplifyVector = TRUE)
    cj$op <- ifelse(cj$op == "value", "value", "value+grad")
    cj$method <- M_JAX
    rows[[length(rows) + 1L]] <- cj[, c("dtype", "variant", "op", "method", "ms")]
  }
  compile_df <- dplyr::bind_rows(rows)
  compile_df$variant_label <- VAR_LABEL[compile_df$variant]
}
```

```{r}
#| label: compile-table
#| echo: false
if (!PERF) {
  htmltools::HTML("<em>Performance report disabled (<code>QNORM_PERF=0</code>).</em>")
} else {
  compile_df |>
    dplyr::transmute(dtype, variant = variant_label, op, method,
                     `compile (ms)` = ms) |>
    dplyr::arrange(dtype, variant, op, method) |>
    datatable(rownames = FALSE, filter = "top",
              options = list(pageLength = 8, scrollX = TRUE)) |>
    formatSignif("compile (ms)", 4)
}
```

The one cost a jit framework charges that a steady-state sweep never sees, and
that a user pays again on **every new input length**: trace, lower, XLA compile,
plus one negligible execution, on a shape never compiled before. Minimum over
three distinct fresh shapes — a single measurement varies by tens of percent and
the first is systematically the highest. Base R has no compile step and does not
appear.

## Graph size {#sec-opcount}

```{r}
#| label: opcount-run

## Non-constant StableHLO ops in the program each stack hands to XLA. A measure
## of algorithm weight that is immune to framework dispatch overhead, so unlike
## the timing panels it does not go stale as anvl's dispatch is optimised.
##
## Both stacks' MLIR is parsed by the one function below, so neither can be
## counted by a rule the other is not.
OPCOUNT_N <- 1000L

## Opcodes in a StableHLO module, with calls expanded.
##
## Counting op *occurrences in the text* is wrong the moment a module outlines
## anything: JAX lowers polynomial evaluation to a private `@polyval` function
## and calls it once per polynomial, so its Horner chain appears once in the text
## and runs five times, while anvl's tracer emits one flat graph with every chain
## written out. Textually that reads as JAX using a third of the ops; expanding
## each call by its callee's body compares the two programs rather than their
## formatting. Constants are excluded throughout -- they are data, not work.
## The two stacks do not print MLIR identically -- JAX writes
## `func.func public @main(%arg0: ...)`, anvl writes `func.func @main (%0: ...)`
## with no visibility keyword and a space before the paren, and anvl uses the
## quoted generic form `"stablehlo.add" (%1, %2)` for most ops. Both spellings
## have to be accepted, and a module that parses to *no* functions is a parser
## failure, not an empty graph: silently returning nothing here dropped every
## anvl row from this panel once already, leaving a table that looked fine and
## compared JAX against itself.
expand_ops <- function(txt) {
  lines <- strsplit(txt, "\n", fixed = TRUE)[[1L]]
  cur <- NA_character_; body <- list(); calls <- list()
  for (ln in lines) {
    m <- regmatches(ln, regexec(
      "func\\.func\\s+(?:public\\s+|private\\s+)?@([A-Za-z0-9_.]+)\\s*\\(",
      ln))[[1L]]
    if (length(m) == 2L) {
      cur <- m[2L]; body[[cur]] <- character(); calls[[cur]] <- character()
      next
    }
    if (is.na(cur)) next
    cm <- regmatches(ln, regexec(
      "\\bcall\"?\\s+@([A-Za-z0-9_.]+)\\s*\\(", ln))[[1L]]
    if (length(cm) == 2L) { calls[[cur]] <- c(calls[[cur]], cm[2L]); next }
    o <- regmatches(ln, gregexpr("(stablehlo|chlo)\\.[a-z_0-9]+", ln))[[1L]]
    o <- o[!grepl("constant", o)]
    if (length(o)) body[[cur]] <- c(body[[cur]], o)
  }
  if (!length(body))
    stop("expand_ops(): no `func.func` found -- the MLIR spelling has changed ",
         "and the op-count panel would silently omit this stack.")
  entry <- if ("main" %in% names(body)) "main" else names(body)[1L]
  active <- character()
  resolve <- function(fn) {
    if (is.null(body[[fn]]) || fn %in% active) return(character())  # cycle guard
    active <<- c(active, fn)
    out <- body[[fn]]
    for (callee in calls[[fn]]) out <- c(out, resolve(callee))
    active <<- setdiff(active, fn)
    out
  }
  resolve(entry)
}

anvl_mlir <- function(log_p, op, dt) tryCatch({
  fn <- if (op == "value") {
    function(z) nv_qnorm(z, log_p = log_p)
  } else {
    value_and_gradient(function(z) nv_reduce_sum(nv_qnorm(z, log_p = log_p)),
                       wrt = "z")
  }
  g <- trace_fn(fn, list(z = nv_aval(shape = OPCOUNT_N, dtype = dt)),
                mode = "toplevel")
  out <- stablehlo(g, donate = character(), donate_unaliased_outputs = TRUE,
                   platform = "cpu")
  paste(stablehlo::repr(out[[1L]]), collapse = "\n")
}, error = function(e) NA_character_)

op_rows <- list()
for (v in VARIANTS) for (dt in DTYPES) for (op in c("value", "value+grad")) {
  txt <- anvl_mlir(v$log_p, op, dt)
  if (is.na(txt)) next
  op_rows[[length(op_rows) + 1L]] <- data.frame(
    dtype = dt, variant = v$id, op = op, method = M_ANVL,
    opcode = expand_ops(txt))
}
if (USEJAX) {
  oj <- file.path(WORK, "opcount_jax.json")
  run_py_chunk("opcount-jax-src", c(sprintf("PJ_N=%d", OPCOUNT_N),
                                    sprintf("PJ_OUT=%s", oj)))
  jo <- fromJSON(oj, simplifyVector = TRUE)
  ## How many polynomial evaluations JAX's value graph performs, counted rather
  ## than asserted: every one is unconditional, and selected between afterwards.
  JAX_POLY <- local({
    k <- which(jo$dtype == "f64" & jo$variant == "value" & jo$op == "value")[1L]
    m <- gregexpr("call @polyval", jo$mlir[k], fixed = TRUE)[[1L]]
    if (m[1L] == -1L) NA_integer_ else length(m)
  })
  for (i in seq_len(nrow(jo))) {
    o <- expand_ops(jo$mlir[i])
    op_rows[[length(op_rows) + 1L]] <- data.frame(
      dtype = jo$dtype[i], variant = jo$variant[i], op = jo$op[i],
      method = M_JAX, opcode = o)
  }
}
opcode_df <- dplyr::bind_rows(op_rows)
opcode_df$variant_label <- VAR_LABEL[opcode_df$variant]

## Share of the graph that moves values into position rather than computing on
## them. Computed rather than transcribed into the prose below, so it cannot
## drift from the artifact.
SHAPING_OPS <- paste0("stablehlo.",
                      c("broadcast_in_dim", "convert", "slice", "reshape"))
op_slice <- function(m, dt) {
  dplyr::filter(opcode_df, method == m, dtype == dt, variant == "value",
                op == "value")
}
pct <- function(m, dt) {
  d <- op_slice(m, dt)
  if (!nrow(d)) return("n/a")
  sprintf("%.0f%%", 100 * mean(d$opcode %in% SHAPING_OPS))
}
## Total, or arithmetic only, for the value graph of one stack and dtype.
op_n <- function(m, dt, what = "total") {
  d <- op_slice(m, dt)
  if (!nrow(d)) return(NA_integer_)
  if (identical(what, "total")) nrow(d) else sum(!d$opcode %in% SHAPING_OPS)
}
## What the log_p variant costs anvl over the plain one, in ops.
LOGP_DELTA <- local({
  n <- table(dplyr::filter(opcode_df, method == M_ANVL, dtype == "f64",
                           op == "value")$variant)
  as.integer(n[["log"]] - n[["value"]])
})
```

### As handed to XLA

```{r}
#| label: opcount-table
#| echo: false
opcode_df |>
  dplyr::group_by(dtype, variant = variant_label, op, method) |>
  dplyr::summarise(`non-constant ops` = dplyr::n(),
                   arithmetic = sum(!opcode %in% SHAPING_OPS),
                   shaping = sum(opcode %in% SHAPING_OPS),
                   `chlo composites` = sum(startsWith(opcode, "chlo.")),
                   `distinct opcodes` = dplyr::n_distinct(opcode),
                   .groups = "drop") |>
  dplyr::arrange(dtype, variant, op, method) |>
  datatable(rownames = FALSE, filter = "top",
            options = list(pageLength = 8, scrollX = TRUE))
```

How much work each stack asks XLA to do, independent of what the framework
charges to dispatch it. Both sides are read at the
same stage (the StableHLO handed to XLA: `trace_fn()` → `stablehlo()` for anvl,
`jit(...).lower(x).as_text()` for JAX) and parsed by the same function, with
`stablehlo.constant` excluded because a constant is data rather than work. Base R
has no traced graph at all — it calls compiled C — so it has no entry here.

**Calls are expanded.** The two stacks format the same program differently: JAX
outlines polynomial evaluation into a private `@polyval` function and calls it
once per polynomial, whereas anvl's tracer writes every chain out into one flat
graph. Counting op occurrences in the text would therefore charge JAX for one
Horner chain where it runs `r if (USEJAX) "six" else "several"`, and made JAX
look like it used a third of anvl's ops. Each call is expanded by its callee's
body, so what is compared is the program rather than its formatting.

```{r}
#| label: opcount-note
#| echo: false
#| results: asis
if (USEJAX) cat(sprintf(
  "What the arithmetic column shows: the 32-bit coefficient sets cut anvl's value graph from %d arithmetic ops to %d, while JAX's stays at %d in both dtypes — it evaluates the same Cephes polynomials whatever the dtype, so narrowing buys it narrower arithmetic and nothing else. Two structural differences account for the rest. JAX evaluates **%d polynomials unconditionally** (all three rational pairs — central, tail, far tail) and selects afterwards; anvl selects between the tail and far-tail *coefficients* first, so it evaluates one tail rational rather than two. And anvl's `log_p = TRUE` graph costs only %d more op%s than its `log_p = FALSE` one, because the flag is static and the branch is resolved at trace time rather than compiled in.\n",
  op_n(M_ANVL, "f64", "arithmetic"), op_n(M_ANVL, "f32", "arithmetic"),
  op_n(M_JAX, "f64", "arithmetic"), JAX_POLY,
  LOGP_DELTA, if (abs(LOGP_DELTA) == 1L) "" else "s"))
```

### After XLA optimisation

```{r}
#| label: hlo-run

## The counts above are the program each stack *hands* to XLA. This is what XLA
## made of it: the optimised, scheduled HLO of the executable that actually runs,
## after inlining, constant folding, CSE and fusion.
##
## JAX exposes it directly (`.compile().as_text()`). anvl compiles through PJRT,
## which has no such accessor, so the program is re-run in a subprocess under
## XLA_FLAGS=--xla_dump_to and the dump is read back. A subprocess per cell, not
## one for all of them: XLA_FLAGS is process-wide and dumps every compile it
## sees, and this document compiles hundreds of graphs -- attribution would be
## guesswork otherwise.
SHAPING_HLO <- c("broadcast", "reshape", "bitcast", "copy", "convert", "slice",
                 "transpose", "get-tuple-element", "tuple")
STRUCTURAL_HLO <- c("fusion", "call", "custom-call")

## Post-optimisation HLO is a different syntax from StableHLO -- instructions
## read `%name = shape opcode(operands), metadata=...` -- so it needs its own
## parser. Calls are expanded exactly as before, and `parameter`/`constant` are
## dropped. Returns the whole program and, separately, the entry computation,
## whose `fusion` instructions are the kernels that actually get launched.
expand_hlo <- function(txt) {
  lines <- strsplit(txt, "\n", fixed = TRUE)[[1L]]
  cur <- NA_character_; entry <- NA_character_; body <- list(); calls <- list()
  for (ln in lines) {
    h <- regmatches(ln, regexec(
      "^(ENTRY\\s+)?(%[A-Za-z0-9_.-]+)\\s*\\(.*\\{\\s*$", ln))[[1L]]
    if (length(h) == 3L) {
      cur <- h[3L]; body[[cur]] <- character(); calls[[cur]] <- character()
      if (nzchar(h[2L])) entry <- cur
      next
    }
    if (is.na(cur)) next
    im <- regmatches(ln, regexec(
      "^\\s+(?:ROOT\\s+)?%[A-Za-z0-9_.-]+\\s*=\\s*.*?([a-z][a-z0-9-]*)\\(",
      ln, perl = TRUE))[[1L]]
    if (length(im) != 2L) next
    op <- im[2L]
    if (op %in% c("constant", "parameter")) next
    body[[cur]] <- c(body[[cur]], op)
    cm <- regmatches(ln, gregexpr("calls=(%[A-Za-z0-9_.-]+)", ln))[[1L]]
    if (length(cm)) calls[[cur]] <- c(calls[[cur]], sub("^calls=", "", cm))
  }
  if (is.na(entry))
    stop("expand_hlo(): no ENTRY computation -- the HLO dump format has changed.")
  active <- character()
  resolve <- function(fn) {
    if (is.null(body[[fn]]) || fn %in% active) return(character())
    active <<- c(active, fn)
    out <- body[[fn]]
    for (callee in calls[[fn]]) out <- c(out, resolve(callee))
    active <<- setdiff(active, fn)
    out
  }
  list(all = resolve(entry), entry = body[[entry]])
}

## Compile one cell in a fresh R subprocess with HLO dumping on, and read back
## the post-optimisation module. Returns NA if the plugin declined to dump.
anvl_hlo <- function(log_p, op, dt) tryCatch({
  dd <- file.path(WORK, sprintf("hlo_%s_%s_%s", dt, log_p, gsub("\\W", "", op)))
  dir.create(dd, showWarnings = FALSE, recursive = TRUE)
  script <- file.path(dd, "cell.R")
  writeLines(c(
    "suppressPackageStartupMessages(library(anvl))",
    sprintf("a <- nv_array(runif(%d), dtype = '%s')", OPCOUNT_N, dt),
    if (op == "value") {
      sprintf("invisible(as.vector(nv_qnorm(a, log_p = %s)))", log_p)
    } else {
      c(sprintf(paste0("vg <- jit(value_and_gradient(function(z) ",
                       "nv_reduce_sum(nv_qnorm(z, log_p = %s)), wrt = 'z'))"),
                log_p),
        "r <- vg(a)",
        "invisible(as.vector(r$value)); invisible(as.vector(r$grad[[1L]]))")
    }), script)
  system2(file.path(R.home("bin"), "Rscript"), shQuote(script),
          env = quote_env(c(
            sprintf("XLA_FLAGS=--xla_dump_to=%s --xla_dump_hlo_as_text", dd),
            paste0("PJRT_PLATFORM=", DEVICE))),
          stdout = FALSE, stderr = FALSE)
  ## cpu_ or gpu_ depending on the device. If more than one module was dumped,
  ## keep the largest -- an incidental transfer/convert module is tiny next to
  ## the kernel under test.
  f <- list.files(dd, pattern = "_after_optimizations\\.txt$", full.names = TRUE)
  if (!length(f)) return(NA_character_)
  paste(readLines(f[which.max(file.size(f))]), collapse = "\n")
}, error = function(e) NA_character_)

hlo_summ <- list(); hlo_codes <- list()
add_hlo <- function(txt, dt, vid, op, method) {
  if (is.na(txt)) return(invisible(NULL))
  r <- expand_hlo(txt)
  hlo_summ[[length(hlo_summ) + 1L]] <<- data.frame(
    dtype = dt, variant = vid, op = op, method = method,
    instructions = length(r$all),
    arithmetic = sum(!r$all %in% c(SHAPING_HLO, STRUCTURAL_HLO)),
    shaping = sum(r$all %in% SHAPING_HLO),
    kernels = sum(r$entry == "fusion"))
  hlo_codes[[length(hlo_codes) + 1L]] <<- data.frame(
    dtype = dt, variant = vid, op = op, method = method, opcode = r$all)
}

for (v in VARIANTS) for (dt in DTYPES) for (op in c("value", "value+grad")) {
  add_hlo(anvl_hlo(v$log_p, op, dt), dt, v$id, op, M_ANVL)
}
if (USEJAX) {
  for (i in seq_len(nrow(jo))) {
    add_hlo(jo$hlo[i], jo$dtype[i], jo$variant[i], jo$op[i], M_JAX)
  }
}
hlo_df <- dplyr::bind_rows(hlo_summ)
hlo_codes_df <- dplyr::bind_rows(hlo_codes)
if (nrow(hlo_df)) hlo_df$variant_label <- VAR_LABEL[hlo_df$variant]
```

```{r}
#| label: hlo-table
#| echo: false
if (!nrow(hlo_df)) {
  htmltools::HTML("<em>No post-optimisation HLO was available: the PJRT plugin did not honour <code>--xla_dump_to</code>.</em>")
} else {
  hlo_df |>
    dplyr::transmute(dtype, variant = variant_label, op, method,
                     instructions, arithmetic, shaping, kernels) |>
    dplyr::arrange(dtype, variant, op, method) |>
    datatable(rownames = FALSE, filter = "top",
              options = list(pageLength = 8, scrollX = TRUE))
}
```

The same program after XLA has finished with it — inlined, constant-folded,
CSE'd and fused — which is what actually executes. `kernels` counts the `fusion`
instructions in the entry computation, i.e. how many separate passes over the
data the executable makes.

```{r}
#| label: hlo-note
#| echo: false
#| results: asis
if (nrow(hlo_df) && USEJAX) {
  g <- function(m, dt, col) {
    d <- dplyr::filter(hlo_df, method == m, dtype == dt, variant == "value",
                       op == "value")
    if (!nrow(d)) NA_integer_ else d[[col]][1L]
  }
  ## `shaping` pools several opcodes; the sentence below names `broadcast`, so
  ## count that one rather than letting the pooled figure stand in for it.
  bcast <- function(m, dt) sum(dplyr::filter(
    hlo_codes_df, method == m, dtype == dt, variant == "value",
    op == "value")$opcode == "broadcast")
}
```

::: {.callout-note title="Every opcode, per graph and stage" collapse="true"}
```{r}
#| label: opcount-detail
#| echo: false
dplyr::bind_rows(
  dplyr::mutate(opcode_df[, c("dtype", "variant_label", "op", "method",
                              "opcode")],
                stage = "handed to XLA (StableHLO)"),
  if (nrow(hlo_codes_df)) {
    hlo_codes_df |>
      dplyr::mutate(variant_label = VAR_LABEL[variant],
                    stage = "after XLA (HLO)") |>
      dplyr::select(dtype, variant_label, op, method, opcode, stage)
  }) |>
  dplyr::count(stage, dtype, variant = variant_label, op, method, opcode,
               name = "count") |>
  dplyr::arrange(stage, dtype, variant, op, method, dplyr::desc(count)) |>
  datatable(rownames = FALSE, filter = "top",
            options = list(pageLength = 15, scrollX = TRUE))
```
:::

# Accuracy

```{r}
#| label: gold

## --- 512-bit Rmpfr ground truth -------------------------------------------
##
## Phi^-1 is obtained by Newton in *log* space, log Phi(Q) = lp, which stays
## well-scaled at any depth: a direct Phi(Q) = p iteration dies once p underflows.
## Seeded from base R's qnorm, which is accurate enough to converge in a few
## steps; convergence is asserted below via the returned relative residual.

## MPFR's default exponent floor is 2^(1-2^30), so erfc(-Q/sqrt(2)) underflows to
## zero -- silently, giving log(0) = -Inf -- once Q^2/2 exceeds ~7.4e8, i.e. at
## |Q| ~ 3.86e4, well inside the range a log-probability input can reach. Widen
## as far as the build allows; this is global MPFR state and widening it can only
## remove spurious underflow.
for (.e in c(2^61, 2^40, 2^31)) {
  suppressWarnings(.mpfr_erange_set("Emin", -.e))
  if (.mpfr_erange("Emin") <= -.e) break
}

pi_m <- Const("pi", PREC)
s2   <- sqrt(mpfr(2, PREC))
l2pi <- log(2 * pi_m)

## Even widened, the erfc route dies at log Phi ~ 2^Emin. Past that, switch to the
## Mills asymptotic, which has no such limit. 0.9 keeps a margin.
Q_ERFC_MAX <- sqrt(2 * log(2) * abs(as.numeric(.mpfr_erange("Emin"))) * 0.9)

## log(1 - 1/t^2 + 3/t^4 - 15/t^6 + ...), the bracket of the Mills-ratio
## asymptotic for the lower tail. Only ever called at t > Q_ERFC_MAX (~1.7e9),
## where successive terms fall by ~19 decades each. The loop stops on term
## magnitude rather than a fixed count, so it stays correct if the switch moves.
mills_log <- function(t) {
  neg_it2 <- -1 / (t * t)
  term <- mpfr(rep(1, length(t)), PREC)
  s <- term
  tiny <- mpfr(2, PREC)^(-(PREC + 16L))
  for (k in seq_len(200L)) {
    term <- term * (2 * k - 1) * neg_it2
    s <- s + term
    if (max(abs(term)) < tiny) break
  }
  log(s)
}

## log Phi, cancellation-free on both sides and valid at any depth. The deep
## branch never materialises Phi itself, which is the point: the direct form
## underflows long before the quantile domain does (a log-probability input
## reaches |x| ~ 1.9e154).
logPhi_m <- function(Q) {
  out <- Q
  fin <- is.finite(Q)
  deep <- fin & Q < -Q_ERFC_MAX
  lo <- fin & !deep & Q <= 0
  hi <- fin & !deep & Q > 0
  if (any(deep)) {
    t <- -Q[deep]
    out[deep] <- -t * t / 2 - l2pi / 2 - log(t) + mills_log(t)
  }
  if (any(lo)) out[lo] <- log(0.5 * erfc(-Q[lo] / s2))
  if (any(hi)) out[hi] <- log1p(-0.5 * erfc(Q[hi] / s2))
  out
}
logphi_m <- function(Q) -Q * Q / 2 - l2pi / 2
Phi_m <- function(x) 0.5 * erfc(-x / s2)

## Newton on g(Q) = log Phi(Q) - lp, with g'(Q) = phi(Q)/Phi(Q).
solve_q <- function(lp, seed) {
  Q <- mpfr(seed, PREC)
  for (k in 1:8) Q <- Q - (logPhi_m(Q) - lp) / exp(logphi_m(Q) - logPhi_m(Q))
  Q
}

## Deep lower tail, where log Phi(Q) and lp are both ~ -Q^2/2 and Newton's
## `log Phi(Q) - lp` subtracts two nearly equal huge numbers -- costing
## log2(Q^2 / (2 log|Q|)) bits, so by |Q| ~ 2e78 nothing 512 bits can hold
## survives. Solve instead for u = Q^2/2, where the same equation reads
##   u = s - log(4 pi u)/2 + mills_log(sqrt(2u)),   s = -lp,
## and the correction is subtracted *from* a large number instead of *between*
## two of them, so u keeps full relative precision at any depth. The map
## contracts like 1/(2u); the last step is a conservative stand-in for the error.
solve_deep <- function(s) {
  u <- s; du <- s
  for (k in 1:6) {
    u_new <- s - log(4 * pi_m * u) / 2 + mills_log(sqrt(2 * u))
    du <- u_new - u
    u <- u_new
  }
  t <- sqrt(2 * u)
  list(Q = -t,
       ## dQ/dlp = Phi/phi = M(t)/t, formed directly: the exp(lp - logphi) route
       ## subtracts two ~Q^2/2 quantities and loses the lot.
       grad = exp(mills_log(t)) / t,
       rel = as.numeric(abs(du) / (2 * u)))          # |dQ/Q| = |du| / (2u)
}

## The double-precision input that targets quantile x.
gold_inputs <- function(x, log_p) {
  xm <- mpfr(x, PREC)
  as.numeric(if (log_p) logPhi_m(xm) else Phi_m(xm))
}

## Value and derivative at the *exact* input supplied (not at the nominal grid
## point), plus the solver's own relative residual. Must tolerate inputs the f64
## grid never produces -- a probability that has rounded to 0 or 1, or a log
## probability rounded to 0 -- since it is called once per dtype at the narrowed
## input; those are marked NaN and dropped by the caller's `keep` filter.
gold_eval <- function(input, log_p) {
  lp_d <- if (log_p) input else base::log(input)
  ok <- is.finite(lp_d) & lp_d < 0                    # strictly 0 < p < 1
  n <- length(input)
  value <- mpfr(rep(NaN, n), PREC)
  grad  <- mpfr(rep(NaN, n), PREC)
  residual <- rep(0, n)
  if (any(ok)) {
    lp <- if (log_p) mpfr(input[ok], PREC) else log(mpfr(input[ok], PREC))
    m <- sum(ok)
    Q <- mpfr(rep(0, m), PREC); gr <- mpfr(rep(0, m), PREC); res <- numeric(m)
    deep <- lp < -(Q_ERFC_MAX^2 / 2)                  # same split as logPhi_m
    if (!all(deep)) {
      pin <- input[ok][!deep]
      seed <- if (log_p) stats::qnorm(pin, log.p = TRUE) else stats::qnorm(pin)
      q <- solve_q(lp[!deep], seed)
      Q[!deep] <- q
      ## dQ/dp = 1/phi(Q); dQ/dlog p = Phi(Q)/phi(Q) = exp(lp - logphi(Q)).
      gr[!deep] <- if (log_p) exp(lp[!deep] - logphi_m(q)) else exp(-logphi_m(q))
      ## The residual must be *relative*: Newton's is in log Phi, so divide by
      ## dlogPhi/dQ to get dQ, then by |Q|. An absolute bound would be
      ## meaningless -- at |Q| ~ 1e15 a perfect 512-bit root still leaves
      ## |dlogPhi| ~ 1e-124, and by |Q| ~ 1e154 it exceeds 1.
      dQ <- (logPhi_m(q) - lp[!deep]) * exp(logPhi_m(q) - logphi_m(q))
      res[!deep] <- as.numeric(abs(dQ) / abs(q))
    }
    if (any(deep)) {
      d <- solve_deep(-lp[deep])
      Q[deep] <- d$Q
      gr[deep] <- d$grad                # deep is only reachable with log_p = TRUE
      res[deep] <- d$rel
    }
    value[ok] <- Q; grad[ok] <- gr; residual[ok] <- res
  }
  list(value = value, grad = grad, residual = residual)
}
```

```{r}
#| label: acc-grid

## --- the accuracy grid, specified in target quantiles x --------------------
##
## Specify the grid in terms of the *output*, and derive the input from it
## through the mpfr reference. Specifying inputs directly is how a grid ends up
## silently never reaching the tail: a linear or even log-uniform ladder in p
## cannot place a point at x = -1e100.

## Dense through the body, then a geometric ladder into each tail.
tail_ladder <- function(to, dense_to = 3, dense_by = 0.2, n_tail = 28) {
  if (to <= dense_to) return(seq(dense_by, to, by = dense_by))
  c(seq(dense_by, dense_to, by = dense_by),
    exp(seq(log(dense_to * 1.15), log(to), length.out = n_tail)))
}
grid_symmetric <- function(to, upper_to = to, n_tail = 28, n_tail_upper = n_tail) {
  sort(unique(c(-tail_ladder(to, n_tail = n_tail),
                tail_ladder(upper_to, n_tail = n_tail_upper))))
}
## Pin a regime boundary with a point either side of it, and one on it. A ladder
## that merely passes within a few percent of a seam does not test it: both
## branches agree there, and the failures worth finding are visible only across
## the cut itself.
straddle <- function(z, rel = 1e-9) sort(c(z * (1 - rel), z, z * (1 + rel)))

## The Cephes branch cuts, as target quantiles |x|, derived from the thresholds
## the code actually tests rather than typed as decimals. Both anvl and JAX
## evaluate the same two regimes on the smaller tail probability t:
##   t >  e^-2     central rational in w = t - 1/2
##   t <= e^-2     tail rational in 1/z, z = sqrt(-2 log t), refitted at z = 8.
r_seam <- function(rp) abs(stats::qnorm(-rp^2, log.p = TRUE))
SEAMS <- c(r_seam(sqrt(2)), r_seam(sqrt(32)))    # central<->tail, z = 8
NEAR_CENTRE <- c(0.001, 0.01, 0.05)              # approaches the p = 1/2 reflection

## Reachability is strongly asymmetric, and differently so per variant.
##  * log_p = FALSE: p in [~2.2e-308, 1-eps] maps to x in [-37.5, +8.2] -- the
##    left tail is deep, the right capped eight sigma out. (In f32 the cap is
##    tighter still, x <= +5.29, because 1-p cannot be resolved below 2^-24; the
##    grid points that collapse are dropped automatically.)
##  * log_p = TRUE: the left tail runs to x ~ -1.9e154, the deepest a double log
##    probability can express, and the right still reaches ~ +37.
ACC_X <- list(
  value = c(grid_symmetric(to = 37, upper_to = 8.2, n_tail = 34),
            outer(c(-1, 1), straddle(SEAMS)), outer(c(-1, 1), NEAR_CENTRE)),
  ## Dense to just past the last seam, then a sparse ladder over the remaining
  ## 145 decades -- one geometric ladder across the whole range would step by
  ## five orders at a time and vault straight over the cuts.
  log   = c(grid_symmetric(to = 2e9, upper_to = 35, n_tail = 60,
                           n_tail_upper = 26),
            -10^seq(9.5, 154, length.out = 16),
            outer(c(-1, 1), straddle(SEAMS)), outer(c(-1, 1), NEAR_CENTRE))
)
## Reduce for a quick render, always keeping both extremes.
thin_grid <- function(x, n = 15L) {
  if (length(x) <= n) return(x)
  idx <- unique(round(seq(1, length(x), length.out = n)))
  sort(unique(c(x[1], x[idx], x[length(x)])))
}
prep_targets <- function(x) {
  x <- sort(unique(as.numeric(x)))
  x <- x[x != 0]              # Phi^-1(1/2) = 0 exactly; a relative error is 0/0
  if (FULL) x else thin_grid(x)
}
```

```{r}
#| label: acc-run

## Evaluate JAX on inputs chosen by R. One subprocess per call; `variants` is a
## list of list(id, log_p, inputs).
JAX_VER <- NA_character_
jax_eval <- function(variants, dtypes, tag) {
  if (!USEJAX) return(NULL)
  sf <- file.path(WORK, paste0("accspec_", tag, ".json"))
  ## digits = 17 round-trips f64 exactly; jsonlite's NA/I() do NOT, and would
  ## turn 1 - 1e-16 into 1, i.e. a probability into a boundary.
  ## as.list() keeps a single dtype a JSON array: auto_unbox would turn "f64"
  ## into a bare string, which python then iterates over character by character.
  write_json(list(variants = variants, dtypes = as.list(dtypes)), sf,
             digits = 17, auto_unbox = TRUE)
  of <- file.path(WORK, paste0("acc_", tag, ".json"))
  run_py_chunk("acc-jax-src", c(sprintf("PJ_SPEC=%s", sf), sprintf("PJ_OUT=%s", of)))
  jx <- fromJSON(of, simplifyVector = TRUE)
  JAX_VER <<- jx$meta$jax
  jx$data
}
## The python side tags non-finite results as "inf" / "-inf" / "nan", so a mixed
## array arrives as character; map it back rather than letting as.numeric() turn
## all three into an undifferentiated NA.
as_num <- function(z) {
  if (is.numeric(z)) return(as.numeric(z))
  if (is.list(z)) {
    z[vapply(z, is.null, TRUE)] <- NA_real_
    z <- unlist(z)
  }
  z <- as.character(z)
  out <- suppressWarnings(as.numeric(z))
  out[z == "inf"] <- Inf
  out[z == "-inf"] <- -Inf
  out[z == "nan"] <- NaN
  out
}

## The whole accuracy pipeline for one set of target grids, run for each dtype:
## targets -> exact input -> narrow to dtype -> reference at the narrowed input
## -> all three implementations on that same input.
run_accuracy <- function(targets, dtypes, grid_label, tag) {
  spec <- list(); store <- list()
  for (v in VARIANTS) {
    x_all <- prep_targets(targets[[v$id]])
    input_f64 <- gold_inputs(x_all, v$log_p)
    ## JAX receives the f64 inputs and narrows them itself, exactly as anvl does,
    ## so both stacks see bit-identical values in each dtype.
    spec[[length(spec) + 1L]] <- list(id = v$id, log_p = v$log_p,
                                      inputs = input_f64)
    store[[v$id]] <- list(x_all = x_all, input_f64 = input_f64)
  }
  jx <- jax_eval(spec, dtypes, tag)

  rows <- list()
  for (v in VARIANTS) {
    x_all <- store[[v$id]]$x_all
    for (dt in dtypes) {
      input <- round_to_dtype(store[[v$id]]$input_f64, dt)
      g <- gold_eval(input, v$log_p)
      ## Drop points where the reference itself is unusable (an input that has
      ## rounded onto a boundary) rather than reporting a meaningless error, and
      ## points whose input has collapsed onto one already covered -- in a narrow
      ## dtype whole stretches of the grid land on the same value.
      keep <- is.finite(g$value) & g$value != 0 & is.finite(g$grad) &
        !duplicated(input)
      x <- x_all[keep]; inp <- input[keep]
      gv <- g$value[keep]; gg <- g$grad[keep]
      ## The reference's own solver must have converged, in relative terms.
      stopifnot(max(g$residual[keep]) < 1e-40)

      arr <- nv_array(inp, dtype = dt)
      av <- anvl_value(arr, v$log_p)
      ag <- as.vector(anvl_grad_fn(v$log_p)(arr)[[1L]])
      add <- function(method, op, computed, gold_m) {
        rows[[length(rows) + 1L]] <<- data.frame(
          grid = grid_label, variant = v$id, variant_label = VAR_LABEL[[v$id]],
          dtype = dt, method = method, op = op, x = x, input = inp,
          computed = computed, rel_err = rel_err(computed, gold_m),
          ulp = ulp_vs_gold(computed, gold_m, dt))
      }
      add(M_ANVL, "value", av, gv)
      add(M_ANVL, "gradient", ag, gg)
      if (identical(dt, "f64")) {           # base R is double precision only
        rv <- r_value(inp, v$log_p)
        add(M_R, "value", rv, gv)
        add(M_R, "gradient", r_grad(inp, v$log_p, rv), gg)
      }
      if (USEJAX) {
        e <- jx[[v$id]][[dt]]
        add(M_JAX, "value", as_num(e$value)[keep], gv)
        add(M_JAX, "gradient", as_num(e$grad)[keep], gg)
      }
    }
  }
  dplyr::bind_rows(rows)
}

acc_df <- run_accuracy(ACC_X, DTYPES, "curated grid", "curated")
```

Relative error against the `r PREC`-bit `Rmpfr` ground truth — **all three
implementations against the same reference, on bit-identical inputs**. The grid
is specified in *target quantiles* $x$ and the input is derived from it as
$\Phi(x)$ (or $\log\Phi(x)$) in `Rmpfr`, which is what gives controlled coverage
of the deep tail; the reference is then evaluated at the input **after narrowing
to the dtype under test**, since that is the value the kernel actually receives.
(Scoring an `f32` run against an `f64`-input reference charges the caller's
rounding to the algorithm and makes every implementation report the same error.)
Grid points whose narrowed input collapses onto one already covered are dropped,
so a narrow dtype shows fewer points rather than repeats.

The $x$-axis is a **signed-log** of the target quantile, so both tails and the
body are visible at once; $y$ is log relative error. The dotted line marks 100%
relative error; triangles in the upper band are results worse than 1000%,
parked there so a single enormous error cannot flatten the rest of the panel
(the true figure is in the tooltip), and $\times$ markers in the top band are
non-finite results. Hover for exact values.

One asymmetry to read the gradient panels with: base R's derivative is the
analytic $p/\phi(x)$ composed as `exp(lp - dnorm(q, log = TRUE))`, which is exact
in principle but subtracts two quantities of size $x^2/2$ in the exponent. Deep
in the log tail those are both $\sim 10^{307}$ and nothing survives the
subtraction — a property of composing the derivative by hand, not of `qnorm`
itself, and precisely the arithmetic that reverse-mode AD does not have to do.

```{r}
#| label: acc-plot-fn

## Two bands above the data. A relative error can be arbitrarily large here --
## base R's composed log-input derivative reaches 1e154 where the exponent
## cancels -- and one such point would stretch a shared log axis over 170 decades
## and flatten everything else into a line. Anything past CLIP_HI is parked on
## its own marker with the true figure in the tooltip; non-finite results get a
## second band above that.
CLIP_HI <- 10                               # 1000% relative error
UF_Y <- 60                                  # marker band for non-finite results
FLOOR_CLIP <- 1e-18
## Axis breaks are chosen per panel, from the panel's own slog limits: the
## log_p = TRUE panel spans 154 decades, where a break set fine enough for the
## other panel piles half a dozen labels on top of each other next to the origin.
inv_slog <- function(y) sign(y) * (10^abs(y) - 1)
acc_breaks <- function(lims) {
  mx <- max(abs(lims), na.rm = TRUE)
  cand <- if (mx > 40) c(0, 1e10, 1e40, 1e80, 1e120, 1e150)
          else c(0, 1, 3, 8, 30, 100, 1e3, 1e6, 1e9)
  b <- slog(sort(unique(c(-cand, cand)))) + 0        # + 0 normalises -0 to 0
  b[b >= lims[1] & b <= lims[2]]
}
acc_labels <- function(b) {
  v <- inv_slog(b) + 0
  ifelse(abs(v) < 1e4, formatC(round(v), format = "d"),
         formatC(v, format = "g", digits = 2))
}

acc_plot <- function(dt, which_op) {
  d <- dplyr::filter(acc_df, dtype == dt, op == which_op)
  d$status <- ifelse(is.na(d$rel_err), "non-finite",
                     ifelse(d$rel_err > CLIP_HI, "off scale", "ok"))
  d$re_plot <- ifelse(d$status == "ok", pmax(d$rel_err, FLOOR_CLIP),
                      ifelse(d$status == "off scale", CLIP_HI, UF_Y))
  d$tt <- sprintf("x = %.6g\n%s: %s\nvalue = %.10g\nulp = %.3g",
                  d$x, d$method,
                  ifelse(d$status == "non-finite", "non-finite",
                         formatC(d$rel_err, format = "e", digits = 2)),
                  d$computed, d$ulp)
  g <- ggplot(d, aes(slog(x), re_plot, colour = method, group = method,
                     text = tt)) +
    geom_hline(yintercept = 1, linetype = "dotted", colour = "grey60") +
    geom_line(data = subset(d, status == "ok"), linewidth = 0.4) +
    geom_point(aes(shape = status), size = 1.2) +
    facet_wrap(~variant_label, scales = "free_x", ncol = 1) +
    ## Label the signed-log axis in the original x, not in slog(x): the log_p
    ## panel spans 154 decades, and an unlabelled slog tick reading "-150" would
    ## be read as x = -150 rather than x = -1e150. Free scales mean each panel
    ## keeps only the breaks inside its own range.
    scale_x_continuous(breaks = acc_breaks, labels = acc_labels) +
    scale_y_log10() +
    scale_shape_manual(values = c(ok = 16, `off scale` = 17, `non-finite` = 4),
                       guide = "none") +
    scale_colour_manual(values = PAL) +
    labs(x = "target quantile x (signed-log scale)", y = "relative error",
         colour = NULL,
         title = sprintf("%s-bit %s — relative error vs %d-bit mpfr",
                         sub("f", "", dt), which_op, PREC)) +
    theme_minimal(base_size = 12)
  ggplotly(g, tooltip = "text") |> legend_below()
}
```

## 64-bit

```{r}
#| label: acc-f64-value
acc_plot("f64", "value")
```

```{r}
#| label: acc-f64-grad
acc_plot("f64", "gradient")
```

## 32-bit

```{r}
#| label: acc-f32-value
acc_plot("f32", "value")
```

```{r}
#| label: acc-f32-grad
acc_plot("f32", "gradient")
```

## Dense random sweep

```{r}
#| label: acc-dense

## A curated grid answers "is every seam and extreme right"; it cannot answer
## "what is the worst case", because it can only ever report the worst of the
## points someone thought to include. This does the complementary job: random
## targets filling the gaps *between* consecutive grid points, f64 only (that is
## where the accuracy claims live, and a second dtype would double the mpfr cost
## for no new question).
##
## Sampling between adjacent pairs rather than log-uniformly across the whole
## reach matters once a variant spans 157 decades: a flat draw would put almost
## every point past |x| = 1e3, where a single asymptotic holds, and none in the
## body where the fitted rationals live. Interpolating the grid inherits the
## density the grid already encodes.
dense_targets <- function(acc_x, n) {
  x <- acc_x[acc_x != 0 & is.finite(acc_x)]
  out <- numeric()
  for (s in c(-1, 1)) {
    side <- sort(unique(abs(x[sign(x) == s])))
    if (length(side) < 2L) next
    k <- max(1L, round(n * length(side) / length(x)))
    i <- sample.int(length(side) - 1L, k, replace = TRUE)
    out <- c(out, s * 10^stats::runif(k, log10(side[i]), log10(side[i + 1L])))
  }
  out
}
set.seed(SEED)
dense_x <- lapply(ACC_X, dense_targets, n = DENSE_N)
dense_df <- run_accuracy(dense_x, "f64", "dense random", "dense")
```

```{r}
#| label: acc-summary
#| echo: false

## Deliberately long rather than pivoted: pivoting orders the value/gradient
## columns alphabetically, which makes the two easy to misread.
##
## `worst rel err` is a maximum over the points that produced a number, so on its
## own it flatters an implementation that declines to produce one. Two companion
## columns stop that: `non-finite` counts Inf/NaN, and `zeroed` counts points
## returned as exactly 0 against a non-zero reference -- a silent wrong answer,
## which in a sampler is worse than a NaN because nothing downstream notices.
summ <- dplyr::bind_rows(acc_df, dense_df) |>
  dplyr::group_by(grid, dtype, variant_label, op, method) |>
  dplyr::summarise(
    n = dplyr::n(),
    `worst rel err` = suppressWarnings(max(rel_err, na.rm = TRUE)),
    `worst ulp` = suppressWarnings(max(ulp, na.rm = TRUE)),
    `worst at x` = x[which.max(replace(rel_err, is.na(rel_err), -1))],
    `non-finite` = sum(!is.finite(computed)),
    zeroed = sum(computed == 0 & is.finite(computed), na.rm = TRUE),
    .groups = "drop") |>
  dplyr::mutate(dplyr::across(c(`worst rel err`, `worst ulp`),
                              ~ ifelse(is.finite(.x), .x, NA_real_))) |>
  dplyr::arrange(`non-finite` > 0 | zeroed > 0, dtype, variant_label, op,
                 `worst rel err`)
datatable(summ, rownames = FALSE, filter = "top",
          options = list(pageLength = 12, scrollX = TRUE)) |>
  formatSignif(c("worst rel err", "worst ulp", "worst at x"), 4)
```

`r format(DENSE_N, big.mark = ",")` random targets per variant, drawn between
adjacent curated-grid points, are pooled with the curated grid in the table above
(`grid` distinguishes them). Rows with any non-finite or silently zeroed result
sort last.

# Behaviour at the seams

```{r}
#| label: seams
#| echo: false

## A relative-error panel samples points and never looks between two of them.
## Offsets are log-spaced from one ulp out to 1e-6 relative, both signs: a
## uniform cluster only straddles the boundary if the declared seam is accurate
## to ~1e-13 relative, whereas the jump metric -- max|d(computed) - d(reference)|,
## a change in *error* between neighbours -- is insensitive to spacing, so a step
## discontinuity registers in full however wide the cluster is.
N_LOCAL <- 33L

## Build every cluster first, then evaluate JAX once per dtype: a subprocess per
## cluster would be sixteen JAX start-ups for thirty-three points each.
clusters <- list()
for (v in VARIANTS) for (dt in DTYPES) {
  eps <- DTYPE_EPS[[dt]]
  d <- 10^seq(log10(2 * eps), -6, length.out = (N_LOCAL - 1L) %/% 2L)
  rel <- sort(c(-d, 0, d))
  for (s in SEAMS) for (sgn in c(-1, 1)) {
    xs <- sort(sgn * s * (1 + rel))
    input <- round_to_dtype(gold_inputs(xs, v$log_p), dt)
    if (anyNA(input) || length(unique(input)) < length(xs)) next  # dtype too coarse
    g <- gold_eval(input, v$log_p)
    gv <- as.numeric(g$value)
    if (!all(is.finite(gv))) next
    clusters[[length(clusters) + 1L]] <- list(
      key = sprintf("c%d", length(clusters) + 1L), v = v, dt = dt, s = sgn * s,
      input = input, gold = gv, ulp_here = 2^floor(log2(abs(s))) * eps)
  }
}
jax_seam <- list()
if (USEJAX) for (dt in DTYPES) {
  sel <- Filter(function(cl) cl$dt == dt, clusters)
  if (!length(sel)) next
  jj <- jax_eval(lapply(sel, function(cl)
                   list(id = cl$key, log_p = cl$v$log_p, inputs = cl$input)),
                 dt, paste0("seam_", dt))
  for (cl in sel) jax_seam[[cl$key]] <- as_num(jj[[cl$key]][[dt]]$value)
}

seam_rows <- list()
for (cl in clusters) {
  dg <- diff(cl$gold)
  ## Only judge monotonicity where the reference itself moves by several ulp: the
  ## innermost offsets are ~1 ulp apart, and at that spacing a step that does not
  ## move, or moves the other way, is ordinary rounding.
  resolvable <- abs(dg) > 4 * cl$ulp_here
  vals <- list()
  vals[[M_ANVL]] <- anvl_value(nv_array(cl$input, dtype = cl$dt), cl$v$log_p)
  if (identical(cl$dt, "f64")) vals[[M_R]] <- r_value(cl$input, cl$v$log_p)
  if (USEJAX) vals[[M_JAX]] <- jax_seam[[cl$key]]
  for (nm in names(vals)) {
    dv <- diff(vals[[nm]])
    seam_rows[[length(seam_rows) + 1L]] <- data.frame(
      variant = VAR_LABEL[[cl$v$id]], dtype = cl$dt, seam = cl$s, method = nm,
      `excess jump (ulp)` = if (all(is.finite(dv))) max(abs(dv - dg)) / cl$ulp_here
                            else NA_real_,
      reversals = sum(resolvable & sign(dv) == -sign(dg), na.rm = TRUE),
      `flat steps` = sum(resolvable & dv == 0, na.rm = TRUE),
      `non-finite` = sum(!is.finite(vals[[nm]])),
      check.names = FALSE)
  }
}
seam_df <- dplyr::bind_rows(seam_rows)
if (!nrow(seam_df)) {
  htmltools::HTML("<em>No seam is resolvable in these dtypes.</em>")
} else {
  seam_df |>
    dplyr::arrange(dplyr::desc(`excess jump (ulp)`)) |>
    datatable(rownames = FALSE, filter = "top",
              options = list(pageLength = 12, scrollX = TRUE)) |>
    formatSignif(c("excess jump (ulp)", "seam"), 4)
}
```

Both anvl and JAX switch between a central rational and a tail rational at
$t = e^{-2}$ ($|x| = `r signif(SEAMS[1], 6)`$), and between two tail fits at
$z = 8$ ($|x| = `r signif(SEAMS[2], 6)`$). Two properties live *between* grid
points, and neither shows up in a relative-error panel.

**Excess jump** is how far the step across the cut exceeds the reference's own
step, in ulps, measured on `r N_LOCAL` points spanning $\pm 10^{-6}$ relative
either side of each seam (and its mirror image in the upper tail). A few ulp is
rounding; a large value is a genuine discontinuity — invisible above, and a cliff
to a sampler, an optimiser, or anything that assumes a smooth map.

**Reversals** and **flat steps** are counted separately, wherever the reference's
own step exceeds 4 ulp. A reversal is genuine non-monotonicity, which breaks
inverse-transform sampling and bisection outright. A flat step means the
implementation's parametrisation discarded resolution the input still carried —
different diagnosis, different fix, so they are not pooled.

# Boundary and invalid inputs

```{r}
#| label: domain
#| echo: false

## The accuracy grid is built from valid targets and then filtered to points
## where the reference is finite and non-zero, so p = 0, p = 1, p outside [0,1]
## and NaN can never appear in it -- even though all three implementations carry
## explicit code for them, and even though these are exactly the inputs a sampler
## hands over when a proposal goes bad. p = 1/2 is in the same position: x = 0 is
## dropped because a relative error there is 0/0, so the one input at which the
## lower/upper reflection changes hands is otherwise never evaluated.
domain_cases <- function(log_p) {
  if (log_p) {
    data.frame(
      label = c("log p = -Inf (p = 0)", "log p = 0 (p = 1)", "log p = log(1/2)",
                "log p > 0 (invalid)", "NaN", "log p = -2 exactly (seam)",
                "log p = -32 exactly (z = 8 seam)"),
      input = c(-Inf, 0, base::log(0.5), 0.5, NaN, -2, -32),
      expected = c(-Inf, Inf, 0, NaN, NaN, NA_real_, NA_real_))
  } else {
    data.frame(
      label = c("p = 0", "p = 1", "p = 1/2", "p < 0 (invalid)", "p > 1 (invalid)",
                "NaN", "p = exp(-2) (seam)", "p = 1 - exp(-2) (reflected seam)"),
      input = c(0, 1, 0.5, -0.25, 1.25, NaN, exp(-2), 1 - exp(-2)),
      expected = c(-Inf, Inf, 0, NaN, NaN, NaN, NA_real_, NA_real_))
  }
}

## NaN must match NaN and +-Inf must match the same infinity -- a sign flip or a
## silent Inf where NaN is correct is the whole point of the panel. Finite values
## are compared with a tolerance in ulps *of the dtype*, since the two seam rows
## are judged against the mpfr reference and an f32 answer cannot be nearer than
## its own rounding.
same <- function(a, b, dtype) {
  tol <- 8 * DTYPE_EPS[[dtype]]
  ifelse(is.nan(a) | is.nan(b), is.nan(a) & is.nan(b),
         ifelse(is.finite(a) != is.finite(b), FALSE,
                ifelse(is.finite(a), abs(a - b) <= tol * pmax(abs(b), 1),
                       a == b)))
}
## Display helper: Inf/NaN survive into the table as words rather than blanks.
fmt_val <- function(z) {
  ifelse(is.nan(z), "NaN",
         ifelse(is.infinite(z), ifelse(z > 0, "Inf", "-Inf"),
                formatC(z, digits = 10, format = "g")))
}

## One JAX subprocess per dtype, both variants inside it.
jax_dom <- list()
if (USEJAX) for (dt in DTYPES) {
  jax_dom[[dt]] <- jax_eval(
    lapply(VARIANTS, function(v)
      list(id = v$id, log_p = v$log_p,
           inputs = round_to_dtype(domain_cases(v$log_p)$input, dt))),
    dt, paste0("dom_", dt))
}

dom_rows <- list()
for (v in VARIANTS) {
  dc <- domain_cases(v$log_p)
  for (dt in DTYPES) {
    input <- round_to_dtype(dc$input, dt)
    g <- gold_eval(input, v$log_p)
    ## `expected = NA_real_` means "judge against the mpfr reference", which is
    ## how the two exact-seam cases are checked: a clamp that ties with the
    ## constant the code branches on leaves the value correct while splitting the
    ## derivative, so the seam inputs are worth evaluating even though the value
    ## is ordinary. NaN is a legitimate *expectation* elsewhere in the column, so
    ## test for it explicitly rather than with is.na(), which is true for both.
    want <- ifelse(is.na(dc$expected) & !is.nan(dc$expected),
                   as.numeric(g$value), dc$expected)
    got <- list()
    got[[M_ANVL]] <- anvl_value(nv_array(input, dtype = dt), v$log_p)
    if (identical(dt, "f64")) got[[M_R]] <- suppressWarnings(r_value(input, v$log_p))
    if (USEJAX) got[[M_JAX]] <- as_num(jax_dom[[dt]][[v$id]][[dt]]$value)
    for (nm in names(got)) {
      dom_rows[[length(dom_rows) + 1L]] <- data.frame(
        variant = VAR_LABEL[[v$id]], dtype = dt, case = dc$label, method = nm,
        expected = fmt_val(want), got = fmt_val(got[[nm]]),
        verdict = ifelse(same(got[[nm]], want, dt), "ok", "MISMATCH"))
    }
  }
}
dplyr::bind_rows(dom_rows) |>
  dplyr::arrange(verdict != "MISMATCH", variant, case, method) |>
  datatable(rownames = FALSE, filter = "top",
            options = list(pageLength = 12, scrollX = TRUE))
```

Where `expected` is a plain limit ($\pm\infty$, `0`, `NaN`) it is the documented
answer; on the two seam rows it is the `Rmpfr` reference, since the interesting
question there is not the limit but whether landing exactly on a branch cut
changes the answer. `NaN` where a probability is out of range is the correct
outcome — silent `±Inf` there would be the worst result of all, because nothing
downstream notices.

# Detailed results

Full numeric results, collapsed by default.

::: {.callout-note title="Performance — every cell (time per call, ms)" collapse="true"}
```{r}
#| label: perf-table
if (!PERF) {
  htmltools::HTML("<em>Performance report disabled (<code>QNORM_PERF=0</code>).</em>")
} else {
  dplyr::bind_rows(perf_df, regime_df) |>
    dplyr::transmute(dtype, size, regime = as.character(regime),
                     variant = variant_label, op, method,
                     `time (ms)` = time_ms, `spread (max/min)` = spread) |>
    dplyr::arrange(dtype, variant, op, regime, size, method) |>
    datatable(rownames = FALSE, filter = "top",
              options = list(pageLength = 15, scrollX = TRUE)) |>
    formatSignif(c("time (ms)", "spread (max/min)"), 3)
}
```
:::

::: {.callout-note title="Accuracy — every grid point (relative error, ulps & value)" collapse="true"}
```{r}
#| label: acc-table
## Rounding is display-only (formatSignif); the underlying data stays exact.
dplyr::bind_rows(acc_df, dense_df) |>
  dplyr::transmute(grid, dtype, variant = variant_label, op, method, x, input,
                   `rel err` = rel_err, ulp, computed) |>
  dplyr::arrange(grid, dtype, variant, op, x, method) |>
  datatable(rownames = FALSE, filter = "top",
            options = list(pageLength = 15, scrollX = TRUE)) |>
  formatSignif(c("rel err", "ulp"), 3) |>
  formatSignif(c("x", "input", "computed"), 8)
```
:::

---

```{r}
#| label: colophon
#| echo: false
#| results: asis

## Verify -- rather than assert in prose -- that f64 literals inside a jitted
## graph keep full precision. Without it every f64 relative error below would sit
## on a uniform ~1.7e-8 floor whatever the algorithm, which is exactly the shape
## of error a reader would blame on the implementation. Also confirm f32 graphs
## are not silently widened.
lit_note <- local({
  f <- function(x) x * pi
  exact <- tryCatch(identical(as.vector(jit(f)(nv_array(1, dtype = "f64"))), pi),
                    error = function(e) NA)
  narrow <- tryCatch(
    identical(unname(format(dtype(jit(f)(nv_array(1, dtype = "f32"))))), "32"),
    error = function(e) NA)
  if (isTRUE(exact) && isTRUE(narrow))
    "f64 literal precision **verified** (`jit(\\(x) x * pi)` is bit-exact in f64, and an f32 graph stays f32)."
  else if (isTRUE(narrow))
    "**WARNING: f64 literals are rounding to f32 inside jitted graphs** (anvl's `default_dtype()` materialises an R double at `FloatType(32)`, so a coefficient loses its low 29 bits before promotion ever sees it). Every f64 relative error above is then floored near 1.7e-8 regardless of algorithm, and the accuracy panels cannot be read in f64 as rendered."
  else
    "**WARNING: the f64 literal self-check did not run.** Treat every f64 accuracy number above as unverified."
})

cpu_model <- function() {
  out <- tryCatch({
    if (identical(Sys.info()[["sysname"]], "Darwin")) {
      system2("sysctl", c("-n", "machdep.cpu.brand_string"), stdout = TRUE)
    } else {
      sub("^model name\\s*:\\s*", "",
          grep("^model name", readLines("/proc/cpuinfo"), value = TRUE)[1L])
    }
  }, error = function(e) NULL)
  if (length(out) && nzchar(out[1L])) trimws(out[1L]) else "unknown CPU"
}

## Points kept per (variant, dtype), not pooled: a narrow dtype legitimately
## keeps fewer, and a single total would hide that.
n_acc <- acc_df |>
  dplyr::distinct(variant_label, dtype, x) |>
  dplyr::count(variant_label, dtype) |>
  dplyr::mutate(txt = sprintf("%s/%s: %d", variant_label, dtype, n))
cat(sprintf(
  "Generated %s · %s · device %s · anvl %s · JAX %s · %s · Rmpfr %s (%d-bit) · bench %s · google-benchmark.\n\n%s %s Curated accuracy points kept (%s); %s random targets per variant in the dense sweep.\n",
  format(Sys.Date()), cpu_model(), toupper(DEVICE),
  as.character(packageVersion("anvl")),
  if (is.na(JAX_VER)) "not run" else JAX_VER,
  R.version.string, as.character(packageVersion("Rmpfr")), PREC,
  as.character(packageVersion("bench")), lit_note,
  if (PERF) sprintf("Performance sizes %s.",
                    paste(format(SIZES, scientific = FALSE, trim = TRUE),
                          collapse = ", ")) else "Performance report disabled.",
  paste(n_acc$txt, collapse = ", "), format(DENSE_N, big.mark = ",")))
```
