---
title: "`nv_pnorm` vs JAX — performance & accuracy"
subtitle: "Standard normal CDF: value & gradient, `log.p` on/off, 32- & 64-bit"
date: today
author: Opus 4.8 supervised by Louis Aslett
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-tools: true
    code-summary: "Show code"
    embed-resources: true
    theme: cosmo
    fig-width: 9
    fig-height: 6.5
    tbl-cap-location: top
execute:
  warning: false
  message: false
  cache: false
---

```{r}
#| label: setup
#| include: false

## Everything regenerates on `quarto render`; no result is hard-coded.
##
## Portable configuration -- no machine-specific path is baked in. Override on
## the render line, e.g.
##   PNORM_PY=/path/to/venv/bin/python quarto render pnorm-jax-comparison.qmd
##
##   PNORM_PY    Python interpreter that provides jax, numpy and
##               google-benchmark. Default: this repo's uv venv at
##               ../../../py-benchmarks/.venv/bin/python, else python3 on PATH.
##   PNORM_FULL  "0" for a fast reduced-grid preview; default runs the full grid.
##   PNORM_PERF  "0" skips the performance report entirely (it renders blank);
##               accuracy still runs. Independent of PNORM_FULL -- use it when
##               iterating on the accuracy panels without waiting on timings.
##   DEVICE      Compute device for BOTH anvl and JAX: "cpu" (default) or "cuda"
##               for GPU. GPU needs the matching PJRT plugin (anvl) and a
##               CUDA-enabled jaxlib (JAX) installed on the target machine.
env_or <- function(var, default = "") {
  v <- Sys.getenv(var, unset = ""); if (nzchar(v)) v else default
}
FULL   <- env_or("PNORM_FULL", "1") != "0"
PERF   <- env_or("PNORM_PERF", "1") != "0"
DEVICE <- env_or("DEVICE", "cpu")
## anvl reads PJRT_PLATFORM in default_device(); nv_array()/jit() then place and
## compile on it. JAX is steered per-subprocess via JAX_PLATFORMS (run_py_chunk).
Sys.setenv(PJRT_PLATFORM = DEVICE)

suppressPackageStartupMessages({
  library(anvl)
  library(Rmpfr)
  library(jsonlite)
  library(bench)
  library(ggplot2)
  library(plotly)
  library(dplyr)
  library(tidyr)
  library(DT)
  library(htmltools)
})

## Python interpreter (needs jax, numpy, google-benchmark); env-overridable.
PY_BIN <- env_or("PNORM_PY")
if (!nzchar(PY_BIN)) {
  repo_py <- "../../../py-benchmarks/.venv/bin/python"
  PY_BIN <- if (file.exists(repo_py)) repo_py else unname(Sys.which("python3"))
}
if (!nzchar(PY_BIN) || !file.exists(PY_BIN))
  stop("No Python interpreter found. Set PNORM_PY=/path/to/venv/bin/python ",
       "(needs jax, numpy, google-benchmark).")
if (!identical(0L, suppressWarnings(system2(PY_BIN,
      c("-c", shQuote("import jax, numpy, google_benchmark")),
      stdout = FALSE, stderr = FALSE))))
  stop("Python at '", PY_BIN, "' lacks jax / numpy / google-benchmark; ",
       "point PNORM_PY at an environment that has all three.")

WORK <- tempfile("pnorm_jax_"); dir.create(WORK)

## --- run an embedded (eval:false) python chunk in the project venv --------
run_py_chunk <- function(label, envvars = character()) {
  src <- knitr::knit_code$get(label)
  py  <- file.path(WORK, paste0(label, ".py"))
  writeLines(src, py)
  status <- system2(PY_BIN, shQuote(py),
                    env = c(envvars, paste0("JAX_PLATFORMS=", DEVICE)),
                    stdout = TRUE, stderr = TRUE)
  st <- attr(status, "status")
  if (!is.null(st) && st != 0)
    stop("python chunk '", label, "' failed:\n", paste(status, collapse = "\n"))
  invisible(status)
}

## --- shared configuration -------------------------------------------------
DTYPES  <- c("f64", "f32")
SIZES   <- if (FULL) c(1, 10, 100, 1e3, 1e4, 1e5, 1e6, 1e7) else c(1, 100, 1e4)
# Input-generation recipe shared verbatim with the python side (perf-jax-src).
# For timing only the *distribution* of inputs must match, not the exact draws,
# so each language samples locally from the same band with a fixed seed.
REGIONS <- list(
  deep_left  = c(-40, -25),   # below the asymptotic (Mills-ratio) threshold
  middle     = c(-3,   3),    # bulk of the distribution
  deep_upper = c( 25,  40),   # far upper tail (log1p / near-1 regime)
  random     = c(-40,  40)    # scattered across every branch
)
SEED <- 1L
PREC <- 512L                  # bits of mpfr ground-truth precision
```

```{python}
#| label: perf-jax-src
#| eval: false

# JAX performance for one (dtype, size): times cdf/logcdf value & gradient
# across the 4 input regions with google-benchmark. Parameters arrive via
# environment variables; a google-benchmark JSON is written for R to read.
import os
import jax
jax.config.update("jax_enable_x64", True)
import jax.numpy as jnp
import numpy as np
import google_benchmark as gb

DT = {"f64": jnp.float64, "f32": jnp.float32}
REGIONS = {"deep_left": (-40, -25), "middle": (-3, 3),
           "deep_upper": (25, 40), "random": (-40, 40)}

value_fn = {False: jax.jit(lambda x: jax.scipy.stats.norm.cdf(x)),
            True:  jax.jit(lambda x: jax.scipy.stats.norm.logcdf(x))}
grad_fn = {False: jax.jit(jax.vmap(jax.grad(lambda z: jax.scipy.stats.norm.cdf(z)))),
           True:  jax.jit(jax.vmap(jax.grad(lambda z: jax.scipy.stats.norm.logcdf(z))))}

dt = DT[os.environ["PJ_DTYPE"]]
size = int(os.environ["PJ_SIZE"])
rng = np.random.default_rng(int(os.environ["PJ_SEED"]))
arrays = {r: jnp.asarray(rng.uniform(lo, hi, size), dtype=dt)
          for r, (lo, hi) in REGIONS.items()}


def make_bench(fn, arr):
    def b(state):
        while state:
            np.asarray(fn(arr))          # force host materialisation, cf. as.vector()
    return b


for region, arr in arrays.items():
    for logp in (False, True):
        for op, table in (("value", value_fn), ("grad", grad_fn)):
            fn = table[logp]
            np.asarray(fn(arr))          # warm up / compile
            gb.register(make_bench(fn, arr),
                        name=f"{region}|{'T' if logp else 'F'}|{op}")

gb.main([
    "perf_jax", "--benchmark_format=json",
    f"--benchmark_out={os.environ['PJ_OUT']}", "--benchmark_out_format=json",
    f"--benchmark_min_time={os.environ['PJ_MINTIME']}s",
    f"--benchmark_repetitions={os.environ['PJ_REPS']}",
])
```

```{python}
#| label: acc-jax-src
#| eval: false

# JAX values on a shared z-grid: cdf, logcdf, and their elementwise gradients,
# for both dtypes. Reads the grid JSON from PJ_GRID, writes results to PJ_OUT.
# Results are cast to float64 for lossless transport (f32 is a subset of f64).
import os, json
import jax
jax.config.update("jax_enable_x64", True)
import jax.numpy as jnp
import numpy as np

DT = {"f64": jnp.float64, "f32": jnp.float32}
cdf = jax.jit(lambda x: jax.scipy.stats.norm.cdf(x))
logcdf = jax.jit(lambda x: jax.scipy.stats.norm.logcdf(x))
gcdf = jax.jit(jax.vmap(jax.grad(lambda z: jax.scipy.stats.norm.cdf(z))))
glogcdf = jax.jit(jax.vmap(jax.grad(lambda z: jax.scipy.stats.norm.logcdf(z))))

grid = json.load(open(os.environ["PJ_GRID"]))
out = {"meta": {"jax": jax.__version__}}
for name, dt in DT.items():
    z = jnp.asarray(grid[name], dtype=dt)
    out[name] = {
        "value_F": np.asarray(cdf(z), np.float64).tolist(),
        "value_T": np.asarray(logcdf(z), np.float64).tolist(),
        "grad_F":  np.asarray(gcdf(z), np.float64).tolist(),
        "grad_T":  np.asarray(glogcdf(z), np.float64).tolist(),
    }
json.dump(out, open(os.environ["PJ_OUT"], "w"))
```

This document compares the proposed anvl `nv_pnorm()` against
JAX's `jax.scipy.stats.norm` on `r toupper(DEVICE)`. Every combination of **dtype** (32-/64-bit),
**`log.p`** (the value $\Phi$ vs the log-CDF $\log\Phi$), and **quantity** (the value
vs its gradient w.r.t. the input) is measured, for the standard normal
(`mean = 0`, `sd = 1`, `lower.tail = TRUE`). Performance uses the R
[`bench`](https://bench.r-lib.org) package and Python
[`google-benchmark`](https://github.com/google/benchmark); accuracy uses
`Rmpfr` at `r PREC`-bit precision as ground truth.

# Performance

```{r}
#| label: perf-run

anvl_perf_cell <- function(arr, logp, op) {
  if (op == "value") {
    call_fn <- function() as.vector(nv_pnorm(arr, log.p = logp))
  } else {
    gfn <- jit(gradient(function(q) nv_reduce_sum(nv_pnorm(q, log.p = logp)),
                        wrt = "q"))
    call_fn <- function() as.vector(gfn(arr)[[1L]])
  }
  invisible(call_fn())                         # warm up / compile
  b <- bench::mark(call_fn(), min_iterations = 5, min_time = 0.05,
                   max_iterations = 200, check = FALSE, memory = FALSE,
                   filter_gc = FALSE)
  as.numeric(b$min) * 1000                      # robust minimum, ms
}

perf_df <- NULL
if (PERF) {                    # skip the whole performance grid when PNORM_PERF=0
perf_rows <- list()
for (dt in DTYPES) {
  for (size in SIZES) {
    n <- as.integer(size)

    ## ---- anvl (bench) ----
    for (rg in names(REGIONS)) {
      lo <- REGIONS[[rg]][1]; hi <- REGIONS[[rg]][2]
      arr <- nv_array(runif(n, lo, hi), dtype = dt)
      for (logp in c(FALSE, TRUE)) for (op in c("value", "grad")) {
        perf_rows[[length(perf_rows) + 1L]] <- data.frame(
          dtype = dt, size = size, region = rg, logp = logp, op = op,
          impl = "anvl", time_ms = anvl_perf_cell(arr, logp, op))
      }
      rm(arr); gc(FALSE)
    }

    ## ---- JAX (google-benchmark, one subprocess per (dtype, size)) ----
    out_json <- file.path(WORK, sprintf("perf_%s_%d.json", dt, n))
    run_py_chunk("perf-jax-src", c(
      sprintf("PJ_DTYPE=%s", dt), sprintf("PJ_SIZE=%d", n),
      sprintf("PJ_OUT=%s", out_json), sprintf("PJ_SEED=%d", SEED),
      "PJ_MINTIME=0.05", "PJ_REPS=5"))
    jx <- fromJSON(out_json, simplifyVector = FALSE)$benchmarks
    reps <- Filter(function(b) b$run_type == "iteration", jx)
    tmin <- tapply(vapply(reps, function(b) b$real_time, numeric(1)),
                   vapply(reps, function(b) b$run_name, character(1)), min)
    for (nm in names(tmin)) {
      parts <- strsplit(nm, "|", fixed = TRUE)[[1]]
      perf_rows[[length(perf_rows) + 1L]] <- data.frame(
        dtype = dt, size = size, region = parts[1],
        logp = parts[2] == "T", op = parts[3],
        impl = "jax", time_ms = tmin[[nm]] / 1e6)   # ns -> ms
    }
  }
}
perf_df <- dplyr::bind_rows(perf_rows)
perf_df$region <- factor(perf_df$region,
                         levels = c("deep_left", "middle", "deep_upper", "random"))
perf_df$logp_lab <- ifelse(perf_df$logp, "log.p = TRUE", "log.p = FALSE")
}
```

```{r}
#| label: perf-intro
#| echo: false
#| results: asis
if (PERF) cat(sprintf(
  "Steady-state wall-clock time **per call** — a robust minimum over repeated timings (`bench`'s fastest iteration for anvl, google-benchmark's fastest repetition for JAX), with host materialisation forced on both sides — as the input grows from a single element to %s. Four input regimes are shown: `deep_left` (all below the asymptotic-tail threshold), `middle` (the bulk), `deep_upper` (far upper tail), and `random` (scattered across every branch). Hover any point for exact values.\n",
  format(max(SIZES), big.mark = ",", scientific = FALSE)))
if (!PERF) cat("*Performance report disabled (`PNORM_PERF=0`). Set `PNORM_PERF=1` to include it.*\n")
```

```{r}
#| label: perf-plot-fn

perf_plot <- function(dt, op) {
  d <- dplyr::filter(perf_df, dtype == dt, op == !!op)
  d$tt <- sprintf("size %s | %s | %s\n%s: %.3g ms",
                  format(d$size, big.mark = ",", scientific = FALSE),
                  d$region, d$logp_lab, d$impl, d$time_ms)
  g <- ggplot(d, aes(size, time_ms, colour = impl, linetype = logp_lab,
                     group = interaction(impl, logp_lab), text = tt)) +
    geom_line(linewidth = 0.5) + geom_point(size = 1.4) +
    facet_wrap(~region, ncol = 2) +
    scale_x_log10(breaks = SIZES,
                  labels = format(SIZES, scientific = FALSE, big.mark = ",",
                                  trim = TRUE)) +
    scale_y_log10() +
    scale_colour_manual(values = c(anvl = "#2c7fb8", jax = "#d95f0e")) +
    labs(x = "input length", y = "time per call (ms)",
         colour = NULL, linetype = NULL,
         title = sprintf("%s-bit · %s", sub("f", "", dt),
                         if (op == "value") "value" else "gradient")) +
    theme_minimal(base_size = 12)
  ggplotly(g, tooltip = "text") |> layout(legend = list(orientation = "h", y = -0.15))
}
```

```{r}
#| label: perf-h-64
#| echo: false
#| results: asis
if (PERF) cat("\n## 64-bit\n")
```

```{r}
#| label: perf-f64-value
if (PERF) perf_plot("f64", "value")
```

```{r}
#| label: perf-f64-grad
if (PERF) perf_plot("f64", "grad")
```

```{r}
#| label: perf-h-32
#| echo: false
#| results: asis
if (PERF) cat("\n## 32-bit\n")
```

```{r}
#| label: perf-f32-value
if (PERF) perf_plot("f32", "value")
```

```{r}
#| label: perf-f32-grad
if (PERF) perf_plot("f32", "grad")
```

# Accuracy

```{r}
#| label: acc-config

## z-grid: dense through the body, geometric into both tails. Extents are
## chosen wide enough to bracket, on both sides, the region where either
## implementation still holds accuracy (the deep-tail floor and the tail
## under/overflow walls); per-panel trims below focus each view.
mags <- sort(unique(c(seq(0, 6, by = 0.2), seq(6, 45, by = 0.5),
                      10^seq(log10(6), 4, length.out = 60))))
zgrid <- sort(unique(c(-mags, mags)))

## Per-(dtype, op, log.p) display window (signed z). Deep-tail (log.p = TRUE)
## windows run to the mpfr-reliable limit on the left; value/gradient windows
## bracket the tail under/overflow walls.
trims <- list(
  f64 = list(value_F = c(-44, 9), grad_F = c(-44, 44),
             value_T = c(-1e4, 44), grad_T = c(-1e4, 44)),
  f32 = list(value_F = c(-16, 9), grad_F = c(-16, 16),
             value_T = c(-1e4, 16), grad_T = c(-1e4, 16)))
COMBOS <- c("value_F", "value_T", "grad_F", "grad_T")
```

```{r}
#| label: acc-run

## --- 512-bit mpfr ground truth (dtype-independent, computed once) ---------
s2 <- sqrt(mpfr(2, PREC)); pi_m <- Const("pi", PREC)
gold_all <- function(z) {
  zm  <- mpfr(z, PREC)
  Phi <- 0.5 * erfc(-zm / s2)
  phi <- exp(-(zm * zm) / 2) / sqrt(2 * pi_m)
  logPhi <- mpfr(numeric(length(z)), PREC)          # cancellation-free logPhi
  lo <- z <= 0
  if (any(lo))   logPhi[lo]  <- log(0.5 * erfc(-zm[lo] / s2))
  if (!all(lo))  logPhi[!lo] <- log1p(-0.5 * erfc(zm[!lo] / s2))
  list(value_F = Phi, value_T = logPhi, grad_F = phi, grad_T = phi / Phi)
}
## Cast computed doubles up to mpfr, form the relative error at full precision,
## and only then cast the single scalar back to a double.
rel_err <- function(computed, gold_m) {
  re <- as.numeric(abs(mpfr(computed, PREC) - gold_m) / abs(gold_m))
  re[!is.finite(computed)] <- NA_real_
  re
}
gold <- gold_all(zgrid)

## --- anvl on the grid -----------------------------------------------------
anvl_acc <- function(z, dt) {
  arr <- nv_array(z, dtype = dt)
  gF <- jit(gradient(function(q) nv_reduce_sum(nv_pnorm(q, log.p = FALSE)), wrt = "q"))
  gT <- jit(gradient(function(q) nv_reduce_sum(nv_pnorm(q, log.p = TRUE)),  wrt = "q"))
  list(value_F = as.vector(nv_pnorm(arr, log.p = FALSE)),
       value_T = as.vector(nv_pnorm(arr, log.p = TRUE)),
       grad_F  = as.vector(gF(arr)[[1L]]),
       grad_T  = as.vector(gT(arr)[[1L]]))
}

## --- JAX on the same grid (one subprocess) --------------------------------
grid_json <- file.path(WORK, "acc_grid.json")
write_json(list(f64 = zgrid, f32 = zgrid), grid_json, digits = 17, auto_unbox = TRUE)
acc_out <- file.path(WORK, "acc_jax.json")
run_py_chunk("acc-jax-src", c(sprintf("PJ_GRID=%s", grid_json),
                              sprintf("PJ_OUT=%s", acc_out)))
jax_acc <- fromJSON(acc_out, simplifyVector = TRUE)
JAX_VER <- jax_acc$meta$jax

op_of   <- c(value_F = "value", value_T = "value", grad_F = "grad", grad_T = "grad")
logp_of <- c(value_F = FALSE,   value_T = TRUE,     grad_F = FALSE,  grad_T = TRUE)

acc_rows <- list()
for (dt in DTYPES) {
  av <- anvl_acc(zgrid, dt)
  for (cb in COMBOS) {
    ra <- rel_err(av[[cb]], gold[[cb]])
    rj <- rel_err(jax_acc[[dt]][[cb]], gold[[cb]])
    acc_rows[[length(acc_rows) + 1L]] <- data.frame(
      dtype = dt, combo = cb, op = op_of[[cb]], logp = logp_of[[cb]], z = zgrid,
      anvl_val = av[[cb]], jax_val = jax_acc[[dt]][[cb]],
      anvl_re = ra, jax_re = rj)
  }
}
acc_df <- dplyr::bind_rows(acc_rows)
```

Relative error against the `r PREC`-bit `Rmpfr` ground truth (for gradients, the
analytic derivative $\phi$ or $\phi/\Phi$, also computed in `Rmpfr`). The
$x$-axis is a **signed-log** of the input $z$ so both tails and the body are
visible at once; $y$ is log relative error. The dotted line marks 100% relative
error — a finite result that has lost all accuracy (e.g. tail underflow to
zero); $\times$ markers in the top band are non-finite results. Each panel's
$x$-range spans the region where at least one implementation still holds
accuracy (bounded, on the deep-tail side, by how far the mpfr reference itself
stays exact). Hover for exact values.

```{r}
#| label: acc-plot-fn

slog <- function(z) sign(z) * log10(1 + abs(z))
UF_Y <- 5                                   # marker band for non-finite results
FLOOR <- 1e-17
z_breaks <- c(-1e4, -1e3, -100, -30, -10, -3, -1, 0, 1, 3, 10, 30, 100, 1e3, 1e4)

acc_plot <- function(dt) {
  long <- acc_df |>
    dplyr::filter(dtype == dt) |>
    tidyr::pivot_longer(c(anvl_re, jax_re), names_to = "impl", values_to = "re") |>
    dplyr::mutate(impl = sub("_re", "", impl),
                  value = ifelse(impl == "anvl", anvl_val, jax_val))
  ## per-panel trim to its display window
  keep <- mapply(function(dd, cc, zz) {
    tr <- trims[[dd]][[cc]]; zz >= tr[1] & zz <= tr[2]
  }, long$dtype, long$combo, long$z)
  long <- long[keep, ]

  long$status <- ifelse(is.na(long$re), "non-finite", "ok")
  long$re_plot <- ifelse(long$status == "ok", pmax(long$re, FLOOR), UF_Y)
  long$panel <- factor(paste0(long$op, ",  log.p = ", long$logp),
                       levels = c("value,  log.p = FALSE", "value,  log.p = TRUE",
                                  "grad,  log.p = FALSE",  "grad,  log.p = TRUE"))
  long$tt <- sprintf("z = %.4g\n%s: %s\nvalue = %.6g",
                     long$z, long$impl,
                     ifelse(long$status == "ok",
                            formatC(long$re, format = "e", digits = 2), "non-finite"),
                     long$value)

  g <- ggplot(long, aes(slog(z), re_plot, colour = impl, group = impl, text = tt)) +
    geom_hline(yintercept = 1, linetype = "dotted", colour = "grey60") +
    geom_line(data = subset(long, status == "ok"), linewidth = 0.4) +
    geom_point(aes(shape = status), size = 1.3) +
    facet_wrap(~panel, scales = "free_x", ncol = 2) +
    scale_x_continuous(breaks = slog(z_breaks),
                       labels = format(z_breaks, scientific = FALSE, trim = TRUE)) +
    scale_y_log10() +
    scale_shape_manual(values = c(ok = 16, `non-finite` = 4), guide = "none") +
    scale_colour_manual(values = c(anvl = "#2c7fb8", jax = "#d95f0e")) +
    labs(x = "input z (signed-log scale)", y = "relative error", colour = NULL,
         title = sprintf("%s-bit relative error vs %d-bit mpfr",
                         sub("f", "", dt), PREC)) +
    theme_minimal(base_size = 12)
  ggplotly(g, tooltip = "text") |> layout(legend = list(orientation = "h", y = -0.18))
}
```

## 64-bit

```{r}
#| label: acc-f64
acc_plot("f64")
```

## 32-bit

```{r}
#| label: acc-f32
acc_plot("f32")
```

# Detailed results

Full numeric results, three significant figures. Collapsed by default.

::: {.callout-note title="Performance — every cell (time per call, ms)" collapse="true"}
```{r}
#| label: perf-table
if (!PERF) {
  htmltools::HTML("<em>Performance report disabled (<code>PNORM_PERF=0</code>).</em>")
} else {
  perf_tab <- perf_df |>
    dplyr::transmute(dtype, size, region = as.character(region),
                     `log.p` = logp, op, impl, `time (ms)` = time_ms) |>
    dplyr::arrange(dtype, op, region, size, impl)
  ## Rounding is display-only (formatSignif); the underlying data stays exact.
  datatable(perf_tab, rownames = FALSE, filter = "top",
            options = list(pageLength = 15, scrollX = TRUE)) |>
    formatSignif("time (ms)", 3)
}
```
:::

::: {.callout-note title="Accuracy — every grid point (relative error & value)" collapse="true"}
```{r}
#| label: acc-table
## Relative error is rounded to 3 s.f. for readability (display-only); the
## value columns keep full double precision, as do all underlying data.
acc_tab <- acc_df |>
  dplyr::transmute(dtype, op, `log.p` = logp, z,
                   `anvl rel.err` = anvl_re, `jax rel.err` = jax_re,
                   `anvl value`   = anvl_val, `jax value`  = jax_val) |>
  dplyr::arrange(dtype, op, `log.p`, z)
datatable(acc_tab, rownames = FALSE, filter = "top",
          options = list(pageLength = 15, scrollX = TRUE)) |>
  formatSignif(c("anvl rel.err", "jax rel.err"), 3) |>
  formatSignif("z", 6)
```
:::

---

```{r}
#| label: colophon
#| echo: false
#| results: asis
grid_info <- if (PERF) {
  sprintf("performance sizes %s; ",
          paste(format(SIZES, scientific = FALSE, trim = TRUE), collapse = ", "))
} else "performance report disabled; "
cat(sprintf(
  "Generated %s · device %s · anvl %s · JAX %s · %s · Rmpfr %s (%d-bit) · bench %s · google-benchmark.\n%s%d accuracy points per combination.\n",
  format(Sys.Date()), toupper(DEVICE), as.character(packageVersion("anvl")), JAX_VER,
  R.version.string, as.character(packageVersion("Rmpfr")), PREC,
  as.character(packageVersion("bench")), grid_info, length(zgrid)))
```
