---
title: "Automatic Differentiation"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

In this article, you will learn how to compute gradients of functions with {anvl}.
If you haven't yet, read the [Get Started](anvl.html) article first, which uses gradients to fit a linear model.

## Computing Gradients

`gradient()` takes a function and returns a new function that computes its gradient.
Consider $f(x) = \sum_i x_i^2$, whose gradient is $2x$:

```{r}
library(anvl)

f <- function(x) sum(x^2)
f_grad <- jit(gradient(f))

f_grad(nv_array(c(1, 2, 3)))
```

The gradient function has the same arguments as `f`, and returns a named list with one gradient per argument.
Each gradient has the same shape and data type as its argument.

Note that the result of `gradient()` must be jitted.
This is because {anvl} computes gradients by transforming the traced program of `f`, which only exists within `jit()`.

## Differentiable Functions

The function that is differentiated must return a single float scalar, such as a loss.
If you are interested in the derivative of a function that returns an array, reduce it to a scalar first.
For an elementwise function such as `sin()`, the gradient of `sum(sin(x))` is exactly the elementwise derivative `cos(x)`:

```{r, error = TRUE}
jit(gradient(function(x) sin(x)))(nv_array(c(0, pi)))

jit(gradient(function(x) sum(sin(x))))(nv_array(c(0, pi)))
```

Gradients can only be computed with respect to float arrays, because the derivative with respect to an integer or boolean is not defined.

```{r, error = TRUE}
jit(gradient(f))(nv_array(1:3))
```

## Selecting Arguments with `wrt`

By default, the gradient is computed with respect to all arguments.
The `wrt` argument selects a subset of them, by name or position.
This is necessary when some arguments are not float arrays, e.g. data that is not being optimized over, or static arguments:

```{r}
power_sum <- function(x, p) sum(x^p)
power_sum_grad <- jit(gradient(power_sum, wrt = "x"), static = "p")

power_sum_grad(nv_array(c(1, 2)), p = 3L)
```

## Nested Inputs

Arguments can also be (nested) lists of arrays, as is common for the parameters of a model.
The gradient then has the same structure as the argument:

```{r}
loss <- function(params, x, y) {
  y_hat <- x * params$w + params$b
  mean((y_hat - y)^2)
}
loss_grad <- jit(gradient(loss, wrt = "params"))

params <- list(w = nv_scalar(1), b = nv_scalar(0))
x <- nv_array(c(1, 2, 3))
y <- nv_array(c(2, 4, 6))

loss_grad(params, x, y)
```

This makes it easy to update all parameters at once, e.g. via `Map()`:

```{r}
grads <- loss_grad(params, x, y)$params
Map(\(p, g) p - 0.1 * g, params, grads)
```

## Value and Gradient

During training, one usually needs both the value of the loss and its gradient.
Instead of calling the function and its gradient separately, `value_and_gradient()` computes both in a single pass:

```{r}
loss_vg <- jit(value_and_gradient(loss, wrt = "params"))
loss_vg(params, x, y)
```

## Composing with `jit()`

`gradient()` works on any function that can be traced, including jitted ones, and the gradient function can itself be called within other jitted functions.
Because nested jitted functions are compiled together, it does not matter whether you jit `f`, its gradient, or the function that calls the gradient.
Typically, you jit the outermost function, e.g. a complete training step:

```{r}
train_step <- jit(function(params, x, y, lr) {
  out <- value_and_gradient(loss, wrt = "params")(params, x, y)
  list(
    loss = out$value,
    params = Map(\(p, g) p - lr * g, params, out$grad$params)
  )
})

for (i in 1:3) {
  step <- train_step(params, x, y, lr = 0.05)
  params <- step$params
  print(as_array(step$loss))
}
```

## Higher-Order Derivatives

Because a gradient function is itself an ordinary function, it can be differentiated again.
Below, we compute the second derivative of $x^3$, which is $6x$:

```{r}
cube_grad <- gradient(function(x) sum(x^3))
second_deriv <- jit(gradient(function(x) sum(cube_grad(x)$x)))

second_deriv(nv_array(c(1, 2)))
```

## Non-Differentiable Operations

Some operations have no meaningful derivative, e.g. comparisons.
Where they only select values, gradients flow through the selected values.
For example, in `nv_ifelse(x > 0, x, 0)`, the gradient is `1` where `x` is positive and `0` elsewhere:

```{r}
relu_sum <- function(x) sum(nv_ifelse(x > 0, x, 0))
jit(gradient(relu_sum))(nv_array(c(-1, 2)))
```

## Limitations

- {anvl} only implements *reverse-mode* differentiation, which is efficient for functions with many inputs and a single scalar output, such as losses.
  Forward-mode differentiation is not available yet.
- The control-flow primitives `nv_while()` and `nv_if()` cannot be differentiated yet.
  If you need gradients through a loop, use an R loop inside the jitted function instead, which is unrolled during tracing (see the [R loops are unrolled](jit.html#r-loops-are-unrolled) section of the JIT deep dive).

## How It Works

When `gradient(f)` is called within `jit()`, {anvl} first traces `f` into a graph of primitive operations, as described in the [JIT Deep Dive](jit.html) article.
Every primitive comes with a *reverse rule* that, given the gradient of the primitive's output, computes the gradients of its inputs.
{anvl} then walks the graph backwards from the output and applies these rules, which yields a new graph that computes the gradients.
To learn how to write a reverse rule for a new primitive, see the [Adding a Primitive](extending_primitive.html) article.
