---
title: "Custom Calls"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Custom Calls}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

{anvl} lowers every primitive to a StableHLO program, which XLA then compiles.
Some operations have no StableHLO counterpart, or one that is awkward or
numerically poor.
For those, XLA offers *custom calls*: you write a C++ function, register it
with the PJRT runtime under a name, and XLA invokes it at exactly that point
in the compiled program.

This article walks the whole path on CPU, and then runs the same operation
on a GPU with a hand-written CUDA kernel -- which needs no C++ handler and
no build setup at all.
{pjrt}'s [Adding Custom Calls](https://r-xla.github.io/pjrt/dev/articles/ffi.html)
article covers the C++ side in much more depth, including CUDA, LAPACK,
buffer aliasing and the full binding API.

Everything here is specific to the **`"pjrt"` backend**.
A custom call is a handler registered with the PJRT runtime and invoked by
the XLA-compiled program, so there is nothing for {anvl}'s `quickr` backend
to call: under `local_backend("quickr")` a function containing a custom call
fails to compile with *does not support these primitives: custom_call*.

## The example

We will implement `hypot(x, y)` -- the length of the hypotenuse,
`sqrt(x^2 + y^2)`, computed without letting the intermediate `x^2 + y^2`
over- or underflow.
StableHLO has no `hypot` op, and the obvious composition loses:

```{r}
library(anvl)
x <- nv_array(c(3, 1e-200), dtype = "f64")
y <- nv_array(c(4, 1e-200), dtype = "f64")
nv_sqrt(x^2 + y^2)
```

The second element should be roughly `1.41e-200`, but `x^2` underflowed to
zero before the square root ever ran; symmetrically, inputs above `1e154`
overflow to `Inf`.
`f32` has the same problem, just at a much smaller exponent -- `1e-30`
already squares to zero:

```{r}
xf <- nv_array(c(3, 1e-30), dtype = "f32")
yf <- nv_array(c(4, 1e-30), dtype = "f32")
nv_sqrt(xf^2 + yf^2)
```

C's `std::hypot` rescales its inputs and gets both right, so we call it --
for both dtypes, since a handler that only ever accepted `f64` would be of
little use in code that computes in single precision.

## Step 1: Write the handler

A custom call handler is a C++ function that receives one argument per
operand and one `Result<>` per result of the call, and returns an
`xla::ffi::Error`.
The buffers are already allocated by XLA; the handler only fills the results
in.

{pjrt} ships XLA's FFI headers in its `inst/include`, and the FFI API is
header-only, so `LinkingTo: pjrt` in a package -- or
`// [[Rcpp::depends(pjrt)]]` with `Rcpp::sourceCpp()`, as here -- is all that
is needed to compile against it.

```{Rcpp}
// [[Rcpp::plugins(cpp20)]]
// [[Rcpp::depends(pjrt)]]
#include <Rcpp.h>
#include <cmath>
#include "xla/ffi/api/ffi.h"

using namespace xla::ffi;

// The actual work, once the element type is known.
template <typename T>
static Error hypot_impl(AnyBuffer x, AnyBuffer y, Result<AnyBuffer> out) {
  size_t n = out->element_count();
  if (x.element_count() != n || y.element_count() != n) {
    return Error::InvalidArgument("hypot: operands must have the same size");
  }
  const T *xp = x.typed_data<T>();
  const T *yp = y.typed_data<T>();
  T *op = out->typed_data<T>();
  for (size_t i = 0; i < n; ++i) {
    op[i] = std::hypot(xp[i], yp[i]);
  }
  return Error::Success();
}

// Dispatch on the operand dtype; f32 and f64 are what std::hypot covers.
static Error do_hypot(AnyBuffer x, AnyBuffer y, Result<AnyBuffer> out) {
  DataType dt = x.element_type();
  if (y.element_type() != dt || out->element_type() != dt) {
    return Error::InvalidArgument("hypot: operands and result must have the "
                                  "same dtype");
  }
  switch (dt) {
    case DataType::F32:
      return hypot_impl<float>(x, y, out);
    case DataType::F64:
      return hypot_impl<double>(x, y, out);
    default:
      return Error::InvalidArgument("hypot: only f32 and f64 are supported");
  }
}

XLA_FFI_DEFINE_HANDLER(hypot_handler, do_hypot,
                       Ffi::Bind()
                           .Arg<AnyBuffer>()
                           .Arg<AnyBuffer>()
                           .Ret<AnyBuffer>());

// [[Rcpp::export]]
SEXP hypot_handler_ptr() {
  return R_MakeExternalPtr((void *)hypot_handler, R_NilValue, R_NilValue);
}
```

Three things are happening here:

- `hypot_impl<T>()` is the implementation, and `do_hypot()` picks the `T` at
  run time. `AnyBuffer` is a buffer of any dtype and rank, so the handler
  gets to look at `element_type()` and decide. The alternative is a
  statically typed `Buffer<DataType::F64>`, which pins the handler to one
  dtype and makes XLA reject everything else -- fine for a kernel that only
  ever runs in one precision, but not for one meant to be used from generic
  code.
- `XLA_FFI_DEFINE_HANDLER()` attaches the *binding specification* to it.
  Without it the runtime has no way to know that `do_hypot` wants two
  operands and one result, in that order. The macro defines `hypot_handler`,
  a plain function pointer.
- `hypot_handler_ptr()` hands that pointer to R as an external pointer,
  which is what the registration function below expects. In a package this
  getter is unnecessary: declare the handler with
  `XLA_FFI_DEFINE_HANDLER_SYMBOL()` instead and look the resulting
  `extern "C"` symbol up with `getNativeSymbolInfo()`.

`AnyBuffer` moves the type checking from XLA into the handler, so the
handler has to do it: `typed_data<T>()` only asserts that `T` matches in a
debug build, and reinterprets the bytes otherwise.
Hence the explicit comparison of the three `element_type()`s and the
`default:` branch -- an `i32` operand comes back as
*hypot: only f32 and f64 are supported* rather than as nonsense.

A handler may be called from XLA's worker threads, so it must not touch R's
C API (no allocation, no `Rf_error()`, no printing).
Report failures by returning an `Error` instead -- XLA aborts the execution
and {pjrt} re-raises it as an R error.

## Step 2: Register it with PJRT

```{r}
pjrt::pjrt_register_custom_call(
  "anvl_hypot",
  list(cpu = hypot_handler_ptr())
)
```

The target name is a process-global key, so prefix it with something
specific to your package.
The second argument is one handler per platform, `"cpu"` (a synonym for
PJRT's own `"host"`) and `"cuda"`.
Platforms you leave out simply have no implementation: running the call
there aborts with *No FFI handler registered for `anvl_hypot`*.

Registration is *deferred*: if the plugin for a platform is not loaded yet,
the handler is queued and registered when it is.
Calling this in a package's `.onLoad()` is therefore fine, and you should
pass `.package = pkgname` there so that the handler is dropped again when
the package is unloaded.

## Step 3: Call it

[`nv_custom_call()`] takes the target -- here the name the handler was
registered under -- the operands, and the shape and dtype of each result.
anvl cannot infer the result types -- the handler decides them -- but
nothing forces you to write them out as literals: [`dtype()`] and [`shape()`]
of the operands are available at trace time, so a generic call reads

```{r}
nv_custom_call("anvl_hypot", x, y, output_types = vt(dtype(x), shape(x)))
```

and works for `f32` just as well:

```{r}
nv_custom_call("anvl_hypot", xf, yf, output_types = vt(dtype(xf), shape(xf)))
```

Wrapping that up gives an ordinary-looking `nv_*` function.
The handler requires both operands to have the same dtype, and anvl's own
API functions promote before they compute, so the wrapper does too --
[`nv_promote_to_common()`] is the same helper the built-in binary operations
use:

```{r}
nv_hypot <- function(x, y) {
  operands <- nv_promote_to_common(x, y)
  nv_custom_call(
    "anvl_hypot",
    operands[[1L]],
    operands[[2L]],
    output_types = vt(dtype(operands[[1L]]), shape(operands[[1L]]))
  )
}

nv_hypot(x, y)
nv_hypot(xf, yf)
# an f32 and an f64 operand meet as f64
nv_hypot(xf, y)
```

That is the whole binding.
It works the same inside a JIT-compiled function, where the custom call
becomes one op among the others -- and since the output types are derived
from the traced operands, one `jit()`ed function serves both dtypes:

```{r}
f <- jit(function(x, y) nv_sum(nv_hypot(x, y)))
f(x, y)
f(xf, yf)
```

The arguments worth knowing about:

- `output_types` is a [`stablehlo::ValueType`], or a list of them for a
  handler with several results, most easily built with [`vt()`]. Pass a bare
  `vt()` and you get a bare array back; pass a list and you get a list.
- `attrs` is a named list of values the handler reads through
  `.Attr<T>("name")` or `.Attrs<Dictionary>()`. Scalars, strings and vectors
  are converted using anvl's usual dtype defaults; pass a
  [`stablehlo::ScalarAttr()`] and friends to pin a different one.
- `operand_layouts` / `result_layouts` fix the memory layout XLA hands the
  handler, in minor-to-major order. They default to row-major, which is what
  `hypot_impl()` -- walking its buffers linearly -- wants. A handler calling
  a column-major library such as LAPACK wants `c(0L, 1L)` instead.
- `aliases` says which operand's buffer each result is written into, so an
  in-place kernel does not force a copy. `aliases = 1L` on a single-result
  call hands the handler the same pointer for operand 1 and the result.
- `has_side_effect` keeps a call XLA would otherwise delete as unused. It
  defaults to `TRUE` exactly when `output_types` is `NULL`, which is how you
  write a printing or logging call; such a call returns its operands
  unchanged.

## Making it a primitive

`nv_custom_call()` is opaque to XLA, and therefore to {anvl}'s
differentiation: `gradient()` will not go through it.
It also shows up in a traced graph as a generic `custom_call` rather than
an operation with a name of its own.

Both are fixed the same way -- by giving the call its own primitive with a
reverse rule, exactly as in
[Adding a Primitive](extending_primitive.html).
The dtype stays generic here too: the shape rule propagates whatever dtype
the operands were traced with, and the lowering rule is handed the
`output_types` that follow from it.
That rule reaches one level lower than `nv_custom_call()` does, emitting
[`stablehlo::hlo_custom_call()`] directly, because a rule is handed
`FuncValue`s rather than arrays:

```{r}
prim_hypot <- new_primitive("hypot", function(x, y) {
  graph_desc_add(self, list(x = x, y = y), infer_fn = function(x, y) {
    list(AbstractArray(dtype = dtype(x), shape = shape(x)))
  })[[1L]]
})

prim_hypot[["stablehlo"]] <- function(x, y, output_types) {
  row_major <- rev(seq_along(shape(x)) - 1L)
  list(stablehlo::hlo_custom_call(
    x,
    y,
    call_target_name = "anvl_hypot",
    api_version = 4L,
    has_side_effect = FALSE,
    output_types = output_types,
    operand_layouts = list(row_major, row_major),
    result_layouts = list(row_major)
  ))
}

# d/dx hypot(x, y) = x / hypot(x, y), and the forward result is already there
prim_hypot[["reverse"]] <- rule_reverse(function(
  inputs,
  outputs,
  grads,
  params,
  required
) {
  x <- inputs[[1L]]
  y <- inputs[[2L]]
  r <- outputs[[1L]]
  grad <- grads[[1L]]
  list(
    if (required[[1L]]) prim_mul(grad, prim_div(x, r)),
    if (required[[2L]]) prim_mul(grad, prim_div(y, r))
  )
})
```

Now the call differentiates, in either precision:

```{r}
g <- function(x, y) nv_sum(prim_hypot(x, y))
a <- nv_array(c(3, 6), dtype = "f64")
b <- nv_array(c(4, 8), dtype = "f64")
jit(gradient(g))(a, b)
jit(gradient(g))(nv_convert(a, "f32"), nv_convert(b, "f32"))
```

A real `nv_hypot()` would then promote its arguments as above and call
`prim_hypot()` instead of `nv_custom_call()`: the promotion belongs in the
`nv_*` layer, the primitive takes the dtypes as they come.

## Shipping this in a package

`Rcpp::sourceCpp()` keeps this article self-contained, but in a package you
would:

1. put the `.cpp` file in `src/`, declare the handler with
   `XLA_FFI_DEFINE_HANDLER_SYMBOL()`, and add `pjrt` to `LinkingTo:` and
   `C++20` to `SystemRequirements:`,
2. register it from `.onLoad()`, passing `.package = pkgname` so the handler
   goes away with the package:

   ```r
   .onLoad <- function(libname, pkgname) {
     pjrt::pjrt_register_custom_call(
       "anvl_hypot",
       list(cpu = getNativeSymbolInfo("HypotHandler", PACKAGE = pkgname)$address),
       .package = pkgname
     )
   }
   ```

## Going to CUDA

On a GPU you do not need an FFI handler at all.
{pjrt} ships one generic handler, `pjrt_cuda_kernel`, that launches any CUDA
kernel you hand it -- the same design as jaxlib's `triton_kernel_call` and
CuPy's `RawKernel`.
So the CUDA version of `hypot` is only the kernel itself:

```{r}
hypot_cuda <- pjrt::pjrt_cuda_module(
  r"(
template <typename T>
__global__ void hypot_kernel(const T *x, const T *y, T *out, int n) {
  int i = blockIdx.x * blockDim.x + threadIdx.x;
  if (i < n) out[i] = hypot(x[i], y[i]);
}
)",
  kernels = c("hypot_kernel<float>", "hypot_kernel<double>")
)
hypot_cuda
```

[`pjrt::pjrt_cuda_module()`] only records the source; it needs neither a GPU
nor a compiler.
The first time one of its kernels runs, {pjrt} compiles the module with
NVRTC -- CUDA's runtime compiler, which the CUDA plugin's `pjrt.cuda`
package already ships, so there is no CUDA toolkit to install and no `nvcc`
to call -- for exactly the GPU it runs on, and caches the result on disk.
`kernels` lists the template instantiations to compile; a kernel declared
`extern "C"` needs no entry.

The kernel receives the call's operands, then its results, as device
pointers, followed by any scalar arguments -- here the element count `n`.
[`cuda_kernel()`] says how to launch it, and goes where the handler name went
before.
Because the platform is only known when the program is lowered, `target`
can also list one target per platform, and the same `nv_hypot()` then runs
the C++ handler on the CPU and the kernel on a GPU:

```{r}
nv_hypot <- function(x, y) {
  operands <- nv_promote_to_common(x, y)
  x <- operands[[1L]]
  y <- operands[[2L]]
  n <- as.integer(prod(shape(x)))
  ctype <- switch(as.character(dtype(x)), f32 = "float", f64 = "double")
  nv_custom_call(
    list(
      cpu = "anvl_hypot",
      cuda = cuda_kernel(
        hypot_cuda,
        sprintf("hypot_kernel<%s>", ctype),
        grid = ceiling(n / 256),
        block = 256L,
        scalars = list(n)
      )
    ),
    x,
    y,
    output_types = vt(dtype(x), shape(x))
  )
}

nv_hypot(x, y)
```

That ran on the CPU. With the operands on a GPU, the kernel runs instead:

```{r, eval = FALSE}
xc <- nv_array(c(3, 1e-200), dtype = "f64", device = "cuda")
yc <- nv_array(c(4, 1e-200), dtype = "f64", device = "cuda")
nv_hypot(xc, yc)
jit(function(x, y) nv_sum(nv_hypot(x, y)))(xc, yc)
```

A few things to know about the kernel's side:

- **Scalars** map from R to C as in R's `.C()` interface: an integer is an
  `int`, a double a `double`, a logical a `bool`. That is why `n` is passed
  through `as.integer()`. Anything else is written with
  [`pjrt::pjrt_cuda_scalar()`], e.g. `pjrt_cuda_scalar(n, "i64")` for a
  `long long`.
- **Mismatches are caught.** Before launching, {pjrt} compares the call with
  the kernel's compiled signature, so passing `prod(shape(x))` (a double)
  for the `int n` fails with *Parameter 4 of CUDA kernel 'hypot_kernel<double>'
  is 4 bytes, but a scalar of 8 bytes was passed* instead of reading garbage.
  Compilation errors come back with NVRTC's log.
- **The launch configuration is static**: `grid` and `block` are fixed when
  the call is traced. That is always possible, because shapes are static,
  but `grid` has to be computed from `shape()`, never from values.
- **Layouts** are row-major by default, as for any custom call.

### Shipping kernels in a package

The CUDA side of a package needs no build system changes at all: nothing is
compiled at install time, and the package installs and loads on machines
without a GPU.
Keep the kernel in `inst/cuda/hypot.cu` and create the module when the
package loads, naming the package as its owner:

```r
the <- new.env()

.onLoad <- function(libname, pkgname) {
  the$hypot <- pjrt::pjrt_cuda_module(
    file = system.file("cuda", "hypot.cu", package = pkgname),
    kernels = c("hypot_kernel<float>", "hypot_kernel<double>"),
    package = pkgname
  )
}
```

That alone works: each user's machine compiles the module once, on first
use, for its own GPU, and later sessions load it from the cache
([`pjrt::pjrt_cuda_cache_dir()`]).

To spare users even that, ship the kernels precompiled, as {pjrt} does for
its own.
[`pjrt::pjrt_cuda_build_kernels()`] compiles the package's modules -- a
cubin per GPU architecture, plus PTX for later ones -- into a tarball, which
you attach to each release (it needs NVRTC, but no GPU, so an ordinary CI
runner does; {pjrt}'s `cuda-kernels` workflow is one to copy).
Name its URL in DESCRIPTION:

```
Config/pjrt/cuda-kernels: https://github.com/org/pkg/releases/download/v{version}/pkg-cuda-kernels-{version}.tar.gz
```

When the CUDA plugin loads, {pjrt} downloads the tarball once, verifies it,
and loads the kernels from it. Whatever it does not cover -- a development
version, a GPU it was not built for, a machine without network -- is
compiled locally instead, so nothing breaks without it.

### When you need a handler after all

`pjrt_cuda_kernel` launches one kernel on XLA's stream.
Work that needs more than that -- calling a CUDA library such as cuBLAS or
cuSOLVER, scratch memory, several dependent launches -- still takes an FFI
handler bound with `.Ctx<PlatformStream<void *>>()` and registered under
`cuda` in `pjrt_register_custom_call()`.
{pjrt}'s [Adding Custom Calls](https://r-xla.github.io/pjrt/dev/articles/ffi.html)
article covers how, using its built-in cuSOLVER handlers as the example.
