---
title: "Custom Calls"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Custom Calls}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

{anvl} lowers every primitive to a StableHLO program, which XLA then compiles.
Some operations have no StableHLO counterpart, or one that is awkward or
numerically poor.
For those, XLA offers *custom calls*: you write a C++ function, register it
with the PJRT runtime under a name, and `stablehlo.custom_call @name(...)`
invokes it at exactly that point in the compiled program.

This article walks the whole path on CPU.
It assumes you have read [Adding a Primitive](extending_primitive.html); the
custom call replaces only *Step 2*, the StableHLO lowering rule.
{pjrt}'s [Adding Custom Calls](https://r-xla.github.io/pjrt/dev/articles/ffi.html)
article covers the C++ side in much more depth, including CUDA, LAPACK,
attributes and buffer aliasing.

## The example

We will implement `hypot(x, y)` -- the length of the hypotenuse,
`sqrt(x^2 + y^2)`, computed without letting the intermediate `x^2 + y^2`
over- or underflow.
StableHLO has no `hypot` op, and the obvious composition loses:

```{r}
library(anvl)
x <- nv_array(c(3, 1e-200), dtype = "f64")
y <- nv_array(c(4, 1e-200), dtype = "f64")
nv_sqrt(x^2 + y^2)
```

The second element should be roughly `1.41e-200`, but `x^2` underflowed to
zero before the square root ever ran; symmetrically, inputs above `1e154`
overflow to `Inf`.
C's `std::hypot` rescales its inputs and gets both right, so we call it.

## Step 1: Write the handler

A custom call handler is a C++ function that receives one argument per
operand and one `Result<>` per result of the `stablehlo.custom_call` op, and
returns an `xla::ffi::Error`.
The buffers are already allocated by XLA; the handler only fills the results
in.

{pjrt} ships XLA's FFI headers in its `inst/include`, and the FFI API is
header-only, so `LinkingTo: pjrt` in a package -- or
`// [[Rcpp::depends(pjrt)]]` with `Rcpp::sourceCpp()`, as here -- is all that
is needed to compile against it.

```{Rcpp}
// [[Rcpp::plugins(cpp20)]]
// [[Rcpp::depends(pjrt)]]
#include <Rcpp.h>
#include <cmath>
#include "xla/ffi/api/ffi.h"

using namespace xla::ffi;

static Error do_hypot(Buffer<DataType::F64> x, Buffer<DataType::F64> y,
                      Result<Buffer<DataType::F64>> out) {
  size_t n = out->element_count();
  if (x.element_count() != n || y.element_count() != n) {
    return Error::InvalidArgument("hypot: operands must have the same size");
  }
  const double *xp = x.typed_data();
  const double *yp = y.typed_data();
  double *op = out->typed_data();
  for (size_t i = 0; i < n; ++i) {
    op[i] = std::hypot(xp[i], yp[i]);
  }
  return Error::Success();
}

XLA_FFI_DEFINE_HANDLER(hypot_handler, do_hypot,
                       Ffi::Bind()
                           .Arg<Buffer<DataType::F64>>()
                           .Arg<Buffer<DataType::F64>>()
                           .Ret<Buffer<DataType::F64>>());

// [[Rcpp::export]]
SEXP hypot_handler_ptr() {
  return R_MakeExternalPtr((void *)hypot_handler, R_NilValue, R_NilValue);
}
```

Three things are happening here:

- `do_hypot()` is the implementation. `Buffer<DataType::F64>` is a
  statically typed buffer -- the handler only accepts `f64` and XLA rejects
  the call otherwise. Use `AnyBuffer` instead if you want to dispatch on
  dtype and rank yourself.
- `XLA_FFI_DEFINE_HANDLER()` attaches the *binding specification* to it.
  Without it the runtime has no way to know that `do_hypot` wants two
  operands and one result, in that order. The macro defines `hypot_handler`,
  a plain function pointer.
- `hypot_handler_ptr()` hands that pointer to R as an external pointer, which
  is what the registration function below expects.

A handler may be called from XLA's worker threads, so it must not touch R's
C API (no allocation, no `Rf_error()`, no printing).
Report failures by returning an `Error` instead -- XLA aborts the execution
and {pjrt} re-raises it as an R error.
The FFI decoder does the same for a call that does not match the binding, so
handing this handler an `f32` array fails with *Wrong buffer dtype: expected
F64 but got F32* rather than reading nonsense.

## Step 2: Register it with PJRT

```{r}
pjrt::pjrt_register_custom_call(
  "anvl_hypot",
  list(cpu = hypot_handler_ptr())
)
```

The target name is a process-global key, so prefix it with something
specific to your package.
The second argument is one handler per platform, `"cpu"` (a synonym for
PJRT's own `"host"`) and `"cuda"`.
Platforms you leave out simply have no implementation: running the call
there aborts with *No FFI handler registered for `anvl_hypot`*.

Registration is *deferred*: if the plugin for a platform is not loaded yet,
the handler is queued and registered when it is.
Calling this in a package's `.onLoad()` is therefore fine, and you should
pass `.package = pkgname` there so that the handler is dropped again when
the package is unloaded.

## Step 3: Define the primitive

Nothing here is custom-call specific -- this is the usual `new_primitive()`
from [Adding a Primitive](extending_primitive.html).
The inference function pins down what the handler can actually cope with:
two `f64` arrays of the same shape.

```{r}
prim_hypot <- new_primitive("hypot", function(x, y) {
  graph_desc_add(self, list(x = x, y = y), infer_fn = function(x, y) {
    if (!identical(shape(x), shape(y))) {
      cli::cli_abort("{.arg x} and {.arg y} must have the same shape.")
    }
    if (dtype(x) != "f64" || dtype(y) != "f64") {
      cli::cli_abort("{.arg x} and {.arg y} must be {.val f64}.")
    }
    list(AbstractArray(dtype = dtype(x), shape = shape(x)))
  })[[1L]]
})
```

## Step 4: Lower it to a custom call

The StableHLO rule emits `stablehlo::hlo_custom_call()` instead of a regular
op:

```{r}
prim_hypot[["stablehlo"]] <- function(x, y, output_types) {
  # row-major, in minor-to-major order: [N-1, ..., 1, 0]
  row_major <- rev(seq_along(shape(x)) - 1L)

  list(stablehlo::hlo_custom_call(
    x,
    y,
    call_target_name = "anvl_hypot",
    api_version = 4L,
    has_side_effect = FALSE,
    output_types = output_types,
    operand_layouts = list(row_major, row_major),
    result_layouts = list(row_major)
  ))
}
```

The arguments that matter:

- `call_target_name` must be the name the handler was registered under.
- `api_version = 4L` selects the FFI calling convention that
  `XLA_FFI_DEFINE_HANDLER()` implements. Always 4.
- `has_side_effect = FALSE` lets XLA drop the call if its result is unused.
  Set it to `TRUE` for handlers that exist *only* for their effect -- a
  printing or logging call -- or the optimizer will delete them.
- `output_types` is a list of [`stablehlo::ValueType`]s. Declaring an
  `output_types` formal on the rule makes {anvl} pass in the types it already
  inferred in Step 3, so there is nothing to recompute. Elsewhere, build them
  with [`vt()`].
- `operand_layouts` / `result_layouts` fix the memory layout XLA hands to the
  handler, in minor-to-major order. `do_hypot()` walks all three buffers
  linearly, so all it needs is that they agree; for a handler that calls a
  column-major library such as LAPACK, this is where you would ask for
  `c(0L, 1L)` instead.

## Step 5: Give it a gradient

The custom call is opaque to XLA, so it is opaque to {anvl}'s
differentiation as well: without a reverse rule, `gradient()` will not go
through it.
The rule is written exactly as for any other primitive.
Since `d/dx hypot(x, y) = x / hypot(x, y)`, we can reuse the value the
forward pass already computed, which arrives in `outputs`:

```{r}
prim_hypot[["reverse"]] <- rule_reverse(function(
  inputs,
  outputs,
  grads,
  params,
  required
) {
  x <- inputs[[1L]]
  y <- inputs[[2L]]
  r <- outputs[[1L]]
  grad <- grads[[1L]]
  list(
    if (required[[1L]]) prim_mul(grad, prim_div(x, r)),
    if (required[[2L]]) prim_mul(grad, prim_div(y, r))
  )
})
```

## Using it

Eagerly:

```{r}
prim_hypot(x, y)
```

Inside a JIT-compiled function, where the custom call is one op among
others:

```{r}
f <- function(x, y) nv_reduce_sum(prim_hypot(x, y))
jit(f)(x, y)
```

And differentiated:

```{r}
a <- nv_array(c(3, 6), dtype = "f64")
b <- nv_array(c(4, 8), dtype = "f64")
jit(gradient(f))(a, b)
```

## Shipping this in a package

`Rcpp::sourceCpp()` keeps this article self-contained, but in a package you
would:

1. put the `.cpp` file in `src/` and add `pjrt` to `LinkingTo:` (and
   `SystemRequirements: C++20`),
2. register the handler from `.onLoad()`, passing `.package = pkgname`:

   ```r
   .onLoad <- function(libname, pkgname) {
     pjrt::pjrt_register_custom_call(
       "anvl_hypot",
       list(cpu = hypot_handler_ptr()),
       .package = pkgname
     )
   }
   ```

3. define the primitive and its rules at build time as usual.

To also support CUDA, register a second handler under `cuda`; see the
[{pjrt} article](https://r-xla.github.io/pjrt/dev/articles/ffi.html) for what
changes on the GPU side.
