---
title: "Custom Calls"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Custom Calls}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

{anvl} lowers every primitive to a StableHLO program, which XLA then compiles.
Some operations have no StableHLO counterpart, or one that is awkward or
numerically poor.
For those, XLA offers *custom calls*: you write a C++ function, register it
with the PJRT runtime under a name, and XLA invokes it at exactly that point
in the compiled program.

This article walks the whole path on CPU.
{pjrt}'s [Adding Custom Calls](https://r-xla.github.io/pjrt/dev/articles/ffi.html)
article covers the C++ side in much more depth, including CUDA, LAPACK,
buffer aliasing and the full binding API.

Everything here is specific to the **`"pjrt"` backend**.
A custom call is a handler registered with the PJRT runtime and invoked by
the XLA-compiled program, so there is nothing for {anvl}'s `quickr` backend
to call: under `local_backend("quickr")` a function containing a custom call
fails to compile with *does not support these primitives: custom_call*.

## The example

We will implement `hypot(x, y)` -- the length of the hypotenuse,
`sqrt(x^2 + y^2)`, computed without letting the intermediate `x^2 + y^2`
over- or underflow.
StableHLO has no `hypot` op, and the obvious composition loses:

```{r}
library(anvl)
x <- nv_array(c(3, 1e-200), dtype = "f64")
y <- nv_array(c(4, 1e-200), dtype = "f64")
nv_sqrt(x^2 + y^2)
```

The second element should be roughly `1.41e-200`, but `x^2` underflowed to
zero before the square root ever ran; symmetrically, inputs above `1e154`
overflow to `Inf`.
`f32` has the same problem, just at a much smaller exponent -- `1e-30`
already squares to zero:

```{r}
xf <- nv_array(c(3, 1e-30), dtype = "f32")
yf <- nv_array(c(4, 1e-30), dtype = "f32")
nv_sqrt(xf^2 + yf^2)
```

C's `std::hypot` rescales its inputs and gets both right, so we call it --
for both dtypes, since a handler that only ever accepted `f64` would be of
little use in code that computes in single precision.

## Step 1: Write the handler

A custom call handler is a C++ function that receives one argument per
operand and one `Result<>` per result of the call, and returns an
`xla::ffi::Error`.
The buffers are already allocated by XLA; the handler only fills the results
in.

{pjrt} ships XLA's FFI headers in its `inst/include`, and the FFI API is
header-only, so `LinkingTo: pjrt` in a package -- or
`// [[Rcpp::depends(pjrt)]]` with `Rcpp::sourceCpp()`, as here -- is all that
is needed to compile against it.

```{Rcpp}
// [[Rcpp::plugins(cpp20)]]
// [[Rcpp::depends(pjrt)]]
#include <Rcpp.h>
#include <cmath>
#include "xla/ffi/api/ffi.h"

using namespace xla::ffi;

// The actual work, once the element type is known.
template <typename T>
static Error hypot_impl(AnyBuffer x, AnyBuffer y, Result<AnyBuffer> out) {
  size_t n = out->element_count();
  if (x.element_count() != n || y.element_count() != n) {
    return Error::InvalidArgument("hypot: operands must have the same size");
  }
  const T *xp = x.typed_data<T>();
  const T *yp = y.typed_data<T>();
  T *op = out->typed_data<T>();
  for (size_t i = 0; i < n; ++i) {
    op[i] = std::hypot(xp[i], yp[i]);
  }
  return Error::Success();
}

// Dispatch on the operand dtype; f32 and f64 are what std::hypot covers.
static Error do_hypot(AnyBuffer x, AnyBuffer y, Result<AnyBuffer> out) {
  DataType dt = x.element_type();
  if (y.element_type() != dt || out->element_type() != dt) {
    return Error::InvalidArgument("hypot: operands and result must have the "
                                  "same dtype");
  }
  switch (dt) {
    case DataType::F32:
      return hypot_impl<float>(x, y, out);
    case DataType::F64:
      return hypot_impl<double>(x, y, out);
    default:
      return Error::InvalidArgument("hypot: only f32 and f64 are supported");
  }
}

XLA_FFI_DEFINE_HANDLER(hypot_handler, do_hypot,
                       Ffi::Bind()
                           .Arg<AnyBuffer>()
                           .Arg<AnyBuffer>()
                           .Ret<AnyBuffer>());

// [[Rcpp::export]]
SEXP hypot_handler_ptr() {
  return R_MakeExternalPtr((void *)hypot_handler, R_NilValue, R_NilValue);
}
```

Three things are happening here:

- `hypot_impl<T>()` is the implementation, and `do_hypot()` picks the `T` at
  run time. `AnyBuffer` is a buffer of any dtype and rank, so the handler
  gets to look at `element_type()` and decide. The alternative is a
  statically typed `Buffer<DataType::F64>`, which pins the handler to one
  dtype and makes XLA reject everything else -- fine for a kernel that only
  ever runs in one precision, but not for one meant to be used from generic
  code.
- `XLA_FFI_DEFINE_HANDLER()` attaches the *binding specification* to it.
  Without it the runtime has no way to know that `do_hypot` wants two
  operands and one result, in that order. The macro defines `hypot_handler`,
  a plain function pointer.
- `hypot_handler_ptr()` hands that pointer to R as an external pointer,
  which is what the registration function below expects. In a package this
  getter is unnecessary: declare the handler with
  `XLA_FFI_DEFINE_HANDLER_SYMBOL()` instead and look the resulting
  `extern "C"` symbol up with `getNativeSymbolInfo()`.

`AnyBuffer` moves the type checking from XLA into the handler, so the
handler has to do it: `typed_data<T>()` only asserts that `T` matches in a
debug build, and reinterprets the bytes otherwise.
Hence the explicit comparison of the three `element_type()`s and the
`default:` branch -- an `i32` operand comes back as
*hypot: only f32 and f64 are supported* rather than as nonsense.

A handler may be called from XLA's worker threads, so it must not touch R's
C API (no allocation, no `Rf_error()`, no printing).
Report failures by returning an `Error` instead -- XLA aborts the execution
and {pjrt} re-raises it as an R error.

## Step 2: Register it with PJRT

```{r}
pjrt::pjrt_register_custom_call(
  "anvl_hypot",
  list(cpu = hypot_handler_ptr())
)
```

The target name is a process-global key, so prefix it with something
specific to your package.
The second argument is one handler per platform, `"cpu"` (a synonym for
PJRT's own `"host"`) and `"cuda"`.
Platforms you leave out simply have no implementation: running the call
there aborts with *No FFI handler registered for `anvl_hypot`*.

Registration is *deferred*: if the plugin for a platform is not loaded yet,
the handler is queued and registered when it is.
Calling this in a package's `.onLoad()` is therefore fine, and you should
pass `.package = pkgname` there so that the handler is dropped again when
the package is unloaded.

## Step 3: Call it

[`nv_custom_call()`] takes the target name, the operands, and the shape and
dtype of each result.
anvl cannot infer the result types -- the handler decides them -- but
nothing forces you to write them out as literals: [`dtype()`] and [`shape()`]
of the operands are available at trace time, so a generic call reads

```{r}
nv_custom_call("anvl_hypot", x, y, output_types = vt(dtype(x), shape(x)))
```

and works for `f32` just as well:

```{r}
nv_custom_call("anvl_hypot", xf, yf, output_types = vt(dtype(xf), shape(xf)))
```

Wrapping that up gives an ordinary-looking `nv_*` function.
The handler requires both operands to have the same dtype, and anvl's own
API functions promote before they compute, so the wrapper does too --
[`nv_promote_to_common()`] is the same helper the built-in binary operations
use:

```{r}
nv_hypot <- function(x, y) {
  operands <- nv_promote_to_common(x, y)
  nv_custom_call(
    "anvl_hypot",
    operands[[1L]],
    operands[[2L]],
    output_types = vt(dtype(operands[[1L]]), shape(operands[[1L]]))
  )
}

nv_hypot(x, y)
nv_hypot(xf, yf)
# an f32 and an f64 operand meet as f64
nv_hypot(xf, y)
```

That is the whole binding.
It works the same inside a JIT-compiled function, where the custom call
becomes one op among the others -- and since the output types are derived
from the traced operands, one `jit()`ed function serves both dtypes:

```{r}
f <- jit(function(x, y) nv_reduce_sum(nv_hypot(x, y)))
f(x, y)
f(xf, yf)
```

The arguments worth knowing about:

- `output_types` is a [`stablehlo::ValueType`], or a list of them for a
  handler with several results, most easily built with [`vt()`]. Pass a bare
  `vt()` and you get a bare array back; pass a list and you get a list.
- `attrs` is a named list of values the handler reads through
  `.Attr<T>("name")` or `.Attrs<Dictionary>()`. Scalars, strings and vectors
  are converted using anvl's usual dtype defaults; pass a
  [`stablehlo::ScalarAttr()`] and friends to pin a different one.
- `operand_layouts` / `result_layouts` fix the memory layout XLA hands the
  handler, in minor-to-major order. They default to row-major, which is what
  `hypot_impl()` -- walking its buffers linearly -- wants. A handler calling
  a column-major library such as LAPACK wants `c(0L, 1L)` instead.
- `aliases` says which operand's buffer each result is written into, so an
  in-place kernel does not force a copy. `aliases = 1L` on a single-result
  call hands the handler the same pointer for operand 1 and the result.
- `has_side_effect` keeps a call XLA would otherwise delete as unused. It
  defaults to `TRUE` exactly when `output_types` is `NULL`, which is how you
  write a printing or logging call; such a call returns its operands
  unchanged.

## Making it a primitive

`nv_custom_call()` is opaque to XLA, and therefore to {anvl}'s
differentiation: `gradient()` will not go through it.
It also shows up in a traced graph as a generic `custom_call` rather than
an operation with a name of its own.

Both are fixed the same way -- by giving the call its own primitive with a
reverse rule, exactly as in
[Adding a Primitive](extending_primitive.html).
The dtype stays generic here too: the shape rule propagates whatever dtype
the operands were traced with, and the lowering rule is handed the
`output_types` that follow from it.
That rule reaches one level lower than `nv_custom_call()` does, emitting
[`stablehlo::hlo_custom_call()`] directly, because a rule is handed
`FuncValue`s rather than arrays:

```{r}
prim_hypot <- new_primitive("hypot", function(x, y) {
  graph_desc_add(self, list(x = x, y = y), infer_fn = function(x, y) {
    list(AbstractArray(dtype = dtype(x), shape = shape(x)))
  })[[1L]]
})

prim_hypot[["stablehlo"]] <- function(x, y, output_types) {
  row_major <- rev(seq_along(shape(x)) - 1L)
  list(stablehlo::hlo_custom_call(
    x,
    y,
    call_target_name = "anvl_hypot",
    api_version = 4L,
    has_side_effect = FALSE,
    output_types = output_types,
    operand_layouts = list(row_major, row_major),
    result_layouts = list(row_major)
  ))
}

# d/dx hypot(x, y) = x / hypot(x, y), and the forward result is already there
prim_hypot[["reverse"]] <- rule_reverse(function(
  inputs,
  outputs,
  grads,
  params,
  required
) {
  x <- inputs[[1L]]
  y <- inputs[[2L]]
  r <- outputs[[1L]]
  grad <- grads[[1L]]
  list(
    if (required[[1L]]) prim_mul(grad, prim_div(x, r)),
    if (required[[2L]]) prim_mul(grad, prim_div(y, r))
  )
})
```

Now the call differentiates, in either precision:

```{r}
g <- function(x, y) nv_reduce_sum(prim_hypot(x, y))
a <- nv_array(c(3, 6), dtype = "f64")
b <- nv_array(c(4, 8), dtype = "f64")
jit(gradient(g))(a, b)
jit(gradient(g))(nv_convert(a, "f32"), nv_convert(b, "f32"))
```

A real `nv_hypot()` would then promote its arguments as above and call
`prim_hypot()` instead of `nv_custom_call()`: the promotion belongs in the
`nv_*` layer, the primitive takes the dtypes as they come.

## Shipping this in a package

`Rcpp::sourceCpp()` keeps this article self-contained, but in a package you
would:

1. put the `.cpp` file in `src/`, declare the handler with
   `XLA_FFI_DEFINE_HANDLER_SYMBOL()`, and add `pjrt` to `LinkingTo:` and
   `C++20` to `SystemRequirements:`,
2. register it from `.onLoad()`, passing `.package = pkgname` so the handler
   goes away with the package:

   ```r
   .onLoad <- function(libname, pkgname) {
     pjrt::pjrt_register_custom_call(
       "anvl_hypot",
       list(cpu = getNativeSymbolInfo("HypotHandler", PACKAGE = pkgname)$address),
       .package = pkgname
     )
   }
   ```

## Going to CUDA

Nothing on the R side changes for a GPU.
A `stablehlo.custom_call` names a *target*, not an implementation, and
`pjrt_register_custom_call()` already takes one handler per platform, so the
same `nv_hypot(x, y)` runs the CUDA handler as soon as its operands live on
a CUDA device:

```{r, eval = FALSE}
x <- nv_array(c(3, 1e-200), dtype = "f64", device = "cuda")
y <- nv_array(c(4, 1e-200), dtype = "f64", device = "cuda")
nv_hypot(x, y)
```

What changes is the handler, and above all how it is built: R's build system
knows how to compile `.cpp`, not `.cu`.

### Keep the kernel and the handler in separate files

Put the kernel in a `.cu` compiled by `nvcc`, and the FFI handler in a
`.cpp` compiled by R's ordinary C++ toolchain, with one `extern "C"`
launcher bridging them.
Typing the stream as `void *` at that boundary means the `.cpp` never
includes a CUDA header and the `.cu` never includes an XLA one -- which
matters, because `nvcc`'s C++ support trails the host compiler's.
This is the same split {pjrt} uses for its own CUDA handlers.

The kernel itself is a template, but `extern "C"` cannot be, so the `.cu`
exposes one launcher per supported dtype:

```c++
// src/hypot_kernel.cu -- compiled by nvcc
#include <cuda_runtime.h>

template <typename T>
__global__ void hypot_kernel(const T *x, const T *y, T *out, size_t n) {
  size_t i = blockIdx.x * blockDim.x + threadIdx.x;
  if (i < n) out[i] = hypot(x[i], y[i]);
}

template <typename T>
static void launch(void *stream, const T *x, const T *y, T *out, size_t n) {
  const int threads = 256;
  const int blocks = static_cast<int>((n + threads - 1) / threads);
  hypot_kernel<T><<<blocks, threads, 0, static_cast<cudaStream_t>(stream)>>>(
      x, y, out, n);
}

// The only symbols the rest of the package sees.
extern "C" void anvl_launch_hypot_f32(void *stream, const float *x,
                                      const float *y, float *out, size_t n) {
  launch<float>(stream, x, y, out, n);
}

extern "C" void anvl_launch_hypot_f64(void *stream, const double *x,
                                      const double *y, double *out, size_t n) {
  launch<double>(stream, x, y, out, n);
}
```

```c++
// src/hypot_cuda.cpp -- compiled by R's C++ toolchain, no CUDA headers
#include "xla/ffi/api/ffi.h"

using namespace xla::ffi;

extern "C" void anvl_launch_hypot_f32(void *stream, const float *x,
                                      const float *y, float *out, size_t n);
extern "C" void anvl_launch_hypot_f64(void *stream, const double *x,
                                      const double *y, double *out, size_t n);

static Error do_hypot_cuda(void *stream, AnyBuffer x, AnyBuffer y,
                           Result<AnyBuffer> out) {
  size_t n = out->element_count();
  if (x.element_count() != n || y.element_count() != n) {
    return Error::InvalidArgument("hypot: operands must have the same size");
  }
  DataType dt = x.element_type();
  if (y.element_type() != dt || out->element_type() != dt) {
    return Error::InvalidArgument("hypot: operands and result must have the "
                                  "same dtype");
  }
  // these are *device* pointers; only the kernel may dereference them
  switch (dt) {
    case DataType::F32:
      anvl_launch_hypot_f32(stream, x.typed_data<float>(),
                            y.typed_data<float>(), out->typed_data<float>(), n);
      return Error::Success();
    case DataType::F64:
      anvl_launch_hypot_f64(stream, x.typed_data<double>(),
                            y.typed_data<double>(), out->typed_data<double>(),
                            n);
      return Error::Success();
    default:
      return Error::InvalidArgument("hypot: only f32 and f64 are supported");
  }
}

XLA_FFI_DEFINE_HANDLER_SYMBOL(HypotHandlerCuda, do_hypot_cuda,
                              Ffi::Bind()
                                  .Ctx<PlatformStream<void *>>()
                                  .Arg<AnyBuffer>()
                                  .Arg<AnyBuffer>()
                                  .Ret<AnyBuffer>());
```

The binding gains one entry: `.Ctx<PlatformStream<void *>>()` puts the CUDA
stream XLA dispatched the op on in front of the buffers.
Everything the handler enqueues must go on *that* stream.

### You do not need a system CUDA toolkit

Installing {pjrt}'s CUDA plugin also installs the
[`cuda12.8`](https://mlverse.r-universe.dev/cuda12.8) R package, which
downloads the CUDA redistributables from PyPI at install time -- and it keeps
the `include/` directory of every component it unpacks, plus `nvcc` itself:

```r
cuda12.8::bin_path("nvcc")           # nvcc, ptxas, cicc
cuda12.8::include_path("runtime")    # cuda_runtime.h and friends
cuda12.8::include_path("cublas")     # likewise for cublas, cusolver, cudnn, ...
cuda12.8::lib_path()                 # all the .so files, libcudart.so included
```

So everything the kernel above needs to compile and link is already on the
machine that runs the plugin, and a package can find it from R instead of
guessing at `/usr/local/cuda`.
Two caveats: the package is Linux and Windows only (no macOS), and its
version is pinned -- {pjrt} defaults to `cuda12.8` and only that version is
known to match the plugin, though `PJRT_CUDA_R_PACKAGE` can point at another
one.

### Teaching `R CMD INSTALL` about `.cu`

R has no rule for `.cu`, so `src/Makevars` supplies one and lists every
object explicitly:

```make
CUDA_BIN = $(shell "$(R_HOME)/bin/Rscript" -e 'cat(cuda12.8::bin_path("nvcc"))')
CUDA_INC = $(shell "$(R_HOME)/bin/Rscript" -e 'cat(cuda12.8::include_path("runtime"))')
CUDA_LIB = $(shell "$(R_HOME)/bin/Rscript" -e 'cat(cuda12.8::lib_path())')

NVCC = $(CUDA_BIN)/nvcc

OBJECTS = hypot.o hypot_cuda.o hypot_kernel.o
PKG_LIBS = -L$(CUDA_LIB) -lcudart -Wl,-rpath,$(CUDA_LIB)

%.o: %.cu
	$(NVCC) -std=c++17 -O2 -I$(CUDA_INC) -Xcompiler "$(CXXPICFLAGS)" -c $< -o $@
```

`nvcc` finds its own CRT headers next to its `bin/`, so only the runtime
headers have to be named explicitly.
A package meant to install on machines without a GPU wants a `configure` that
checks whether `cuda12.8` (or a system `nvcc`) is there at all, and generates
a `Makevars` without the CUDA objects when it is not.

### Registering both

`.onLoad()` then offers whichever handlers were actually built:

```r
symbol <- function(name, pkgname) {
  tryCatch(getNativeSymbolInfo(name, PACKAGE = pkgname)$address, error = function(e) NULL)
}

.onLoad <- function(libname, pkgname) {
  handlers <- list(
    cpu = symbol("HypotHandler", pkgname),
    cuda = symbol("HypotHandlerCuda", pkgname)
  )
  pjrt::pjrt_register_custom_call(
    "anvl_hypot",
    Filter(Negate(is.null), handlers),
    .package = pkgname
  )
}
```

Registering only `cpu` is not an error; a program that needs the missing
platform is what fails, with *No FFI handler registered for `anvl_hypot`*.
