---
title: "Data Types and Promotion Rules"
---

## Data Types

Every `AnvlArray` has a data type, which `dtype()` reports.
The data types fall into three categories, where the number is the width in bits:

| Category | Data types |
|----------|------------|
| Boolean | `bool` |
| Integer | signed: `i8`, `i16`, `i32`, `i64`; unsigned: `ui8`, `ui16`, `ui32`, `ui64` |
| Float | `f32`, `f64` |

Other floating-point data types, such as `f16`, and complex numbers are currently not supported.

The data type of a new `AnvlArray` is chosen with the `dtype` argument, and `nv_convert()` converts an existing one:

```{r}
library(anvl)
x <- nv_array(1:3, dtype = "i16")
dtype(x)
nv_convert(x, "f64")
```

Without a `dtype`, an R `logical` becomes a `bool`, and R `double`s and `integer`s become the default float and integer data types, which `default_dtypes()` reports and the `anvl.default_dtypes` option configures:

```{r}
default_dtypes()
```

Converting an `AnvlArray` back to R with `as_array()` gives the R type that can represent its values, e.g. a `double` for an `f32`; see `?as_array` for the full list.
Help pages name groups of data types that an argument accepts with single words, such as *numeric* or *integerish*, which `?dtypes` defines.

The rest of this article covers which data type the result of an operation has when its inputs have different ones.

## Type Promotion Rules

When combining arrays of different types (e.g., adding an `f32` to an `i32`), {anvl} needs to determine a common type.
For example, below we are adding an `f32` to an `f64`, where the former is promoted to the latter's type, because it's more expressive.

```{r}
nv_add(
  nv_scalar(1.0, dtype = "f32"),
  nv_scalar(1.0, dtype = "f64")
)
```

The rules are defined by the `common_dtype()` function.

```{r}
common_dtype("f64", "f32")
common_dtype("i64", "f32")
```

The table below shows the common data type of every pair of data types.
The cells are colored by the category of the result:
<span style="background-color: rgba(128, 128, 128, 0.25); padding: 0 0.3em;">boolean</span>,
<span style="background-color: rgba(66, 133, 244, 0.25); padding: 0 0.3em;">signed integer</span>,
<span style="background-color: rgba(52, 168, 83, 0.25); padding: 0 0.3em;">unsigned integer</span>,
<span style="background-color: rgba(251, 140, 0, 0.3); padding: 0 0.3em;">float</span>, and
<span style="background-color: rgba(229, 57, 53, 0.25); padding: 0 0.3em;">none</span>.

```{r, echo = FALSE}
dtypes <- c("bool", "i8", "i16", "i32", "i64", "ui8", "ui16", "ui32", "ui64", "f32", "f64")
# Semi-transparent, so the colors work with both the light and the dark theme.
colors <- c(
  bool = "rgba(128, 128, 128, 0.25)",
  int = "rgba(66, 133, 244, 0.25)",
  uint = "rgba(52, 168, 83, 0.25)",
  float = "rgba(251, 140, 0, 0.3)",
  none = "rgba(229, 57, 53, 0.25)"
)
category <- function(dt) {
  if (dt == "bool") "bool" else if (startsWith(dt, "ui")) "uint" else if (startsWith(dt, "i")) "int" else "float"
}
cell <- function(text, color) {
  sprintf('<span style="display: block; background-color: %s; text-align: center;">%s</span>', color, text)
}
tbl <- matrix(NA_character_, length(dtypes), length(dtypes))
for (i in seq_along(dtypes)) {
  for (j in seq_along(dtypes)) {
    # `ui64` and a signed integer have no common data type.
    common <- tryCatch(as.character(common_dtype(dtypes[i], dtypes[j])), error = function(e) NULL)
    tbl[i, j] <- if (is.null(common)) cell("--", colors[["none"]]) else cell(common, colors[[category(common)]])
  }
}
# The column headers are bold as table headers; the row labels are cells, so
# they are emphasized here to match.
rownames(tbl) <- sprintf("<strong>%s</strong>", dtypes)
colnames(tbl) <- dtypes
knitr::kable(tbl, format = "html", escape = FALSE, caption = "Common data type of two data types")
```

Two integer types meet at a type that holds every value of both, so `i8` and `ui8` meet at `i16`.
The exception is `ui64`: no signed integer type holds its values, and an integer never becomes a float on its own, so a `ui64` and a signed integer have no common type at all (the red cells above).
This is where JAX and NumPy fall back to a float -- {anvl} instead asks you to say what you want, by converting one of the operands with `nv_convert()`.

R values are handled differently: {anvl}'s type system keeps their full precision, and lets the author of a function decide how they are promoted, without losing precision along the way.
We describe how this works below.

## R Values Have No Data Type

An R number is a 64-bit double, while {anvl} computes in `f32` by default.
There are two simple ways to reconcile the two, and both have a drawback:

- Converting R doubles always to `f64` means that every R number, such as a step size or `pi`, pulls the computation to `f64`, which is considerably slower on GPUs.
- Converting R doubles always to the default `f32` means that an R number loses precision before it is even used, also when it is combined with an `f64` `AnvlArray`.

Below, `exact` is what {anvl} computes, and `rounded` is what the second approach would give, where `pi` is rounded to an `f32` first:

```{r}
exact <- nv_add(pi, nv_scalar(2, dtype = "f64"))
rounded <- nv_add(
  nv_convert(nv_scalar(pi, dtype = "f32"), "f64"),
  nv_scalar(2, dtype = "f64")
)
print(c(exact = as_array(exact), rounded = as_array(rounded), base_r = pi + 2), digits = 16)
```

{anvl} avoids both drawbacks: an R value has no data type of its own, and takes the data type of the `AnvlArray` it is combined with.
Combined with an `f64`, `pi` keeps its full precision and gives the same result as base R, while combined with an `f32`, the computation stays fast.
This makes {anvl} well suited for statistical computations, where precision often matters a lot.

In {anvl}'s type system, R objects do not have a concrete data type.

```{r, error = TRUE}
dtype(1)
```

The type promotion table from above therefore does not apply to them.
However, it is possible to apply an {anvl} function to R values (length-1 vectors and arrays):

```{r}
nv_exp(1)
nv_add(1, nv_scalar(1, "f64"))
nv_add(1, 2)
```

It is therefore important to understand the rules that govern the materialization of R objects as `AnvlArray`s.
There is no uniform rule how this is done, functions can decide this for themselves, but the two most common routes are:

1. The R value is materialized at its default data type, which `default_dtypes()` reports: a `double` and an `integer` take the defaults of the active backend, configurable through the `anvl.default_dtypes` option, and a `logical` takes `bool`.
   This is e.g. the case in unary functions such as `nv_exp`.
2. The R value's data type is inferred from the other arguments, as in the `nv_add` call above.
   When no concrete data type is present, `nv_add(1, 2)` falls back to that default again.

Either way the value is *built at* that data type rather than converted into it, so one the data type cannot hold is an error:

```{r, error = TRUE}
jit(function(x) x + (-2L))(nv_scalar(1L, "ui8"))
```

Converting an array is a different matter and keeps XLA's wraparound semantics:

```{r}
nv_convert(nv_scalar(-2L, "i32"), "ui8")
```

Note that these rules are not universal and exceptions exist.
Some functions, such as `nv_clamp`, prioritize the data type of a specific argument, in this case `x`, the value that is being clamped.
It fails if the boundary values cannot be promoted to `x` without loss of precision:

```{r, error = TRUE}
nv_clamp(
  x = nv_scalar(0, "f32"),
  min = nv_scalar(0, "f64"),
  max = nv_scalar(0, "f64")
)
```

Functions document their behavior, so consult their respective help page for more information on how they promote their inputs.

When you are writing your own {anvl} functions, you can decide the promotion rules yourself.
This is possible via the `.promote` argument of `as_anvl_arrays()`.
It takes a `PromotionRule`, which is a special function that takes in the arguments and outputs a data type for each one.

```{r}
as_anvl_arrays(1, 2, .promote = promotion_dtype("f64"))
```

One common rule is `promotion_common()`, which is used by functions such as `nv_add` above.
It computes the common data type of the inputs, so below, the R value `1` takes the data type of the `f64` input:

```{r}
as_anvl_arrays(1, nv_scalar(2, dtype = "f64"), .promote = promotion_common())
```

When none of the inputs has a data type of its own, e.g. `as_anvl_arrays(1, 2, .promote = promotion_common())`, it falls back to the default data type of an R `double`.
We can override that default via `with_default_dtypes()`, which we use below to ask for double precision.
The two elements contrast what the override does and does not touch: `materialized` is an R literal with
no data type of its own, so it follows the new default, while `yielded` meets a typed `f32` array and
takes *its* data type -- a default only decides what happens when nothing else does.

```{r}
with_default_dtypes(c(float = "f64"), {
  list(
    materialized = jit(\() 1.0)(),
    yielded = nv_array(1, dtype = "f32") + 1.5
  )
})
```

Below, we compute the common data type of the arguments.

```{r}
promotion_fn <- promotion_common()
args <- list(1, 2, nv_scalar(1L, "i8"))
promotion_fn(args)
```

When used in `as_anvl_arrays()`, it converts the inputs to that data type:

```{r}
do.call(as_anvl_arrays, c(args, list(.promote = promotion_fn)))
```

For more information about the available rules, see `?promotion_rule`.
